---
title: "Data analysis Imperial college, pilot study - Full data."
format: html
tbl-cap-location: bottom
editor: visual
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
author: Philippine Louail, Ana Sanchez Lorenzo
---

```{r setup, include=FALSE}
library(knitr)
library(quarto)
knitr::opts_knit$set(root.dir = './')
```

# Data description

\- Here describe what your data is, important experimental info,.... People
should understand the most important info before running the code.

## loading packages

```{r packages-used, message=FALSE, warning=FALSE}
## General bioconductor package
library(BiocStyle)
library(BiocParallel)
library(knitr)

## Data Import and handling
library(readxl)
library(MsExperiment)
library(SummarizedExperiment)

## Preprocessing of LC-MS data
library(xcms)
library(Spectra)
library(MetaboCoreUtils)

## Statistical analysis
library(limma) # Differential abundance
library(matrixStats) # Summaries over matrices

## Visualisation
library(pander)
library(RColorBrewer)
library(pheatmap)
library(vioplot)
library(ggfortify)   # Plot PCA
library(gridExtra)   # To arrange multiple ggplots into single plots

## Annotation
library(AnnotationHub) # Annotation resources
library(CompoundDb)    # Access small compound annotation data.
library(MetaboAnnotation) # Functionality for metabolite annotation.
```

# Data import

```{r import, message=FALSE, warning=FALSE}
# Data import
#' read the sample descriptions from an xlsx sheet
pd <- read_xlsx("metadata_lcms_rpos.xlsx", sheet = "samples") |>
  as.data.frame()

#' Import the data
MZML_PATH <- file.path("mzml rpos")

lcms1 <- readMsExperiment(file.path(MZML_PATH, "/", pd$mzML_file),
                         sampleData = pd)
idx_qc <- which(sampleData(lcms1)$phenotype == "QC")
idx_blank <- which(sampleData(lcms1)$phenotype == "BLANK")
```

```{r par-process}
#' Set up parallel processing using 3 cores
if (.Platform$OS.type == "unix") {
  register(MulticoreParam(3)) ###chunksize has to be be the same number. 
 } else {
     register(SnowParam(3))
}
```

## Data organisation

```{r}
lcms1
```

Below we provide a brief overview of the data structure and content. The
`sampleData()` function extracts sample information from the object.

```{r phenodata}
#| tbl-cap: "Table 1. Samples from the data set."
sampleData(lcms1) |> kable()
```

There are `r length(sampleData(lcms1))` samples in this data set. We only use
QCs here, this will allows to get an idea of the parameters needed for the
overall analysis without wasting too much time duirng computation.

We will define colors for each of the sample groups based on their sample group
using the *RColorBrewer* package:

```{r define-colors, include=FALSE}
#' Define colors for the different QCs 
col_phenotype <- brewer.pal(7, name = "Set3")
names(col_phenotype) <- unique(sampleData(lcms1)$phenotype)
col_sample <- col_phenotype[sampleData(lcms1)$phenotype]
```

The MS data of this experiment is stored as a *`Spectra` object* (from the
`r Biocpkg("Spectra")` Bioconductor package) within the `MsExperiment` object
and can be accessed using `spectra()` function. Each element in this object is
a spectrum

```{r}
#' Access Spectra Object
spectra(lcms1)
```

We can also summarize the number of spectra and their respective MS level. We
only have MS1 here.

```{r}
#' Count the number of spectra with a specific MS level per file.
spectra(lcms1) |>
    msLevel() |>
    split(fromFile(lcms1)) |>
    lapply(table) |>
    do.call(what = cbind)
```

The present data set thus contains only MS1 data, which is ideal for
quantification of the signal. Also all the samples seem to have a similar
number of scans, which is good. 

As another example, we below determine the retention time range for the entire
data set.

```{r}
#' Retention time range for entire dataset
spectra(lcms1) |>
    rtime() |>
    range()
```

Data obtained from LC-MS experiments are typically analyzed along the retention
time axis, while MS data is organized by spectrum, orthogonal to the retention
time axis.

# Data visualization and general quality assessment

Effective visualization is paramount for inspecting and assessing the quality
of MS data. For a general overview of our LC-MS data, we can:

-   Combine all mass peaks from all (MS1) spectra of a sample into a single
    spectrum in which each mass peak then represents the maximum signal of all
    mass peaks with a similar *m/z*. This spectrum might then be called Base
    Peak Spectrum (BPS), providing information on the most abundant ions of a
    sample.
-   Aggregate mass peak intensities for each spectrum, resulting in the Base
    Peak Chromatogram (BPC). The BPC shows the highest measured intensity for
    each distinct retention time (hence spectrum) and is thus orthogonal to the
    BPS.
-   Sum the mass peak intensities for each spectrum to create a Total Ion
    Chromatogram (TIC).
-   Compare the BPS of all samples in an experiment to evaluate similarity of
    their ion content.
-   Compare the BPC of all samples in an experiment to identify samples with
    similar or dissimilar chromatographic signal.

In addition to such general data evaluation and visualization, it is also
crucial to investigate specific signal of e.g. internal standards or
compounds/ions known to be present in the samples. By providing a reliable
reference, internal standards help achieve consistent and accurate analytical
results.

## Spectra Data Visualization: BPS

This is going to take way to long so not running this.
Maybe run it if you have some suspicions on the dataset later. 

## Chromatographic Data Visualization: BPC and TIC

```{r bpc1}
#| fig-cap: "Figure 5. BPC of all samples colored by phenotype."
#' Extract and plot BPC for full data
bpc <- chromatogram(lcms1, aggregationFun = "max")

plot(bpc, col = paste0(col_sample, 80), main = "BPC", lwd = 1.5)
grid()
legend("topright", col = col_phenotype,
       legend = names(col_phenotype), lty = 1, lwd = 2, horiz = TRUE,
       bty = "n")
```


We can also here compare similarities of the BPCs in a heatmap.

```{r heatmap1}
#| fig-cap: "Figure 7. Heatmap of BPC similarities."
#' Total ion chromatogram
tic <- chromatogram(lcms1, aggregationFun = "sum") |>
  bin(binSize = 2)

#' Calculate similarity (Pearson correlation) between BPCs
ticmap <- do.call(cbind, lapply(tic, intensity)) |>
  cor()

#' Add sample names as rownames and colnames
rownames(ticmap) <- colnames(ticmap) <- sampleData(lcms1)$sample_name
ann <- data.frame(phenotype = sampleData(lcms1)[, "phenotype"])
rownames(ann) <- rownames(ticmap)

#' Plot heatmap
pheatmap(ticmap, annotation_col = ann,
         annotation_colors = list(phenotype = col_phenotype),
         show_rownames = FALSE, show_colnames = FALSE)
```

This is cool. 

### Known compounds

Throughout the entire process, it is crucial to have reference points within
the dataset, such as well-known ions. Most experiments nowadays include
internal standards (IS), and it was the case here. We strongly recommend using
them for visualization throughout the entire analysis.

-   Describe your standards quickly

We also plot all of them and save the images locally to check manually.

```{r EIC_extract_for_internal_standard, include=FALSE}

#' get the list
intern_standard <- read_xlsx("standard_compound_addition.xlsx") |> 
  as.data.frame()
intern_standard <- intern_standard[!is.na(intern_standard$POS), ]
rownames(intern_standard) <- intern_standard$abbreviation

#'generate calcualte formula
intern_standard$mass <- calculateMass(intern_standard$formula)
intern_standard$mz <- mapply(intern_standard$mass,
                             intern_standard$POS,
                             FUN = mass2mz)

#' Fit for each standards
intern_standard$mzmin <- intern_standard$mz - 0.02
intern_standard$mzmax <- intern_standard$mz + 0.02
intern_standard$rtmin <- intern_standard$RT - 15
intern_standard$rtmax <- intern_standard$RT + 15

#' Extract the EICs
eics <- chromatogram(lcms1,
                     mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]),
                     rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                     chunkSize = 2)

rownames(fData(eics)) <- intern_standard$abbreviation

dr <- "EIC_IS/full/raw/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eics[i, ], main = intern_standard$name[i],
         col = paste0(col_sample, 80))
    grid()
    legend("topright", col = col_phenotype,
           legend = names(col_phenotype), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}
```


# Data preprocessing

## Chromatographic peak detection

Reuse the parameters determined in the QC file.

```{r run-find-chrompeak-on-entire-dataset}
#' Using the same settings, but with default snthresh
param <- CentWaveParam(peakwidth = c(5, 20), ppm = 25, integrate = 2, snthresh = 7)
lcms1<- findChromPeaks(lcms1, param = param, chunkSize = 3)

#' Update EIC internal standard object
eics_peakdetect <- chromatogram(lcms1,
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
fData(eics_peakdetect) <- fData(eics)

## plot 
dr <- "EIC_IS/full/peakdetect/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eics_peakdetect[i, ], main = intern_standard$name[i],
         col = paste0(col_sample, 80))
    grid()
    legend("topright", col = col_phenotype,
           legend = names(col_phenotype), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}

save(lcms1, file = "peakdetect_lcms1.RData")
```

### Refine identified chromatographic peaks

```{r}
#' Apply on whole dataset
param <- MergeNeighboringPeaksParam(expandRt = 7, expandMz = 0.0015,
                                    minProp = 0.75)
lcms1 <- refineChromPeaks(lcms1, param = param, chunkSize = 3)
chromPeakData(lcms1)$merged |>
                      table()
```

We check the number of identified chromatographic
peaks in all samples of a data set can help spotting potentially problematic
samples. Below we count the number of chromatographic peaks per sample and show
these numbers in a table.

```{r}

#| tbl-cap: "Table 5. Samples and number of identified chromatographic peaks."
#' Count the number of peaks per sample and summarize them in a table.
data.frame(sample_name = sampleData(lcms1)$sample_name,
           peak_count = as.integer(table(chromPeaks(lcms1)[, "sample"]))) |>
    kable(format = "pipe")
```


## Retention time alignment

We do an inital correspodence round on the QC samples. The alignment will be 
based on the features shared in these samples. 

```{r}
#| fig-cap: "Figure 16. Initial correspondence analysis."
# Initial correspondence analysis
param <- PeakDensityParam(sampleGroups = sampleData(lcms1)$phenotype == "QC",
                          minFraction = 1,
                          binSize = 0.01, ppm = 10,
                          bw = 2)
lcms1 <- groupChromPeaks(lcms1, param = param)
```


```{r}
#' Define parameters of choice
subset <- which(sampleData(lcms1)$phenotype == "QC")
param <- PeakGroupsParam(minFraction = 0.9, extraPeaks = 50, span = 0.5,
                         subsetAdjust = "average",
                         subset = subset)

#' Perform the alignment
lcms1 <- adjustRtime(lcms1, param = param)
```


```{r}
#| fig-cap: "Figure 17. Retention time alignment results."
#' Visualize alignment results
plotAdjustedRtime(lcms1, col = paste0(col_sample, 80), peakGroupsPch = 1)
grid()
legend("topright", col = col_phenotype,
       legend = names(col_phenotype), lty = 1, bty = "n")
```

```{r}
#' Get data before alignment
lcms1_raw <- dropAdjustedRtime(lcms1)

#' Apply the adjusted retention time to our dataset
lcms1 <- applyAdjustedRtime(lcms1)
```



```{r}
#' Update EIC internal standard object
eics_align <- chromatogram(lcms1,
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
fData(eics_align) <- fData(eics)

## plot 
dr <- "EIC_IS/full/align/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eics_align[i, ], main = intern_standard$name[i],
         col = paste0(col_sample, 80))
    grid()
    legend("topright", col = col_phenotype,
           legend = names(col_phenotype), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}
## they all look like they were detected.

save(lcms1, file = "align_lcms1.RData")
```


## Correspondence

```{r}
#' Define the settings for the param
param <- PeakDensityParam(sampleGroups = sampleData(lcms1)$phenotype, ## this groups per devices, but maybe we want to group per triplicate ? 
                          minFraction = 0.75, binSize = 0.01, ppm = 10,
                          bw = 2)

#' Apply to whole data
lcms1 <- groupChromPeaks(lcms1, param = param)

save(lcms1, file = "corres_lcms1.RData")
```

plot eics: 

```{r}
#' Plot again
eics_dens <- chromatogram(lcms1,
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
fData(eics_dens) <- fData(eics)

dr <- "EIC_IS/full/density/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plotChromPeakDensity(eics_dens[i, ], param = param,
                         main = intern_standard$name[i],   
                         col = paste0(col_sample, "80"),
    peakCol = col_sample[chromPeaks(eics_dens[i, ])[, "sample"]],
    peakBg = paste0(col_sample[chromPeaks(eics_dens[i, ])[, "sample"]], 20),
    peakPch = 16)
    grid()
    legend("topright", col = col_phenotype,
           legend = names(col_phenotype), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}
```

Another interesting information to look a the the distribution of these
features along the retention time axis.

```{r}
#| tbl-cap: "Table 5. Distribution of features along the retention time axis."
# Bin features per RT slices
vc <- featureDefinitions(lcms1)$rtmed
breaks <- seq(0, max(vc, na.rm = TRUE) + 1, length.out = 15) |>
    round(0)
cuts <- cut(vc, breaks = breaks, include.lowest = TRUE)

table(cuts) |>
    kable(format = "pipe")
```

The results from the correspondence analysis are now stored, along with the
results from the other preprocessing steps, within our `XcmsExperiment` result
object.

This data frame provides the average *m/z* and retention time (in columns
`"mzmed"` and `"rtmed"`) that characterize a LC-MS feature. Column, `"peakidx"`
contains the indices of all chromatographic peaks that were assigned to that
feature. The actual abundances for these features, which represent also the
final preprocessing results, can be extracted with the `featureValues()`
function:

```{r}
#' Extract feature abundances
featureValues(lcms1, method = "sum") |>
    head()
```

We can note that some features (e.g. F0002) have missing values in
some samples. This is expected to a certain degree as not in all samples
features, respectively their ions, need to be present. We will address this in
the next section.

+ this quick check kinda make sense because we saw that there was less peaks in
in the later QCs. 

```{r}
nrow(featureDefinitions(lcms1))
```


## Gap filling

```{r}
#' Percentage of missing values
sum(is.na(featureValues(lcms1))) /
    length(featureValues(lcms1)) * 100
```
That's a nice number. 


```{r}
#' Fill in the missing values in the whole dataset
lcms1 <- fillChromPeaks(lcms1, param = ChromPeakAreaParam(), chunkSize = 3)

#' Percentage of missing values after gap-filling
sum(is.na(featureValues(lcms1))) /
    length(featureValues(lcms1)) * 100
```
 The gap-filling rescue ALL missing values. 

```{r Detected-vs-filled-signal1}
#' Get only detected signal in QC samples
vals_detect <- featureValues(lcms1, filled = FALSE)

#' Get detected and filled-in signal
vals_filled <- featureValues(lcms1)

#' Replace detected signal with NA
vals_filled[!is.na(vals_detect)] <- NA

#' Identify features with at least one filled peak
has_filled <- is.na(rowSums(vals_detect))

#' Calculate row averages for features with missing values
avg_detect <- rowMeans(vals_detect[has_filled, ], na.rm = TRUE)
avg_filled <- rowMeans(vals_filled[has_filled, ], na.rm = TRUE)

#' Plot the values against each other (in log2 scale)
plot(log2(avg_detect), log2(avg_filled),
     xlim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     ylim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     pch = 21, bg = "#00000020", col = "#00000080")
grid()
abline(0, 1)
```

The detected (x-axis) and gap-filled (y-axis) values for QC samples are highly
correlated. Especially for higher abundances, the agreement is very high, while
for low intensities, as can be expected, differences are higher and trending to
below the correlation line. Below we, in addition, fit a linear regression line
to the data and summarize its results

```{r}
#' fit a linear regression line to the data
l <- lm(log2(avg_filled) ~ log2(avg_detect))
summary(l)
```

The linear regression line has a slope of `r round(coef(l)[2], 2)` and an
intercept of `r round(coef(l)[1], 2)`. This indicates that the filled-in signal
is on average `r round(coef(l)[2], 2)` times higher than the detected signal.

## Preprocessing results

The final results of the LC-MS data preprocessing are stored within the
`XcmsExperiment` object. As well as the exact preprocessing performed and each
setting used. 

```{r Process-history}
#' Check first step of the process history
processHistory(lcms1)[[1]]
```


```{r}
#' Extract results as a SummarizedExperiment
res <- quantify(lcms1, method = "sum", filled = FALSE)
res
```


```{r}
assays(res)$raw_filled <- featureValues(lcms1, method = "sum",
                                        filled = TRUE )

#' Different assay in the SummarizedExperiment object
assayNames(res)
```



```{r}
#' Save the preprocessing results
#' d <- file.path(tempdir(), "objects/lcms1")
# saveMsObject(lcms1, AlabasterParam(path = d))
#' for now let's do R object because the previous method is not implemented yet.
save(lcms1, file = "preprocessed_lcms1.RData")
```
