---
title: "Data analysis of Pilot Study from Imperial - Full data."
format: html
tbl-cap-location: bottom
editor: visual
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
author: Philippine Louail, Ana Sanchez Lorenzo
---

```{r setup, include=FALSE}
library(knitr)
library(quarto)
knitr::opts_knit$set(root.dir = './')
```

# Data description

Rapid Microbore Metabolic Profiling (RAMMP) methods were developed to provide
ultra-fast liquid chromatography analysis cycles of up to five minutes while
maintaining high retention time repeatability. This is achieved by scaling down
column, tubing, and probe dimensions, reducing particle size, and decreasing
flow rates.

This study evaluates the feasibility of using RAMMP method for metabolic
analysis of blood microsampling device (capitainer) and compares its performance
with a conventional UPLC-MS approach, as well as with reference liquid samples,
including plasma and whole blood.

[**Samples of study**]{.underline}

The samples used in this study were venous blood samples collected from 10
healthy individuals ( 5 men and 5 women) in EDTA tubes. Part of this blood was
used as whole blood sample, part was used for plasma extraction and part was
spiked in capitainer devices to simulate sample collection through fingerprick.
Samples were provided by the HUMAN Doctoral Network.

Total number of samples: 52

-   10 x capitainer

-   10 x plasma

-   10 x blood

-   3 x pooled sample from capitainer

-   3 x pooled sample from plasma

-   3 x pooled sample from blood

-   9 x QC sample accross the analysis.

-   2 x double blank beginning

-   1 x blank capitainer

-   1 x blank end

[**Sample preparation and acquisition**]{.underline}

Samples were treated H2O, for sample extraction from device and blood
hidrolisis, and IPA for protein precipitation and lipids extraction. Then,
extracts were analysed in a UPLC-ESI-QTOF system (Synapt-G2S) (Waters
Corporation) using resolution mode for MS1 and FastDDA mode for MS2 data. This
batch of samples only represents the reversed-phase positive ionisation mode
data acquisition.

The convention in-house method was scaled down using a UPLC ACQUITY™ BEH C8 (1.0
× 50 mm, 1.7 µm) column, with a reduced flow rate of 0.3 mL/min over 3.7 minutes
and 1 µL injection volume.

Data acquisition was conducted on a Synapt G2-S system in positive ionisation
mode. In this study,

[**Sample preparation and acquisition**]{.underline}

Samples were treated H2O and IPA for protein precipitation and lipids
extraction. Then extracts were analysed in a UPLC-ESI-QTOF system (Synapt-G2S)
(Waters Corporation) using resolution mode for MS1 and FastDDA for MS2 data.

## Loading packages

```{r packages-used, message=FALSE, warning=FALSE}
#| code-fold: true
#| code-summary: "Show the code"
## General bioconductor package
library(BiocStyle)
library(BiocParallel)
library(knitr)

## Data Import and handling
library(readxl)
library(MsExperiment)
library(SummarizedExperiment)

## Preprocessing of LC-MS data
library(xcms)
library(Spectra)
library(MetaboCoreUtils)

## Statistical analysis
library(limma) # Differential abundance
library(matrixStats) # Summaries over matrices

## Visualisation
library(pander)
library(RColorBrewer)
library(pheatmap)
library(vioplot)
library(ggfortify)   # Plot PCA
library(gridExtra)   # To arrange multiple ggplots into single plots
library(grid)

## Annotation
library(AnnotationHub) # Annotation resources
library(CompoundDb)    # Access small compound annotation data.
library(MetaboAnnotation) # Functionality for metabolite annotation.
```

# Data import

```{r import, message=FALSE}
# Data import
#' read the sample descriptions from an xlsx sheet
pd <- read_xlsx("metadata_lcms_rpos_rammp.xlsx", sheet = "samples") |>
  as.data.frame()

#' Import the data
MZML_PATH <- file.path("data/lpos_rammp")

lcms1 <- readMsExperiment(file.path(MZML_PATH, "/", pd$mzML_file),
                         sampleData = pd)
```

```{r par-process}
#' Set up parallel processing using 3 cores
if (.Platform$OS.type == "unix") {
  register(MulticoreParam(4)) ### chunksize has to be be the same number. 
 } else {
     register(SnowParam(4))
}
```

## Data organisation

## Indexing per sample group

```{r indexing, message=FALSE}
# Index per sample group
idx_qc <- sampleData(lcms1)$device == "QC"
idx_blank <- sampleData(lcms1)$device == "BLANK"
idx_blank_qc <- sampleData(lcms1)$device %in% c("QC","BLANK")
idx_double_blank <- sampleData(lcms1)$blank_type %in% c("BLANK_001", "BLANK_002", "BLANK_END")
```

```{r lcms1}
lcms1
```

Below we provide a brief overview of the data structure and content. The
`sampleData()` function extracts sample information from the object.

```{r phenodata}
#| tbl-cap: "Table 1. Samples from the data set."
sampleData(lcms1) |> kable()
```

There are `r length(sampleData(lcms1))` samples in this data set. We only use
QCs here, this will allows to get an idea of the parameters needed for the
overall analysis without wasting too much time during computation.

## Define colors

We will define colors for each of the sample groups based on their sample group
using the *RColorBrewer* package:

```{r define-colors}
#| echo: true
#' Define colors for full dataset
leg_full <- brewer.pal(12, name = "Set3")[c(9, 12, 5, 6, 8, 4, 1, 2)]
names(leg_full) <- unique(sampleData(lcms1)$device)
col_sample <- leg_full[sampleData(lcms1)$device]

#' Define colors for only QCs in injection order
leg_qc <- brewer.pal(9, name = "Set3")
names(leg_qc)  <- unique(sampleData(lcms1[idx_qc])$sample_name)
col_qc_inj_order <-leg_qc[sampleData(lcms1[idx_qc])$sample_name]

#' Define colors for full dataset without blanks
leg_noblank <- leg_full[c(-1)]
col_noblank <- leg_noblank[sampleData(lcms1[!idx_blank])$device]

#' Define colors for samples only 
leg_devices <- leg_full[-c(1,2)]
col_devices <- leg_devices[sampleData(lcms1[!idx_blank_qc])$device]

#' Define colors for full dataset with different blank type
leg_blank_type <- brewer.pal(12, name ="Set3")[c(2, 9, 3, 12, 5, 6, 8, 4, 1, 2, 11)]
names(leg_blank_type) <- unique(sampleData(lcms1)$blank_type)
col_sample_blank_type <- leg_blank_type[sampleData(lcms1)$blank_type]
```

The MS data of this experiment is stored as a *`Spectra` object* (from the
`r Biocpkg("Spectra")` Bioconductor package) within the `MsExperiment` object
and can be accessed using `spectra()` function. Each element in this object is a
spectrum.

```{r spectra-objec, eval = FALSE}
#' Access Spectra Object
spectra(lcms1)
```

We can also summarize the number of spectra and their respective MS level. We
only have MS1 here.

```{r number_spectra, eval = FALSE}
#' Count the number of spectra with a specific MS level per file.
spectra(lcms1) |>
    msLevel() |>
    split(fromFile(lcms1)) |>
    lapply(table) |>
    do.call(what = cbind)
```

The present data set thus contains only MS1 data, which is ideal for
quantification of the signal. Also all the samples seem to have a similar number
of scans (961), which is good. However, in comparison with conventional method
(\~3500) it is 1/3.

As another example, we below determine the retention time range for the entire
data set.

```{r rt-range}
#' Retention time range for entire dataset
spectra(lcms1) |>
    rtime() |>
    range()
```

Data obtained from LC-MS experiments are typically analyzed along the retention
time axis, while MS data is organized by spectrum, orthogonal to the retention
time axis.

# Data visualization and general quality assessment

Effective visualization is paramount for inspecting and assessing the quality of
MS data. For a general overview of our LC-MS data, we can:

-   Combine all mass peaks from all (MS1) spectra of a sample into a single
    spectrum in which each mass peak then represents the maximum signal of all
    mass peaks with a similar *m/z*. This spectrum might then be called **Base
    Peak Spectrum (BPS)**, providing information on the most abundant ions of a
    sample.
-   Aggregate mass peak intensities for each spectrum, resulting in the **Base
    Peak Chromatogram (BPC)**. The BPC shows the highest measured intensity for
    each distinct retention time (hence spectrum) and is thus orthogonal to the
    BPS.
-   Sum the mass peak intensities for each spectrum to create a **Total Ion
    Chromatogram (TIC).**

In addition to such general data evaluation and visualization, it is also
crucial to investigate **specific signal** of e.g. internal standards or
compounds/ions known to be present in the samples. By providing a reliable
reference, internal standards help achieve consistent and accurate analytical
results.

## Chromatographic Data Visualization: BPC and TIC

```{r bpc1}
#| fig-cap: "Figure 1. BPC before rt filtering of all samples colored by device."
#| code-fold: true
#| code-summary: "Show the code"
#' Extract and plot BPC for full data
bpc <- chromatogram(lcms1, aggregationFun = "max")

plot(bpc, col = paste0(col_sample, 80), main = "BPC", lwd = 1.5)
grid()
legend("topright", col = leg_full, legend = names(leg_full), 
       lty = 1, lwd = 2, horiz = FALSE,
       bty = "n", cex = 0.9)
```

Evaluating the BPC, we detected a intense peak between 0 to 15 seconds of
features that didn't retain in the columns and a typical base line peak coming
from the solvent (15-25). Also after 180 there shouldn 't be any interesting
features. Rentention time filtering will need to be performed.

We can also compare similarities of the TICs in the heatmap. Thus we bin() the
chromatographic signal per sample along the retention time axis into a bins of 2
seconds resulting in the same number of bins/data points. We can then calculate
the pairwise similarities between these data vectors using the cor() function
and visualize the result using pheatmap().

```{r heatmap}
#| fig-cap: "Figure 2. Heatmap of TIC similarities before filtering."
#| code-fold: true
#| code-summary: "Show the code"
#' Total ion chromatogram
tic <- chromatogram(lcms1, aggregationFun = "sum") |>
  bin(binSize = 2)

#' Calculate similarity (Pearson correlation) between BPCs
ticmap <- do.call(cbind, lapply(tic, intensity)) |>
  cor()

#' Add sample names as rownames and colnames
rownames(ticmap) <- colnames(ticmap) <- sampleData(lcms1)$sample_name
ann <- data.frame(blank_type = sampleData(lcms1)[, "blank_type"])
rownames(ann) <- rownames(ticmap)

#' Plot heatmap
pheatmap(ticmap, annotation_col = ann,
         annotation_colors = list(blank_type = leg_blank_type),
         show_rownames = FALSE, show_colnames = FALSE)
```

In this case dataset, Blood and Plasma group together with their correspondent
pool samples. However, capitainer is separated and half of the sample from
capitainer are more similar to QC samples.

If we look at the TIC signal deeper splited per 2-second rt bins along the
x-axis, we could see how samples differed.

```{r evalution-chromatogram-differences}
#| fig-cap: "Figure 3. Heatmap of TIC similarities per 2-second retention time bins along the x-axis"
#| code-fold: true
#| code-summary: "Show the code"
#| fig-height: 5
#| fig-width: 10
ticm <- do.call(cbind, lapply(tic, intensity))
rownames(ticm) <- rtime(tic[, 1L])
colnames(ticm) <- sampleData(lcms1)$sample_name
#' transpose matrix such that rows are samples and columns retention times
ticm <- t(ticm)

#' log2 transforming the intensities and centering them by retention time bin
#' such that each column contains the relative tic for that retention time
ticm_c <- log2(ticm) |>
    scale(center = TRUE, scale = FALSE)
pheatmap(ticm_c, annotation_row = ann, cluster_cols = FALSE, 
         annotation_colors = list(blank_type = leg_blank_type))
```

The heatmap above represents the relative TIC signal per 2-second retention
timebins along the x-axis (relative TIC compared to the average TIC across all
samples in that bin). Samples are shown in columns, and clustered based on their
similarity. Microsampling devices differ mostly in the beginning of the LC. The
largest difference can be observed between some Capitainer samples (7) and pool
Capitainer and all other samples. All these differences are found at the
begining of the run, which should be features that did not retained in the
column, but they are influencing the clustering of the samples.

Below we repeat the plot, focusing on the first 30 seconds of the LC.

```{r evalution-chromatogram-differences-30s}
#| fig-cap: "Figure 4. Heatmap of TIC similarities per 2-second retention time bins along 2 and 30 seconds"
#| code-fold: true
#| code-summary: "Show the code"
tmp <- ticm_c[, as.numeric(colnames(ticm_c)) < 30]
pheatmap(tmp, annotation_row = ann, cluster_cols = FALSE, 
         annotation_colors = list(blank_type = leg_blank_type))

```

The Capitainer samples show a higher signal between 6 and 10 seconds compared to
all other samples. Looking at the chromatogram and gradient, we could filter rt
of injection phase and reconditioning. Thus, I filtered the data removing
everything before 24s and after 190 seconds.

```{r bpc-rtfilter}
#' Filter the data based on retention time
lcms1 <- filterRt(lcms1, c(24, 190))
```

```{r bpc-filtered}
#| fig-cap: "Figure 5. BPC after rt filtering of all samples colored by device."
#| code-fold: true
#| code-summary: "Show the code"
# BPC filtered
bpc_filtered <- chromatogram(lcms1, aggregationFun = "max")
#' Plot after filtering
plot(bpc_filtered, col = paste0(col_sample, 80),
     main = "BPC after filtering retention time", lwd = 1.5)
grid()
legend("topright", col = leg_full,
       legend = names(leg_full), lty = 1, lwd = 2, horiz = FALSE, bty = "n",
       cex = 0.9)

```

We can also here compare similarities of the BPCs in a heatmap.

```{r heatmap-filtered}
#| fig-cap: "Figure 6. Heatmap of TIC similarities after rt filtering."
#| code-fold: true
#| code-summary: "Show the code"

#' Total ion chromatogram
tic <- chromatogram(lcms1, aggregationFun = "sum") |>
  bin(binSize = 2)

#' Calculate similarity (Pearson correlation) between BPCs
ticmap <- do.call(cbind, lapply(tic, intensity)) |>
  cor()

#' Add sample names as rownames and colnames
rownames(ticmap) <- colnames(ticmap) <- sampleData(lcms1)$sample_name
ann <- data.frame(sample_type = sampleData(lcms1)[, "blank_type"])
rownames(ann) <- rownames(ticmap)

#' Plot heatmap
pheatmap(ticmap, annotation_col = ann,
         annotation_colors = list(sample_type = leg_blank_type),
         show_rownames = FALSE, show_colnames = FALSE)
```

After filtering the rt range samples clustered Capitainer and Blood together and
plasma a bit further. All QCs clustered as well together except of 2 samples
that mixed with blood samples. However, the clustring is more distributed.

### Known compounds

Throughout the entire process, it is crucial to have reference points within the
dataset, such as well-known ions, useful for the visualization of data
throughout the entire analysis. Most experiments nowadays include internal
standards (IS). In this case, no standards were spiked, so 12 known compounds
were selected from the National Phenome Center library. This library was built
using the same LCMS method and metabolites were annotated using standards
(<https://github.com/phenomecentre/npc-open-lcms/blob/main/LC-MS%20Metabolite%20Annotations/LPOS_ROI_V_5_1_2.csv>).

-   PC(16:0/18:2)
-   LPC(0:0/18:2)
-   PC(18:0/18:2)
-   Cholesterol
-   LPC(0:0/16:0)
-   LPC(18:0/0:0)
-   LPC(18:1/0:0)
-   LPC(O-18:0/0:0)
-   SM(d16:1/16:0)
-   PC(16:0/20:4)
-   PC(16:0/16:0)
-   SM(d18:2/24:0)

We also plot all of them and save the images locally to check manually.

```{r EIC_extract_for_internal_standard1}
#' get the list
intern_standard <- read_xlsx("standard_compound_addition_rammp.xlsx") |> 
  as.data.frame()
intern_standard <- intern_standard[!is.na(intern_standard$POS), ]
rownames(intern_standard) <- intern_standard$abbreviation

#'generate calcualte formula
intern_standard$mass <- calculateMass(intern_standard$formula)
intern_standard$mz <- mapply(intern_standard$mass,
                             intern_standard$POS,
                             FUN = mass2mz)

#' Fit for each standards
intern_standard$mzmin <- intern_standard$mz - 0.02
intern_standard$mzmax <- intern_standard$mz + 0.02
intern_standard$rtmin <- intern_standard$RT - 15
intern_standard$rtmax <- intern_standard$RT + 15
```

```{r EIC_extract_for_internal_standard2}
#' Extract the EICs
eics <- chromatogram(lcms1[idx_qc],
                     mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]),
                     rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                     chunkSize = 4)

rownames(fData(eics)) <- intern_standard$abbreviation

dr <- "EIC_IS/full/raw/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eics[i, ], main = intern_standard$name[i],
         col = paste0(col_qc_inj_order, 80))
    grid()
    legend("topright", col = leg_qc,
           legend = names(leg_qc), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}
```

EICs extracted show good reproducibility in retention time, but differences in
intensity. In this case, most of the peaks present tail, which could be realted
with excessive sample volume or column overload. Also, a few compounds are s
spicky which could be related with the sensitibility of the good rt
reproducibility, but intensity variability between qcs.

# Data preprocessing

## Chromatographic peak detection

Before the chromatographic peak detection, it is importnat to estimate the ppm
value:

```{r ppm-parameter1}
#' Restrict the data to signal from PC 18:0/18:2 (PC) in the first sample (5L) OPTIMISE PPM
PC <- lcms1[4L] |>
  spectra() |>
  filterRt(rt = c(95, 105)) |>
  filterMzRange(mz = fData(eics)["PC_18_0_18_2", c("mzmin", "mzmax")])

#' Show the number of peaks per m/z filtered spectra
lengths(PC)

#' Calculate the difference in m/z values between scans
mz_diff <- PC |>
    mz() |>
    unlist() |>
    diff() |>
    abs()

#' Express differences in ppm
range(mz_diff * 1e6 / mean(unlist(mz(PC)), na.rm = FALSE), na.rm = FALSE)
```

PPM estimated: 33.75 The ppm estimated using PC_18_0_18_2 in the fist QC sample
is of 34ppm for the settings of peak detection. In comparison with the
conventional method, I increased the ppm value for RAMMP LPOS dataset.

```{r run-find-chrompeak-on-entire-dataset, eval=!file.exists("peakdetect_lcms1.RData")}
#' Using the same settings, but with default snthresh
param <- CentWaveParam(peakwidth = c(2, 10), ppm = 35, integrate = 2, 
                       snthresh = 2)
#param <- CentWaveParam(peakwidth = c(2, 10), ppm = 35, integrate = 2, 
#                       snthresh = 7, noise = 600)
#test in EIC: decrease snthre to 3. 
# lcms1_test <- findChromPeaks(lcms1[idx_qc], param = param, chunkSize = 4)
lcms1 <- findChromPeaks(lcms1, param = param, chunkSize = 4)
save(lcms1, file = "peakdetect_lcms1.RData")
```

Above we found the parameters that have beeing optimised. We decided to continue
with snthresh = 7 and noise = 600 as there were not that much difference in the
detection of peaks, but the number of features decreased from \~ ????? to
\~????.

```{r}
# eics_peakdetect_test <- chromatogram(lcms1_test,
#                        rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
#                        mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
# fData(eics_peakdetect_test) <- fData(eics)
# 
# ## plot 
# dr <- "EIC_IS/full/peakdetecttest/"
# dir.create(dr, recursive = TRUE, showWarnings = FALSE)
# for (i in seq_len(nrow(intern_standard))) {
#     png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
#         width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
#     ## assign the sample color to each chromatographic peak
#     col_bg <- leg_qc[chromPeaks(eics_peakdetect_test[i, ])[, "column"]]
#     plot(eics_peakdetect_test[i, ], main = intern_standard$name[i],
#          col = paste0(col_qc_inj_order, 80), peakBg = paste0(col_bg, 40))
#     grid()
#     legend("topright", col = leg_qc,
#            legend = names(leg_qc), lty = 1)
#     abline(v = intern_standard$RT[i], col = "red", lty = 3)
#     dev.off()
# }
```

```{r run-find-chrompeak-on-entire-dataset2, eval=FALSE,eval=!file.exists("peakdetect_lcms1.RData")}
#' Update EIC internal standard object
eics_peakdetect <- chromatogram(lcms1[idx_qc],
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
fData(eics_peakdetect) <- fData(eics)

## plot 
dr <- "EIC_IS/full/peakdetect/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    ## assign the sample color to each chromatographic peak
    col_bg <- leg_qc[chromPeaks(eics_peakdetect[i, ])[, "column"]]
    plot(eics_peakdetect[i, ], main = intern_standard$name[i],
         col = paste0(col_qc_inj_order, 80), peakBg = paste0(col_bg, 40))
    grid()
    legend("topright", col = leg_qc,
           legend = names(leg_qc), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}
```

```{r run-find-chrompeak-on-entire-dataset-2, eval=file.exists("peakdetect_lcms1.RData")}
load("peakdetect_lcms1.RData")
```

All peaks of known compounds were detected so no further optimisation was
performed.

### Refine identified chromatographic peaks

As the CentWave algorith can miss some peaks due to overlapping or split peaks
it is important to use from XCMS refinechromPeaks() and
MergeNeighboringPeaksParam() to merge some split peaks. expandMz is important to
keep it small to avoid merging isotopes. ExpanRt: half size of av rt width.
minProp: to determine

```{r refine1, eval=!file.exists("refine_identified_lcms1.RData")}
#' Apply on whole dataset
param <- MergeNeighboringPeaksParam(expandRt = 2, expandMz = 0.0015,
                                    minProp = 0.75)
# lcms1_test <- refineChromPeaks(lcms1[idx_qc], param = param, chunSize = 4)
# chromPeakData(lcms1_test)$merged |> table()
lcms1 <- refineChromPeaks(lcms1, param = param, chunkSize = 4)
chromPeakData(lcms1)$merged |>
                      table()
save(lcms1, file = "refine_identified_lcms1.RData")
```

After refinement, the chrompeak detected decreased from 59309 to 54848.

```{r eval=file.exists("refine_identified_lcms1.RData")}
load("refine_identified_lcms1.RData")
```

We check the number of identified chromatographic peaks in all samples of a data
set can help spotting potentially problematic samples. Below we count the number
of chromatographic peaks per sample and show these numbers in a table.

```{r refine3}
#| tbl-cap: "Table 2. Samples and number of identified chromatographic peaks."
#' Count the number of peaks per sample and summarize them in a table.
data.frame(sample_name = sampleData(lcms1)$sample_name,
           peak_count = as.integer(table(chromPeaks(lcms1)[, "sample"]))) |>
    kable(format = "pipe") 
```

There is a decrease of identified peaks related with injection order. In average
there will be \~ 150 peaks detected in blanks and \~ 1000 peaks detected in
study samples. NOTE: FIX. shouldn t be less than 1000?

## Retention time alignment

We do an initial correspondence round on the QC samples. The alignment will be
based on the features shared in these samples.

```{r rt_alignment_1, eval=!file.exists("align_lcms1.RData")}
#| fig-cap: "Figure 7. Initial correspondence analysis."
# Initial correspondence analysis
param <- PeakDensityParam(sampleGroups = sampleData(lcms1)$device == "QC",
                          minFraction = 1,
                          binSize = 0.01, ppm = 10,
                          bw = 2)
lcms1 <- groupChromPeaks(lcms1, param = param)
```

```{r rt_alignment_2, eval=!file.exists("align_lcms1.RData")}
#' Define parameters of choice
subset <- which(sampleData(lcms1)$device == "QC")
param <- PeakGroupsParam(minFraction = 0.9, extraPeaks = 50, span = 0.5,
                         subsetAdjust = "average",
                         subset = subset)

#' Perform the alignment

lcms1 <- adjustRtime(lcms1, param = param)

```

```{r rt_alignment_plot, eval = FALSE}
#| fig-cap: "Figure 8. Retention time alignment results."
#| fig-height: 5
#| fig-width: 10
#' Visualize alignment results
par(mar = c(5, 5, 6, 10))
plotAdjustedRtime(lcms1, col = paste0(col_sample, 80), peakGroupsPch = 1,
                  cex.lab = 1.5)
grid()
legend("topright", inset = c(-0.3, 0), col = leg_full, legend = names(leg_full), lty = 1, bty = "n", horiz = FALSE, cex = 0.9, xpd = TRUE)
# legend("topright", col = leg_full,
       # legend = names(leg_full), lty = 1, bty = "n", horiz = FALSE, cex = 0.9)
```

The differences in retention times are -1s to +0.5s. This will require retention
time alignment.

```{r rt_alignment_4, eval=!file.exists("align_lcms1.RData")}
#' Get data before alignment
lcms1_raw <- dropAdjustedRtime(lcms1)

#' Apply the adjusted retention time to our dataset
lcms1 <- applyAdjustedRtime(lcms1)

# Save aligment object 
save(lcms1, file = "align_lcms1.RData")
```

```{r eval=file.exists("align_lcms1.RData")}
load("align_lcms1.RData")
```

```{r rt_alignment_5, eval=FALSE}
# RUN WITH QCs (idx)
#' Update EIC internal standard object
eics_align <- chromatogram(lcms1[idx_qc],
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
fData(eics_align) <- fData(eics)

## plot 
dr <- "EIC_IS/full/align/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    col_bg <- leg_qc[chromPeaks(eics_align[i, ])[, "column"]]
    plot(eics_align[i, ], main = intern_standard$name[i],
         col = paste0(col_qc_inj_order, 80), peakBg = paste0(col_bg, 40))
    legend("topright", col = leg_qc,
           legend = names(leg_qc), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}
```

Use the next code to test some extraction of chromatograms.

```{r test-ch, eval=FALSE}
# Test chromatogram()
c_test <- chromatogram(lcms1[5L], rt = c(30, 55), mz = c(600, 750)) 
plot(c_test, peakType = "none") 
```

NOTE: what is peakType doing?

## Correspondence

As described above, for the present data set we don't need to perform a
retention time alignment. Thus we load below the data that was saved **before**
performing the alignment and proceed the analysis with that data.

```{r eval=file.exists("refine_identified_lcms1.RData")}
load("refine_identified_lcms1.RData")
```

Generally, the goal of the correspondence analysis is to identify
chromatographic peaks that originate from the same types of ions in all samples
of an experiment and to group them into LC-MS features. Using the function
PlotChromPeakDensity() form xcms package, it is possible to simulate a
corresopndece analyis with the default values for Peak Groups on the EIC of
known compounds.

```{r correspondence_1, eval=!file.exists("corres_lcms1.RData")}
pd_new <- read_xlsx("metadata_lcms_rpos_rammp.xlsx", sheet = "samples") |>
  as.data.frame()
pd_new <- DataFrame(pd_new)
sampleData(lcms1) <- pd_new

sampleData(lcms1)$device
sampleData(lcms1)$sex

vec <- c("sex", "device")
p <- do.call(paste, c(as.list(sampleData(lcms1)[,vec]), sep = "_"))
p[p %in% c("NA_BLANK", "NA_QC")] <- NA
f <- factor(p)
f

#' Define the settings for the param
param <- PeakDensityParam(sampleGroups = f, ## this groups per devices 
                          minFraction = 0.75, binSize = 0.01, ppm = 35,
                          bw = 2)

#' Apply to whole data
lcms1 <- groupChromPeaks(lcms1, param = param)
save(lcms1, file = "corres_lcms1.RData")
nrow(featureDefinitions(lcms1))
```

Binsize depends on the resolution of the instrument. We decreate from 0.25 to
0.01. MinFraction: defines features only if chrompeak was identified in at least
75% of the samples of one of the sample groups. bw depends on the resolution of
the instrument as well, but so we decreased form 30 (default) to 2.

The total number of features detected is 1427. PPM (changed to 35 as tested).
????? why now it is 4256???

```{r featureDefinitions}
featureDefinitions(lcms1) |> head()
```

```{r eval=file.exists("corres_lcms1.RData")}
load("corres_lcms1.RData")
```

plot eics:

```{r correspondence_2, eval=FALSE}
#' Plot again
eics_dens <- chromatogram(lcms1[idx_qc],
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]),
                       chunkSize = 4)
fData(eics_dens) <- fData(eics)

dr <- "EIC_IS/full/density/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plotChromPeakDensity(eics_dens[i, ], param = param,
                         main = intern_standard$name[i],   
                         col = paste0(col_sample, "80"),
    peakCol = leg_qc[chromPeaks(eics_dens[i, ])[, "sample"]],
    peakBg = paste0(col_qc_inj_order[chromPeaks(eics_dens[i, ])[, "sample"]], 20),
    peakPch = 16)
    grid()
    legend("topright", col = leg_qc,
           legend = names(leg_qc), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}
```

Another interesting information to look a the the distribution of these features
along the retention time axis.

```{r correspondence_3}
#| tbl-cap: "Table 3. Distribution of features along the retention time axis."
# Bin features per RT slices
vc <- featureDefinitions(lcms1)$rtmed
breaks <- seq(0, max(vc, na.rm = TRUE) + 1, length.out = 15) |>
    round(0)
cuts <- cut(vc, breaks = breaks, include.lowest = TRUE)

table(cuts) |>
    kable(format = "pipe")
```

The results from the correspondence analysis are now stored, along with the
results from the other preprocessing steps, within our `XcmsExperiment` result
object.

This data frame provides the average *m/z* and retention time (in columns
`"mzmed"` and `"rtmed"`) that characterize a LC-MS feature. Column, `"peakidx"`
contains the indices of all chromatographic peaks that were assigned to that
feature. The actual abundances for these features, which represent also the
final preprocessing results, can be extracted with the `featureValues()`
function:

```{r feature_abundance}
#'Extract feature abundances
# featureValues(lcms1, method = "sum") |>
#    head()
```

We can note that some features (e.g. F0002) have missing values in some samples.
This is expected to a certain degree as not in all samples features,
respectively their ions, need to be present. We will address this in the next
section.

-   this quick check kinda make sense because we saw that there was less peaks
    in the later QCs.

```{r feature_abundance_2}
nrow(featureDefinitions(lcms1))
```

## Gap filling

The missing values (NA) is not always because the ion is not present but because
of the failure of the preceding chromatographic peak detection steps. In that
case, for qualitative and differential abundance analysis we will need to do
imputation. However, this gap filling step we can recover the missing values of
the latter category as much as possible to reduce the eventual need for data
imputation.

```{r gapfilling1}
#' Percentage of missing values
sum(is.na(featureValues(lcms1))) /
    length(featureValues(lcms1)) * 100
```

Gap filling: 43%.

43% of ions detected in different samples are missing. this could be due to some
missing data point in the middle of the peak. Instead of imputing of all them,
the gap filling to recues their original intensity value from the raw data. for
that, fillChromPeaks() function (xcms) was used to define an m/z-rt are for each
feature based on detected peaks, where the signal for the respective ion is
expected. Then it integrates the intensities in that area and report it as
feature abundance.

```{r gapfilling2, , eval=!file.exists("gapfilling_lcms1.RData")}
#' Fill in the missing values in the whole dataset
lcms1 <- fillChromPeaks(lcms1, param = ChromPeakAreaParam(), chunkSize = 4)
save(lcms1, file = "gapfilling_lcms1.RData")
```

```{r eval=file.exists("gapfilling_lcms1.RData")}
load("gapfilling_lcms1.RData")
```

```{r gapfilling3}
#' Percentage of missing values after gap-filling
sum(is.na(featureValues(lcms1))) /
    length(featureValues(lcms1)) * 100
```

The gap-filling rescue up to 6.2% Missing Value.

```{r}
lcms1_noblanks <- lcms1[sampleData(lcms1)$device!="BLANK", keepFeatures = TRUE]
```

```{r Detected-vs-filled-signal1}
#| fig-cap: "Figure 9. Detect vs filled signal (all dataset)."
#Distribution of data 

# A. DISTRUBTION in ALL SAMPLES. 

#' Get only detected signal in all samples
vals_detect <- featureValues(lcms1, filled = FALSE)
#' Get detected and filled-in signal
vals_filled <- featureValues(lcms1)
#' Replace detected signal with NA
vals_filled[!is.na(vals_detect)] <- NA
#' Identify features with at least one filled peak
has_filled <- is.na(rowSums(vals_detect))
#' Calculate row averages for features with missing values
avg_detect <- rowMeans(vals_detect[has_filled, ], na.rm = TRUE)
avg_filled <- rowMeans(vals_filled[has_filled, ], na.rm = TRUE)
#' Plot the values against each other (in log2 scale)
plot(log2(avg_detect), log2(avg_filled),
     xlim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     ylim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     pch = 21, bg = "#00000020", col = "#00000080",  
     xlab = "log2(avg_detected)"
     )
grid()
abline(0, 1)

```

Detected and gap-filled values are correlated, but for a considerable number of
features the gap-filled values are lower than the respective data from the
detected chromatographic peaks. This plot is created with full data

```{r}
#' Identify gap-filled features with a log2 difference in abundance being > 3
#' Note: log2 difference of 2 represents a 2^3 = 8-fold difference in natural scale
#| fig-cap: "Figure 10. Detect vs filled signal (heatmap)."
# sel <- which((log2(avg_detect) - log2(avg_filled)) > 3)
# 
# #' Get the feature values for these
# vals <- featureValues(lcms1)[has_filled, ]
# vals <- vals[sel, ]
# 
# colnames(vals) <- sampleData(lcms1)$sample_name
# ## Just use BLANK instead of the individual BLANKS
# ann$device[grep("BLANK", ann$device)] <- "BLANK"
# 
# pheatmap(log2(vals), annotation_col = ann,
#          annotation_colors = list(device = leg_full),
#          show_rownames = FALSE, show_colnames = FALSE)

```

We next evaluate the gap-filling using the QC samples. For these we can assume
the same sample composition and gap-filled signal should thus be more similar to
detected signal.

```{r Detected-vs-filled-signal1-qc}
#Distribution of data 
#| fig-cap: "Figure 11. Detect vs filled signal (QCs)."
qc <- lcms1[sampleData(lcms1)$device == "QC", keepFeatures = TRUE]

#' Get only detected signal in all samples
vals_detect <- featureValues(qc, filled = FALSE)
#' Get detected and filled-in signal
vals_filled <- featureValues(qc)
#' Replace detected signal with NA
vals_filled[!is.na(vals_detect)] <- NA
#' Identify features with at least one filled peak
has_filled <- is.na(rowSums(vals_detect))
#' Calculate row averages for features with missing values
avg_detect <- rowMeans(vals_detect[has_filled, ], na.rm = TRUE)
avg_filled <- rowMeans(vals_filled[has_filled, ], na.rm = TRUE)
#' Plot the values against each other (in log2 scale)
plot(log2(avg_detect), log2(avg_filled),
     xlim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     ylim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     pch = 21, bg = "#00000020", col = "#00000080",  
     xlab = "log2(avg_detected)"
     )
grid()
abline(0, 1)

```

The detected (x-axis) and gap-filled (y-axis) values for QC samples are highly
correlated. Especially for higher abundances, the agreement is very high, while
for low intensities, as can be expected, differences are higher and trending to
below the correlation line. Below we, in addition, fit a linear regression line
to the data and summarize its results

```{r lr}
#' fit a linear regression line to the data
l <- lm(log2(avg_filled) ~ log2(avg_detect))
summary(l)
```

Call: lm(formula = log2(avg_filled) \~ log2(avg_detect))

Residuals: Min 1Q Median 3Q Max -4.0534 -0.4925 -0.0300 0.4779 3.9436

Coefficients: Estimate Std. Error t value Pr(\>\|t\|)\
(Intercept) 0.486679 0.088185 5.519 3.75e-08 ***log2(avg_detect) 0.954437
0.008353 114.264 \< 2e-16*** --- Signif. codes: 0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05
‘.’ 0.1 ‘ ’ 1

Residual standard error: 0.7703 on 2564 degrees of freedom (434 observations
deleted due to missingness) Multiple R-squared: 0.8359, Adjusted R-squared:
0.8358 F-statistic: 1.306e+04 on 1 and 2564 DF, p-value: \< 2.2e-16

R quared smaller than 0.9 (0.83). Intercept low (0.48) and slope 0.95.

## Filtering Features: Blank flagging

Now we flagged features present mainly in the blank with Blankflag() function
from xcms package. This can be done in the "XcmsExperiment" or in the
"SummarizedExperiment" object.

There will be an extra column created in "featureDefinition" or "rowData" called
"*possible_contaminants*"with a logical value for each feature.

```{r blankflagfilter}
filter<- BlankFlag(blankIndex = idx_double_blank,
                   qcIndex = idx_qc,
)
lcms1 <- filterFeatures(lcms1, filter)

#number of flagged features : 
featureDefinitions(lcms1)$possible_contaminants |> sum(na.rm = TRUE)
```

Blankflag result filter: 3270 total features remain. 972 features were flagged.

### Ion Map

We will represent the m/z vs the retention time coloring the compounds that have
been flagged in the blank and that are present in possible_contaminant column.

```{r ion-map, eval=FALSE}
ion_map <- ggplot(featureDefinitions(lcms1), aes(x = rtmed, y = mzmed, color = possible_contaminants)) +
    geom_point() +
    scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red", "NA" = "grey")) +
    labs(x = "retention Time (rtmed)", y = "m/z (mzmed)", color = "Possible Contaminants") +
    theme_minimal(base_size = 10) +
    theme(panel.border = element_rect(color = "black", fill = NA, size = 0.8))
grid.arrange(ion_map, ncol = 1, heights = unit(4, "in"),widths = unit(7.5, "in"))

#featureDefinitions(lcms1)$possible_contaminants 
```

Representation with base R.

```{r}

fa <- featureArea(lcms1)
xl <- range(fa[, c("rtmin", "rtmax")])

## BPC
par(mfrow = c(2, 1), mar = c(0, 4.5, 0.5, 0.5))
plot(bpc, xlim = xl, main = "", xlab = "", xaxt = "n")
grid()
par(mar = c(4.5, 4.5, 0, 0.5))
## Feature "area"
## color for features
feat_border <- rep("#00000040", nrow(fa))
feat_border[featureDefinitions(lcms1)$possible_contaminants] <- "#ff000040"
plot(NA, NA, xlim = range(fa[, c("rtmin", "rtmax")]),
     ylim = range(fa[, c("mzmin", "mzmax")]), xlab = "retention time",
     ylab = "m/z")
rect(xleft = fa[, "rtmin"], xright = fa[, "rtmax"],
     ybottom = fa[, "mzmin"], ytop = fa[, "mzmax"], border = feat_border)
grid()
```

Some of the potential contaminant signal can be observed clustered as horizontal
lines along the full retention time range. These could represent contamination
with e.g. salts. Also, vertically aligned groups of features are present. These
could represent multiple ions/adducts of the same compound.

## Preprocessing results

The final results of the LC-MS data preprocessing are stored within the
`XcmsExperiment` object. As well as the exact preprocessing performed and each
setting used.

```{r Process-history}
#' Check first step of the process history
processHistory(lcms1)[[1]]
```

```{r extract_SumExp}
#' Extract results as a SummarizedExperiment
res <- quantify(lcms1, method = "sum", filled = FALSE)
res
```

```{r different_assays_sumExp}
assays(res)$raw_filled <- featureValues(lcms1, method = "sum",
                                        filled = TRUE )
#' Different assay in the SummarizedExperiment object
assayNames(res)
```

```{r save_preprocessed}
#' Save the preprocessing results
#' d <- file.path(tempdir(), "objects/lcms1")
# saveMsObject(lcms1, AlabasterParam(path = d))
#' for now let's do R object because the previous method is not implemented yet.
save(lcms1, file = "preprocessed_lcms1.RData")
save(res, file = "res.RData")
```

### Noise Comparison

Below we compare the noise signals between devices. We first calculate the
overall signal in the dataset and then calculate the signal that is in the
chromatographic peaks detection. We then subtract the two to get the noise
signal.

```{r Noise_comparison, echo=FALSE}
# overall signal in the dataset 
#' - for each file calculate the sum of intensities 
background  <- spectra(lcms1[!idx_blank_qc]) |>
    split(fromFile(lcms1[!idx_blank_qc])) |>
    lapply(tic) |>
    lapply(sum) |>
    unlist()
res[idx_qc]
# Overall signal that is in the chromatographic peaks detection
detected <- apply(assay(res[,!idx_blank_qc]), 2, function(x) sum(x, na.rm = TRUE))
 
# substract and plot
names(background) <- names(detected) <- res$device[!idx_blank_qc]
noise <- background - detected
 
f <- factor(names(noise), levels = unique(names(noise)))
group <- split(noise, f)
 
plot(NULL, xlim = c(1, length(group)), ylim = range(unlist(group)), 
     xaxt = "n", xlab = "Devices", ylab = "Noise", 
     main = "Noise comparison between Devices")
for (i in seq_along(group)) {
  points(rep(i, length(group[[i]])), group[[i]], pch = 19, 
col = leg_devices[i])
}
axis(1, at = seq_along(group), labels = names(group))

```

In this case, blood and plasma show more background noise than capitainer. Pool
sample form plasma has less reproducbility as well.
