---
title: "Data analysis of Pilot Study from Imperial - Full data."
format: html
tbl-cap-location: bottom
editor: visual
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
author: Philippine Louail, Ana Sanchez Lorenzo
---

```{r setup, include=FALSE}
library(knitr)
library(quarto)
knitr::opts_knit$set(root.dir = './')
```

# Data description

The aim of this pilot study is to compare three microsampling devices profiles
with conventional liquid samples, plasma and whole blood, in an untargeted LCMS
analysis. The final objective is to determine which device perform a similar
profile to plasma, gold standard sample, generating less artefacts. For this,
there will be a qualitative comparison (number of features, metabolites
annotated) and a semi-quantitative comparison (intensity and statistical
differences).

[**Samples of study**]{.underline}

The samples used in this study were venous blood samples collected from 10
healthy individuals ( 5 men and 5 women) in EDTA tubes. Part of this blood was
used as whole blood sample, part was used for plasma extraction and part was
spiked in devices to simulate sample collection through fingerprick.

Total number of samples: 66

double blanks:

-   BLANK_001(beginning)

-   BLANK_002 (beginning)

-   BLANK_003 (beginning)

-   BLANK_END

Blanks from devices:

-   Blank_W

-   Blank_M

-   Blank_C

Randomise Samples:

-   10 x WHATMAN

-   10 x MITRA

-   10 x CAPITAINER

-   10 x PLASMA

-   10 x BLOOD

-   10 x QCs

[**Sample preparation and acquisition**]{.underline}

Samples were treated H2O and IPA for protein precipitation and lipids
extraction. Then extracts were analysed in a UPLC-ESI-QTOF system (Synapt-G2S)
(Waters Corporation) using resolution mode for MS1 and FastDDA for MS2 data.

## Loading packages

```{r packages-used, message=FALSE, warning=FALSE}
#| code-fold: true
#| code-summary: "Show the code"
## General bioconductor package
library(BiocStyle)
library(BiocParallel)
library(knitr)

## Data Import and handling
library(readxl)
library(MsExperiment)
library(SummarizedExperiment)

## Preprocessing of LC-MS data
library(xcms)
library(Spectra)
library(MetaboCoreUtils)

## Statistical analysis
library(limma) # Differential abundance
library(matrixStats) # Summaries over matrices

## Visualisation
library(pander)
library(RColorBrewer)
library(pheatmap)
library(vioplot)
library(ggfortify)   # Plot PCA
library(gridExtra)   # To arrange multiple ggplots into single plots
library(grid)

## Annotation
library(AnnotationHub) # Annotation resources
library(CompoundDb)    # Access small compound annotation data.
library(MetaboAnnotation) # Functionality for metabolite annotation.
```

# Data import

```{r import, message=FALSE}
# Data import
#' read the sample descriptions from an xlsx sheet
pd <- read_xlsx("metadata_lcms_rpos.xlsx", sheet = "samples") |>
  as.data.frame()

#' Import the data
MZML_PATH <- file.path("data/lpos_conv")

lcms1 <- readMsExperiment(file.path(MZML_PATH, "/", pd$mzML_file),
                          sampleData = pd)
```

## Indexing per sample group

```{r indexing, message=FALSE}
# Index per sample group
idx_qc <- which(sampleData(lcms1)$device == "QC")
idx_blank <- which(sampleData(lcms1)$device == "BLANK")
idx_blank_qc <- sampleData(lcms1)$device %in% c("QC","BLANK")
idx_double_blank <- sampleData(lcms1)$blank_type %in% c("BLANK_001", "BLANK_002", "BLANK_END")
```

```{r par-process}
#' Set up parallel processing using 3 cores
if (.Platform$OS.type == "unix") {
  register(MulticoreParam(10)) ### chunksize has to be be the same number. 
 } else {
     register(SnowParam(10))
}
```

## Data organisation

```{r lcms1}
lcms1
```

Below we provide a brief overview of the data structure and content. The
`sampleData()` function extracts sample information from the object.

```{r phenodata}
#| tbl-cap: "Table 1. Samples from the data set."
sampleData(lcms1) |> kable()
```

METABONAUT: when I am rendering this dataset, the message below says the lenght
of sampleData(lcms1) is 8, like if there where 8 samples but is not correct. 
(jo: true - that's a bug in Metabonaut, it should be `nrow` instead as `length`
just gets the number of columns, not rows = samples)

There are `r nrow(sampleData(lcms1))` samples in this data set. We only use
QCs here, this will allows to get an idea of the parameters needed for the
overall analysis without wasting too much time during computation.

## Define colors

We will define colors for each of the sample groups based on their sample group
using the *RColorBrewer* package:

```{r define-colors, include=FALSE}
#' Define colors for dataset
col_device <- brewer.pal(7, name = "Set3")
names(col_device) <- unique(sampleData(lcms1)$device)
col_sample <- col_device[sampleData(lcms1)$device]
col_qc <- brewer.pal(10, name = "Set3")
names(col_qc) <- paste0("QC_", seq_along(col_qc))

#' Define colors for dataset without blanks
col_noblank <- col_sample[names(col_sample)!= "BLANK"]

#' Define colors for samples
col_sampleonly <- col_sample[!(names(col_sample) %in% c("BLANK", "QC"))]

#' Define colors for dataset with different blank type
col_blank_type <- brewer.pal(12, name ="Set3")
names(col_blank_type) <- unique(sampleData(lcms1)$blank_type)
```

The MS data of this experiment is stored as a *`Spectra` object* (from the
`r Biocpkg("Spectra")` Bioconductor package) within the `MsExperiment` object
and can be accessed using `spectra()` function. Each element in this object is a
spectrum.

```{r spectra-objec, eval = FALSE}
#' Access Spectra Object
spectra(lcms1)
```

We can also summarize the number of spectra and their respective MS level. We
only have MS1 here.

```{r number_spectra, eval = FALSE}
#' Count the number of spectra with a specific MS level per file.
spectra(lcms1) |>
    msLevel() |>
    split(fromFile(lcms1)) |>
    lapply(table) |>
    do.call(what = cbind)
```

The present data set thus contains only MS1 data, which is ideal for
quantification of the signal. Also all the samples seem to have a similar number
of scans, which is good.

As another example, we below determine the retention time range for the entire
data set.

```{r rt-range}
#' Retention time range for entire dataset
spectra(lcms1) |>
    rtime() |>
    range()
```

Data obtained from LC-MS experiments are typically analyzed along the retention
time axis, while MS data is organized by spectrum, orthogonal to the retention
time axis.

# Data visualization and general quality assessment

Effective visualization is paramount for inspecting and assessing the quality of
MS data. For a general overview of our LC-MS data, we can:

-   Combine all mass peaks from all (MS1) spectra of a sample into a single
    spectrum in which each mass peak then represents the maximum signal of all
    mass peaks with a similar *m/z*. This spectrum might then be called **Base
    Peak Spectrum (BPS)**, providing information on the most abundant ions of a
    sample.
-   Aggregate mass peak intensities for each spectrum, resulting in the **Base
    Peak Chromatogram (BPC)**. The BPC shows the highest measured intensity for
    each distinct retention time (hence spectrum) and is thus orthogonal to the
    BPS.
-   Sum the mass peak intensities for each spectrum to create a **Total Ion
    Chromatogram (TIC).**
-   **Compare the BPS** of all samples in an experiment to evaluate similarity
    of their ion content.
-   **Compare the BPC** of all samples in an experiment to identify samples with
    similar or dissimilar chromatographic signal.

In addition to such general data evaluation and visualization, it is also
crucial to investigate **specific signal** of e.g. internal standards or
compounds/ions known to be present in the samples. By providing a reliable
reference, internal standards help achieve consistent and accurate analytical
results.

## Spectra Data Visualization: BPS

(This is going to take way too long so not running this. Maybe run it if you
have some suspicions on the dataset later).

(Find in metabonaut)

## Chromatographic Data Visualization: BPC and TIC

```{r bpc1, eval=FALSE}
#| fig-cap: "Figure 1. BPC before rt filtering of all samples colored by device."
#| code-fold: true
#| code-summary: "Show the code"
#' Extract and plot BPC for full data
bpc <- chromatogram(lcms1, aggregationFun = "max")

plot(bpc, col = paste0(col_sample, 80), main = "BPC", lwd = 1.5)
grid()
legend("topright", col = col_device, legend = names(col_device), 
       lty = 1, lwd = 2, horiz = FALSE,
       bty = "n", cex = 0.9)
```

After filtering, we realised the Heatmap had a different layout. At the
beginning it seems the Capitainer devices were different to the rest. However,
now it seems all devices are grouped together. That is way, I want to plot the
heatmap again to check this variation is due to filtering or due to another
modification I am not aware of .

We can see separation of samples per group. Only Blood, Whatman and Mitra
samples are mixed, probably due to the same blood extraction. However,
Capitainer is separated from the others. This disappears if we filter the rt and
plot the heatmap again.

```{r heatmap, eval=FALSE}
#| fig-cap: "Figure 2. Heatmap of TIC similarities before filtering."
#| code-fold: true
#| code-summary: "Show the code"
#' Total ion chromatogram
tic <- chromatogram(lcms1, aggregationFun = "sum") |>
  bin(binSize = 2)

#' Calculate similarity (Pearson correlation) between BPCs
ticmap <- do.call(cbind, lapply(tic, intensity)) |>
  cor()

#' Add sample names as rownames and colnames
rownames(ticmap) <- colnames(ticmap) <- sampleData(lcms1)$sample_name
ann <- data.frame(device = sampleData(lcms1)[, "device"])
rownames(ann) <- rownames(ticmap)

#' Plot heatmap
pheatmap(ticmap, annotation_col = ann,
         annotation_colors = list(device = col_device),
         show_rownames = FALSE, show_colnames = FALSE)
```

Evaluating which part of the chromatogram, along the retention time contributes
most to the observed differences, e.g. of Whatman against the other samples.

```{r}
ticm <- do.call(cbind, lapply(tic, intensity))
rownames(ticm) <- rtime(tic[, 1L])
colnames(ticm) <- sampleData(lcms1)$sample_name
#' transpose matrix such that rows are samples and columns retention times
ticm <- t(ticm)

#' log2 transforming the intensities and centering them by retention time bin
#' such that each column contains the relative tic for that retention time
ticm_c <- log2(ticm) |>
    scale(center = TRUE, scale = FALSE)
pheatmap(ticm_c, annotation_row = ann, cluster_cols = FALSE)
```

The heatmap above represents the relative TIC signal per 2-second retention time
bins along the x-axis (relative TIC compared to the average TIC across all
samples in that bin). Samples are shown in columns, and clustered based on their
similarity. Microsampling devices differ mostly in the beginning of the LC. The
largest difference can be observed between Capitainer and all other
samples. Below we repeat the plot, focusing on the first 30 seconds of the LC.

```{r}
tmp <- ticm_c[, as.numeric(colnames(ticm_c)) < 30]
pheatmap(tmp, annotation_row = ann, cluster_cols = FALSE)

```

The Capitainer samples show a higher signal between 16 and 18 seconds compared
to all other samples.

METABONAUT: TIC instead of BPC right?

Looking at the chromatogram and gradient, we could filter rt of injection phase
and reconditioning. Thus, I filtered the data removing everything before 60s and
after 700 seconds.

NOTE: For rt_filtering I createad lcms1_filtered only for testing. Then I called
it again lcms1. (memory problem).

```{r bpc-rtfilter}
#' Filter the data based on retention time
lcms1 <- filterRt(lcms1, c(60, 700))
```

```{r bpc-filtered}
#| fig-cap: "Figure 3. BPC after rt filtering of all samples colored by device."
#| code-fold: true
#| code-summary: "Show the code"
# BPC filtered
bpc_filtered <- chromatogram(lcms1, aggregationFun = "max")
#' Plot after filtering
plot(bpc_filtered, col = paste0(col_sample, 80),
     main = "BPC after filtering retention time", lwd = 1.5)
grid()
legend("topright", col = col_device,
       legend = names(col_device), lty = 1, lwd = 2, horiz = FALSE, bty = "n",
       cex = 0.9)

```

We can also here compare similarities of the BPCs in a heatmap. REMINDER:
include png function with loop. Before plotting the heatmap after rt filtering,
I create a new column with the same device grouping, but with different blank
groups: BLANK_1, BLANK_2, BLANK_W, BLANK_M, BLANK_C, BLANK_END.

```{r heatmap1}
#| fig-cap: "Figure 4. Heatmap of TIC similarities after rt filtering."
#| code-fold: true
#| code-summary: "Show the code"

#' Total ion chromatogram
tic <- chromatogram(lcms1, aggregationFun = "sum") |>
  bin(binSize = 2)

#' Calculate similarity (Pearson correlation) between BPCs
ticmap <- do.call(cbind, lapply(tic, intensity)) |>
  cor()

#' Add sample names as rownames and colnames
rownames(ticmap) <- colnames(ticmap) <- sampleData(lcms1)$sample_name
ann <- data.frame(device = sampleData(lcms1)[, "blank_type"])
ann2 <- data.frame(device = sampleData(lcms1)[, "device"])
rownames(ann) <- rownames(ticmap)

#' Plot heatmap
pheatmap(ticmap, annotation_col = ann,
         annotation_colors = list(blank_type = col_device),
         show_rownames = FALSE, show_colnames = FALSE)
```

### Known compounds

Throughout the entire process, it is crucial to have reference points within the
dataset, such as well-known ions, useful for the visualization of data
throughout the entire analysis. Most experiments nowadays include internal
standards (IS). In this case, no standards were spiked, so 12 known compounds
were selected from the National Phenome Center library. This library was built
using the same LCMS method and metabolites were annotated using standards
(<https://github.com/phenomecentre/npc-open-lcms/blob/main/LC-MS%20Metabolite%20Annotations/LPOS_ROI_V_5_1_2.csv>).

-   PC(16:0/18:2)
-   LPC(0:0/18:2)
-   PC(18:0/18:2)
-   Cholesterol
-   LPC(0:0/16:0)
-   LPC(18:0/0:0)
-   LPC(18:1/0:0)
-   LPC(O-18:0/0:0)
-   SM(d16:1/16:0)
-   PC(16:0/20:4)
-   PC(16:0/16:0)
-   SM(d18:2/24:0)

We also plot all of them and save the images locally to check manually.

```{r EIC_extract_for_internal_standard1}
#' get the list
intern_standard <- read_xlsx("standard_compound_addition.xlsx") |> 
  as.data.frame()
intern_standard <- intern_standard[!is.na(intern_standard$POS), ]
rownames(intern_standard) <- intern_standard$abbreviation

#'generate calcualte formula
intern_standard$mass <- calculateMass(intern_standard$formula)
intern_standard$mz <- mapply(intern_standard$mass,
                             intern_standard$POS,
                             FUN = mass2mz)

#' Fit for each standards
intern_standard$mzmin <- intern_standard$mz - 0.02
intern_standard$mzmax <- intern_standard$mz + 0.02
intern_standard$rtmin <- intern_standard$RT - 15
intern_standard$rtmax <- intern_standard$RT + 15
```

```{r EIC_extract_for_internal_standard2, eval=FALSE}
#' Extract the EICs
eics <- chromatogram(lcms1[idx_qc],
                     mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]),
                     rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                     chunkSize = 10)

rownames(fData(eics)) <- intern_standard$abbreviation

dr <- "EIC_IS/full/raw/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eics[i, ], main = intern_standard$name[i],
         col = paste0(col_qc, 80))
    grid()
    legend("topright", col = col_qc,
           legend = names(col_qc), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}
```

EICs show good rt reproducibility, but intensity variability between samples,
not only between groups.

# Data preprocessing

## Chromatographic peak detection

Reuse the parameters determined in the QC file.

```{r run-find-chrompeak-on-entire-dataset, eval=!file.exists("peakdetect_lcms1.RData")}
#' Using the same settings, but with default snthresh
param <- CentWaveParam(peakwidth = c(5, 20), ppm = 25, integrate = 2, 
                       snthresh = 5)
lcms1 <- findChromPeaks(lcms1, param = param, chunkSize = 10)
save(lcms1, file = "peakdetect_lcms1.RData")
```

```{r run-find-chrompeak-on-entire-dataset2, eval=!file.exists("peakdetect_lcms1.RData")}
#' Update EIC internal standard object
eics_peakdetect <- chromatogram(lcms1[idx_qc],
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
fData(eics_peakdetect) <- fData(eics)

## plot 
dr <- "EIC_IS/full/peakdetect/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    ## assign the sample color to each chromatographic peak
    col_bg <- col_qc[chromPeaks(eics_peakdetect[i, ])[, "column"]]
    plot(eics_peakdetect[i, ], main = intern_standard$name[i],
         col = paste0(col_qc, 80), peakBg = paste0(col_bg, 40))
    grid()
    legend("topright", col = col_qc,
           legend = names(col_qc), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}
```

```{r run-find-chrompeak-on-entire-dataset-2, eval=file.exists("peakdetect_lcms1.RData")}
load("peakdetect_lcms1.RData")
```

### Refine identified chromatographic peaks

```{r refine1, eval=!file.exists("refine_identified_lcms1.RData")}
#' Apply on whole dataset
param <- MergeNeighboringPeaksParam(expandRt = 7, expandMz = 0.0015,
                                    minProp = 0.75)
lcms1 <- refineChromPeaks(lcms1, param = param, chunkSize = 10)
chromPeakData(lcms1)$merged |>
                      table()
save(lcms1, file = "refine_identified_lcms1.RData")
```

```{r eval=file.exists("refine_identified_lcms1.RData")}
load("refine_identified_lcms1.RData")
```

We check the number of identified chromatographic peaks in all samples of a data
set can help spotting potentially problematic samples. Below we count the number
of chromatographic peaks per sample and show these numbers in a table.

```{r refine3}
#| tbl-cap: "Table 2. Samples and number of identified chromatographic peaks."
#' Count the number of peaks per sample and summarize them in a table.
data.frame(sample_name = sampleData(lcms1)$sample_name,
           peak_count = as.integer(table(chromPeaks(lcms1)[, "sample"]))) |>
    kable(format = "pipe") 
```

There is a decrease of identified peaks related with injection order.

## Retention time alignment

We do an initial correspondence round on the QC samples. The alignment will be
based on the features shared in these samples.

```{r rt_alignment_1, eval=!file.exists("align_lcms1.RData")}
#| fig-cap: "Figure 5. Initial correspondence analysis."
# Initial correspondence analysis
param <- PeakDensityParam(sampleGroups = sampleData(lcms1)$device == "QC",
                          minFraction = 1,
                          binSize = 0.01, ppm = 10,
                          bw = 2)
lcms1 <- groupChromPeaks(lcms1, param = param)
```

```{r rt_alignment_2, eval=!file.exists("align_lcms1.RData")}
#' Define parameters of choice
subset <- which(sampleData(lcms1)$device == "QC")
param <- PeakGroupsParam(minFraction = 0.9, extraPeaks = 50, span = 0.5,
                         subsetAdjust = "average",
                         subset = subset)

#' Perform the alignment
lcms1 <- adjustRtime(lcms1, param = param)
```

```{r rt_alignment_3}
#| fig-cap: "Figure 6. Retention time alignment results."
#' Visualize alignment results
plotAdjustedRtime(lcms1, col = paste0(col_sample, 80), peakGroupsPch = 1, cex.lab = 1.2)
grid()
legend("topright", col = col_device,
       legend = names(col_device), lty = 1, bty = "n", horiz = TRUE)
```

The differences in retention times are, in particular if compared with the
retention time width of the identified chromatographic peaks, only very
small. For the present data set we will thus **not** perform a retention time
alignment.

```{r rt_alignment_4, eval=!file.exists("align_lcms1.RData")}
#' Get data before alignment
lcms1_raw <- dropAdjustedRtime(lcms1)

#' Apply the adjusted retention time to our dataset
lcms1 <- applyAdjustedRtime(lcms1)

# Save aligment object 
save(lcms1, file = "align_lcms1.RData")
```

```{r eval=file.exists("align_lcms1.RData")}
load("align_lcms1.RData")
```

```{r rt_alignment_5, eval=FALSE}
# RUN WITH QCs (idx)
#' Update EIC internal standard object
eics_align <- chromatogram(lcms1[idx_qc],
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
fData(eics_align) <- fData(eics)

## plot 
dr <- "EIC_IS/full/align/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eics_align[i, ], main = intern_standard$name[i],
         col = paste0(col_qc, 80))
    grid()
    legend("topright", col = col_device,
           legend = names(col_device), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}
```

```{r rt_alignment_6, eval=FALSE, include=FALSE}
# Test chromatogram()
c_test <- chromatogram(lcms1[1L], rt = c(400, 450), mz = c(100, 600)) 
plot(c_test, peakType = "none") # how to not have the detected peaks?
```

## Correspondence

As described above, for the present data set we don't need to perform a
retention time alignment. Thus we load below the data that was saved **before**
performing the alingnment and proceed the analysis with that data.

```{r eval=file.exists("refine_identified_lcms1.RData")}
load("refine_identified_lcms1.RData")
```

```{r correspondence_1, eval=!file.exists("corres_lcms1.RData")}
pd_new <- read_xlsx("metadata_lcms_rpos.xlsx", sheet = "samples") |>
  as.data.frame()
pd_new <- DataFrame(pd_new)
sampleData(lcms1) <- pd_new

sampleData(lcms1)$device
sampleData(lcms1)$sex

vec <- c("sex", "device")
p <- do.call(paste, c(as.list(sampleData(lcms1)[,vec]), sep = "_"))
p[p %in% c("NA_BLANK", "NA_QC")] <- NA
f <- factor(p)
f

#' Define the settings for the param
param <- PeakDensityParam(sampleGroups = f, ## this groups per devices 
                          minFraction = 0.75, binSize = 0.01, ppm = 25,
                          bw = 2)

#' Apply to whole data
lcms1 <- groupChromPeaks(lcms1, param = param)
save(lcms1, file = "corres_lcms1.RData")
nrow(featureDefinitions(lcms1))
featureDefinitions(lcms1) |> head()
```

```{r eval=file.exists("corres_lcms1.RData")}
load("corres_lcms1.RData")
```

plot eics:

```{r correspondence_2, eval=FALSE}
#' Plot again
eics_dens <- chromatogram(lcms1[idx_qc],
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]),
                       chunkSize = 10)
fData(eics_dens) <- fData(eics)

dr <- "EIC_IS/full/density/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plotChromPeakDensity(eics_dens[i, ], param = param,
                         main = intern_standard$name[i],   
                         col = paste0(col_sample, "80"),
    peakCol = col_qc[chromPeaks(eics_dens[i, ])[, "sample"]],
    peakBg = paste0(col_qc[chromPeaks(eics_dens[i, ])[, "sample"]], 20),
    peakPch = 16)
    grid()
    legend("topright", col = col_qc,
           legend = names(col_qc), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}
```

Another interesting information to look a the the distribution of these features
along the retention time axis.

```{r correspondence_3}
#| tbl-cap: "Table 5. Distribution of features along the retention time axis."
# Bin features per RT slices
vc <- featureDefinitions(lcms1)$rtmed
breaks <- seq(0, max(vc, na.rm = TRUE) + 1, length.out = 15) |>
    round(0)
cuts <- cut(vc, breaks = breaks, include.lowest = TRUE)

table(cuts) |>
    kable(format = "pipe")
```

The results from the correspondence analysis are now stored, along with the
results from the other preprocessing steps, within our `XcmsExperiment` result
object.

This data frame provides the average *m/z* and retention time (in columns
`"mzmed"` and `"rtmed"`) that characterize a LC-MS feature. Column, `"peakidx"`
contains the indices of all chromatographic peaks that were assigned to that
feature. The actual abundances for these features, which represent also the
final preprocessing results, can be extracted with the `featureValues()`
function:

```{r feature_abundance}
#'Extract feature abundances
# featureValues(lcms1, method = "sum") |>
#    head()
```

We can note that some features (e.g. F0002) have missing values in some samples.
This is expected to a certain degree as not in all samples features,
respectively their ions, need to be present. We will address this in the next
section.

-   this quick check kinda make sense because we saw that there was less peaks
    in in the later QCs.

```{r feature_abundance_2}
nrow(featureDefinitions(lcms1))
```

## Gap filling

```{r gapfilling1}
#' Percentage of missing values
sum(is.na(featureValues(lcms1))) /
    length(featureValues(lcms1)) * 100
```

Gap filling: 40%.

we could find different features between plasma vs blood and between devices.
(background noise???)

```{r gapfilling2, , eval=!file.exists("gapfilling_lcms1.RData")}
#' Fill in the missing values in the whole dataset
lcms1 <- fillChromPeaks(lcms1, param = ChromPeakAreaParam(), chunkSize = 10)
save(lcms1, file = "gapfilling_lcms1.RData")
```

```{r eval=file.exists("gapfilling_lcms1.RData")}
load("gapfilling_lcms1.RData")
```

```{r gapfilling3}
#' Percentage of missing values after gap-filling
sum(is.na(featureValues(lcms1))) /
    length(featureValues(lcms1)) * 100
```

The gap-filling rescue up to 2% Missing Value. When we go back to the raw data,
we rescue intensities that were measures for specific features. If there is no
intensity, it will no be filled and then we will need to do imputation in PCA.

```{r}
lcms1_noblanks <- lcms1[sampleData(lcms1)$device!="BLANK", keepFeatures = TRUE]
```

NOTE: when subsetting lcms1 object: use \[\], do not use coma, add keepFeatures
= TRUE to avoid removing feature definition.

```{r Detected-vs-filled-signal1}
#Distribution of data 

# A. DISTRUBTION in ALL SAMPLES. 

#' Get only detected signal in all samples
vals_detect <- featureValues(lcms1, filled = FALSE)
#' Get detected and filled-in signal
vals_filled <- featureValues(lcms1)
#' Replace detected signal with NA
vals_filled[!is.na(vals_detect)] <- NA
#' Identify features with at least one filled peak
has_filled <- is.na(rowSums(vals_detect))
#' Calculate row averages for features with missing values
avg_detect <- rowMeans(vals_detect[has_filled, ], na.rm = TRUE)
avg_filled <- rowMeans(vals_filled[has_filled, ], na.rm = TRUE)
#' Plot the values against each other (in log2 scale)
plot(log2(avg_detect), log2(avg_filled),
     xlim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     ylim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     pch = 21, bg = "#00000020", col = "#00000080",  
     xlab = "log2(avg_detected)"
     )
grid()
abline(0, 1)

```

Detected and gap-filled values are correlated, but for a considerable number of
features the gap-filled values are lower than the respective data from the
detected chromatographic peaks. Note however that the plot above was created on
the full data set, and differences sample composition respectively presence or
abundance of metabolites and features can be expected.

We next identify the gap-filled features with a more than 8-fold lower
gap-filled value compared to the detected value. We visually inspect this data
using a heatmap to potentially identify any relationship between sample groups
and differences in the gap-filled values.

```{r}
#' Identify gap-filled features with a log2 difference in abundance being > 3
#' Note: log2 difference of 2 represents a 2^3 = 8-fold difference in natural scale
sel <- which((log2(avg_detect) - log2(avg_filled)) > 3)

#' Get the feature values for these
vals <- featureValues(lcms1)[has_filled, ]
vals <- vals[sel, ]

colnames(vals) <- sampleData(lcms1)$sample_name
## Just use BLANK instead of the individual BLANKS
ann$device[grep("BLANK", ann$device)] <- "BLANK"

pheatmap(log2(vals), annotation_col = ann,
         annotation_colors = list(device = col_device),
         show_rownames = FALSE, show_colnames = FALSE)

```

Not unexpectedly, the gap-filled abundances in blanks are much lower for all the
selected features. In addition, a clear difference can also be seen for blood
and plasma samples, and to a lower extend also for Mitra samples. Since these
differences are consistent for all samples of the same *device* it can be
assumed that it stems from differences in either sample composition or
systematic device-specific contamination.

We next evaluate the gap-filling using the QC samples. For these we can assume
the same sample composition and gap-filled signal should thus be more similar to
detected signal.

```{r Detected-vs-filled-signal1-qc}
#Distribution of data 

qc <- lcms1[sampleData(lcms1)$device == "QC", keepFeatures = TRUE]

#' Get only detected signal in all samples
vals_detect <- featureValues(qc, filled = FALSE)
#' Get detected and filled-in signal
vals_filled <- featureValues(qc)
#' Replace detected signal with NA
vals_filled[!is.na(vals_detect)] <- NA
#' Identify features with at least one filled peak
has_filled <- is.na(rowSums(vals_detect))
#' Calculate row averages for features with missing values
avg_detect <- rowMeans(vals_detect[has_filled, ], na.rm = TRUE)
avg_filled <- rowMeans(vals_filled[has_filled, ], na.rm = TRUE)
#' Plot the values against each other (in log2 scale)
plot(log2(avg_detect), log2(avg_filled),
     xlim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     ylim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     pch = 21, bg = "#00000020", col = "#00000080",  
     xlab = "log2(avg_detected)"
     )
grid()
abline(0, 1)

```

The detected (x-axis) and gap-filled (y-axis) values for QC samples are highly
correlated. Especially for higher abundances, the agreement is very high, while
for low intensities, as can be expected, differences are higher and trending to
below the correlation line. Below we, in addition, fit a linear regression line
to the data and summarize its results

```{r lr}
#' fit a linear regression line to the data
l <- lm(log2(avg_filled) ~ log2(avg_detect))
summary(l)
```

The slope of the linear fit is close to 1 (1.03) and the intercept is also small
suggesting that gap filled signal is very close to the detected signal. Also,
the R squared is over 0.9 reflecting the high correlation of the values.


## Filtering Features: Missing values

Now we remove features present mainly in the blank with Blankflag() function
from xcms package. This can be done in the "XcmsExperiment" or in the
"SummarizedExperiment" object.

There will be an extra column created in "featureDefinition" or "rowData" called
"*possible_contaminants*"with a logical value for each feature.

```{r blankflagfilter}
filter<- BlankFlag(blankIndex = idx_double_blank,
                   qcIndex = idx_qc,
)
lcms1 <- filterFeatures(lcms1, filter)

#number of flagged features : 
featureDefinitions(lcms1)$possible_contaminants |> sum(na.rm = TRUE)
```

Blankflag result filter: 4166 features remain. 971 features were flagged. If I
only select double_blank only 770 features were flagged.

### Ion Map

We will represent the m/z vs the retention time coloring the compounds that have
been flagged in the blank and that are present in possible_contaminant column.

```{r ion-map}
ion_map <- ggplot(featureDefinitions(lcms1), aes(x = rtmed, y = mzmed, color = possible_contaminants)) +
    geom_point() +
    scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red", "NA" = "grey")) +
    labs(x = "retention Time (rtmed)", y = "m/z (mzmed)", color = "Possible Contaminants") +
    theme_minimal(base_size = 10) +
    theme(panel.border = element_rect(color = "black", fill = NA, size = 0.8))
grid.arrange(ion_map, ncol = 1, heights = unit(4, "in"),widths = unit(7.5, "in"))

#featureDefinitions(lcms1)$possible_contaminants 
```

(jo: just repeating that with base R.)

```{r}

fa <- featureArea(lcms1)
xl <- range(fa[, c("rtmin", "rtmax")])

## BPC
par(mfrow = c(2, 1), mar = c(0, 4.5, 0.5, 0.5))
plot(bpc, xlim = xl, main = "", xlab = "", xaxt = "n")
grid()
par(mar = c(4.5, 4.5, 0, 0.5))
## Feature "area"
## color for features
feat_border <- rep("#00000040", nrow(fa))
feat_border[featureDefinitions(lcms1)$possible_contaminants] <- "#ff000040"
plot(NA, NA, xlim = range(fa[, c("rtmin", "rtmax")]),
     ylim = range(fa[, c("mzmin", "mzmax")]), xlab = "retention time",
     ylab = "m/z")
rect(xleft = fa[, "rtmin"], xright = fa[, "rtmax"],
     ybottom = fa[, "mzmin"], ytop = fa[, "mzmax"], border = feat_border)
grid()
```

Some of the potential contaminant signal can be observed clustered as horizontal
lines along the full retention time range. These could represent contamination
with e.g. salts. Also, vertically aligned groups of features are present. These
could represent multiple ions/adducts of the same compound.


## Preprocessing results

The final results of the LC-MS data preprocessing are stored within the
`XcmsExperiment` object. As well as the exact preprocessing performed and each
setting used.

```{r Process-history}
#' Check first step of the process history
load("gapfilling_lcms1.RData")
processHistory(lcms1)[[1]]
```

```{r extract_SumExp}
#' Extract results as a SummarizedExperiment
res <- quantify(lcms1, method = "sum", filled = FALSE)
res
```

```{r different_assays_sumExp}
assays(res)$raw_filled <- featureValues(lcms1, method = "sum",
                                        filled = TRUE )

#' Different assay in the SummarizedExperiment object
assayNames(res)
```

```{r save_preprocessed}
#' Save the preprocessing results
#' d <- file.path(tempdir(), "objects/lcms1")
# saveMsObject(lcms1, AlabasterParam(path = d))
#' for now let's do R object because the previous method is not implemented yet.
save(lcms1, file = "preprocessed_lcms1.RData")
```

# Data Normalization

## Initial quality assesment

PCA: unsupervised visualization. We need to impute first all missing values
(replaced by random numbers from a uniform distribution from half the smallest
signal measured for a specific feature to its minimal value).

```{r Impute_MV}
#' Impute missing values using an uniform distribution
na_unidis <- function(z) {
    na <- is.na(z)
    if (any(na)) {
        min = min(z, na.rm = TRUE)
        z[na] <- runif(sum(na), min = min/2, max = min)
    }
    z
}

#' Row-wise impute missing values and add the data as a new assay
tmp <- apply(assay(res, "raw_filled"), MARGIN = 1, na_unidis)
assays(res)$raw_filled_imputed <- t(tmp)

```

Before plotting the PCA, we apply a log2 transform, center and scale the data.
The log2 transformation is applied to stabilize the variance while centering to
remove dependency on absolute abundances.

```{r PCA_all}
#| fig-cap: "PCA of all samples colored by device."
# General PCA: blanks, QCs and devices
#' Log2 transform and scale data

vals <- assay(res, "raw_filled_imputed") |>
    log2() |>
    t() |>
    scale(center = TRUE, scale = TRUE)

#' Perform the PCA
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

#' Plot the results
vals_st <- cbind(vals, device = res$device)
pca_12 <- autoplot(pca_res, data = vals_st , colour = 'device', scale = 0) +
    scale_color_manual(values = col_device)
pca_34 <- autoplot(pca_res, data = vals_st, colour = 'device',
                   x = 3, y = 4, scale = 0) +
    scale_color_manual(values = col_device)
grid.arrange(pca_12, pca_34, ncol = 1)
```

PCA: Blank separates from the samples. Plasma and Blood separates from devices.
PC1 represents 51.97% variance and PC2 20.46%. QCs are not as grouped as
expected.

(autoplot gives automatically colors. No need to define specific colors)

In this line, we are subsetting the res object to do not take into account
blanks and/or QCs.

i.e. this line should be for xcms exp object:: blank \<-
res\[sampleData(res)$device == "BLANK", keepFeatures = TRUE]
    this line should be for summarise exp object:: res_blank <- res_full[, res_full$device
== "blank"\] keep feature == TRUE: is to avoid that the description of the
features is removed. Only in xcms experiments.

```{r PCA_indexing}
# PCA without blank

## Data exploration and comparison
 
### We remove blank and QC samples from the study samples

res_blank <- res[, res$device == "BLANK"] #only blank
res_QC <- res[, res$sample_type == "QC"] #only QCs
res_sample <- res[, !res$device %in% c("BLANK", "QC")]  #samples without blanks or QCs
res_noblank <- res[, !res$device %in% c("BLANK")] #samples without blanks
 
#remove features that only have NA in the study samples
na_idx <- apply(assay(res, "raw_filled"), MARGIN = 1, function(x) all(is.na(x)))
res <- res[!na_idx, ]

```

Plot all the PCA to observe the data structure and grouping of the different
samples.

```{r PCA_samples, fig.height=5, fig.width=10}
#| echo: false
#| fig-cap: PCA of all samples without the blank and QCs colored by device.
#' Log2 transform and scale data
vals_sample <- assay(res_sample, "raw_filled_imputed") |>
    log2() |>
    t() |>
    scale(center = TRUE, scale = TRUE)
 
#' Perform the PCA
pca_res_sample <- prcomp(vals_sample, scale = FALSE, center = FALSE)
 
#' Plot the results
vals_st_sample <- cbind(vals_sample, device = res_sample$device)
pca_sample_12 <- autoplot(pca_res_sample, data = vals_st_sample , colour = 'device', scale = 0) +
    scale_color_manual(values = col_device)
pca_sample_34 <- autoplot(pca_res_sample, data = vals_st_sample, colour = 'device',
                   x = 3, y = 4, scale = 0) +
    scale_color_manual(values = col_device)
grid.arrange(pca_sample_12, pca_sample_34, ncol = 2)
```

```{r PCA_noblank, fig.height=5, fig.width=10}
#| echo: false
#| fig-cap: PCA of all samples without blanks colored by device.
#' Log2 transform and scale data
vals_noblank <- assay(res_noblank, "raw_filled_imputed") |>
    log2() |>
    t() |>
    scale(center = TRUE, scale = TRUE)
 
#' Perform the PCA
pca_res_noblank <- prcomp(vals_noblank, scale = FALSE, center = FALSE)
 
#' Plot the results
vals_st_noblank <- cbind(vals_noblank, device = res_noblank$device)
pca_noblank_12 <- autoplot(pca_res_noblank, data = vals_st_noblank , colour = 'device', scale = 0) +
    scale_color_manual(values = col_device)
pca_noblank_34 <- autoplot(pca_res_noblank , data = vals_st_noblank, colour = 'device',
                   x = 3, y = 4, scale = 0) +
    scale_color_manual(values = col_device)
grid.arrange(pca_noblank_12, pca_noblank_34, ncol = 2)
```

```{r PCA_inj_index, echo=TRUE, fig.height=5, fig.width=10}
#' check injection index only samples
vals_st_sample <- cbind(vals_sample, injection_index = res_sample$injection_index)

pca_sample_injidx_12 <- autoplot(pca_res_sample, data = vals_st_sample, colour = 'injection_index', scale = 0)
pca_sample_injidx_34 <- autoplot(pca_res_sample, data = vals_st_sample, colour = 'injection_index',
                   x = 3, y = 4, scale = 0)
grid.arrange(pca_sample_injidx_12, pca_sample_injidx_34, ncol = 2)

```

```{r PCA_inj_order_samples_QCs, echo=TRUE, fig.height=5, fig.width=10}
#' check injection index samples and QCs
vals_st_noblank <- cbind(vals_noblank, injection_index = res_noblank$injection_index)

pca_noblank_injidx_12 <- autoplot(pca_res_noblank, data = vals_st_noblank, colour = 'injection_index', scale = 0)
pca_noblank_injidx_34 <- autoplot(pca_res_noblank, data = vals_st_noblank, colour = 'injection_index',
                   x = 3, y = 4, scale = 0)
grid.arrange(pca_noblank_injidx_12, pca_noblank_injidx_34, ncol = 2)

```

```{r PCA_inj_order}
#' Add age to the PCA plot depending on age
vals_st_age <- cbind(vals_sample, age = res_sample$age)
pca_sample_age_12 <- autoplot(pca_res_sample, data = vals_st_age , colour = 'age', scale = 0)
pca_sample_age_12

```

```{r PCA_age, echo=TRUE, fig.height=5, fig.width=10}
#' Add age to the PCA plot depending on sex
vals_st_sex <- cbind(vals_sample, sex = res_sample$sex)
pca_sample_sex_12 <- autoplot(pca_res_sample, data = vals_st_sex , colour = 'sex', scale = 0)
grid.arrange(pca_sample_sex_12)

```

There is grouping depending on the type of sample. Inside groups it doesn't seem
to be sex separation between samples.


### Intensity evaluation

```{r Intensity_evaluation_all}
#| fig-cap: "Figure XX. Number of detected peaks and feature abundances."
layout(mat = matrix(1:3, ncol = 1), height = c(0.2, 0.2, 0.8))
par(mar = c(0.2, 4.5, 0.2, 3))

barplot(apply(assay(res, "raw"), MARGIN = 2, function(x) sum(!is.na(x))),
        col = paste0(col_sample, 80), border = col_sample,
        ylab = "# detected peaks", xaxt = "n", space = 0.012)
grid(nx = NA, ny = NULL)
barplot(apply(assay(res, "raw_filled"), MARGIN = 2, function(x) sum(!is.na(x))),
        col = paste0(col_sample, 80), border = col_sample,
        ylab = "# detected + filled peaks", xaxt = "n",
        space = 0.012)
grid(nx = NA, ny = NULL)
vioplot(log2(assay(res, "raw_filled")), xaxt = "n",
        ylab = expression(log[2]~feature~abundance),
        col = paste0(col_sample, 80), border = col_sample)
points(colMedians(log2(assay(res, "raw_filled")), na.rm = TRUE), type = "b",
       pch = 1)
grid(nx = NA, ny = NULL)
legend("topright", col = col_device,
       legend = names(col_device), lty=1, lwd = 2, xpd = TRUE, ncol = 3,
       cex = 0.8,  bty = "n")
```

```{r Intensity_evaluation_samples}
#| fig-cap: "Figure XX. Number of detected peaks and feature abundances (only Samples)."
layout(mat = matrix(1:3, ncol = 1), height = c(0.2, 0.2, 0.8))
par(mar = c(0.2, 4.5, 0.2, 3))

barplot(apply(assay(res_sample, "raw"), MARGIN = 2, function(x) sum(!is.na(x))),
        col = paste0(col_sampleonly, 80), border = col_sampleonly,
        ylab = "# detected peaks", xaxt = "n", space = 0.012, legend.text = NULL)
grid(nx = NA, ny = NULL)
barplot(apply(assay(res_sample, "raw_filled"), MARGIN = 2, function(x) sum(!is.na(x))),
        col = paste0(col_sampleonly, 80), border = col_sampleonly,
        ylab = "# detected + filled peaks", xaxt = "n",
        space = 0.012)
grid(nx = NA, ny = NULL)
vioplot(log2(assay(res_sample, "raw_filled")), xaxt = "n",
        ylab = expression(log[2]~feature~abundance),
        col = paste0(col_sampleonly, 80), border = col_sampleonly)
points(colMedians(log2(assay(res_sample, "raw_filled")), na.rm = TRUE), type = "b",
       pch = 1)
grid(nx = NA, ny = NULL)
legend("topright", col = col_device,
       legend = names(col_device), lty=1, lwd = 2, xpd = TRUE, ncol = 3,
       cex = 0.8,  bty = "n")
```

```{r Intensity_evaluation_samples_QCs}
#| code-fold: true
#| code-summary: Show the code
#| fig-cap: Figure XX. Number of detected peaks and feature abundances (SAMPLES AND QCs).
layout(mat = matrix(1:3, ncol = 1), height = c(0.2, 0.2, 0.8))
par(mar = c(0.2, 4.5, 0.2, 3))

barplot(apply(assay(res_noblank, "raw"), 
    MARGIN = 2, function(x) sum(!is.na(x))),
        col = paste0(col_noblank, 80), border = col_noblank,
        ylab = "# detected peaks", xaxt = "n", space = 0.012, legend.text = NULL)
grid(nx = NA, ny = NULL)
barplot(apply(assay(res_noblank, "raw_filled"), 
    MARGIN = 2, function(x) sum(!is.na(x))),
        col = paste0(col_noblank, 80), border = col_noblank,
        ylab = "# detected + filled peaks", xaxt = "n",
        space = 0.012)
grid(nx = NA, ny = NULL)
vioplot(log2(assay(res_noblank, "raw_filled")), xaxt = "n",
        ylab = expression(log[2]~feature~abundance),
        col = paste0(col_noblank, 80), border = col_noblank)
points(colMedians(log2(assay(res_noblank, "raw_filled")), na.rm = TRUE), 
    type = "b",
       pch = 1)
grid(nx = NA, ny = NULL)
legend("topright", col = col_device,
       legend = names(col_device), lty=1, lwd = 2, xpd = TRUE, ncol = 3,
       cex = 0.8,  bty = "n")
```

### CV values

```{r CV_values}
#' Calculate the CV values
#index_study <- res$device %in% c("CTR", "CVD")
index_QC <- res_noblank$device == "QC"
cv_qc <- rowRsd(assay(res_noblank, "raw_filled")[, index_QC],
                    na.rm = TRUE, mad = TRUE)

#' Summarize the values across features
res_df <- quantile(cv_qc, na.rm = TRUE)

kable(res_df, format = "pipe")
```

### RLA

An RLA value is the abundance of a feature in a sample relative to the median
abundance of the same feature across multiple samples.

-   Within group RLA plots: assess tightness of replicates within groups should
    have a median close to zero and low variation around it.
-   across groups, they allow to compare behavior between groups

```{r RLA-log-abundance-all}
par(mfrow = c(1, 1), mar = c(3.5, 4.5, 2.5, 1))
boxplot(MsCoreUtils::rowRla(assay(res, "raw_filled"),
                            f = res$device, transform = "log2"),
        cex = 0.5, pch = 16,
        col = paste0(col_sample, 80), ylab = "RLA",
        border = col_sample, boxwex = 1,
        outline = FALSE, xaxt = "n", main = "Relative log abundance",
        cex.main = 1)
axis(side = 1, at = seq_len(ncol(res)), labels = colData(res)$sample_name)
grid(nx = NA, ny = NULL)
abline(h = 0, lty=3, lwd = 1, col = "black")
legend("topright", col = col_device,
       legend = names(col_device), lty=1, lwd = 2, xpd = TRUE, ncol = 3,
       cex = 0.8,  bty = "n")

#change legend for sample_name
```

```{r RLA-log-abundance-no-blank}
par(mfrow = c(1, 1), mar = c(3.5, 4.5, 2.5, 1))
boxplot(MsCoreUtils::rowRla(assay(res_noblank, "raw_filled"),
                            f = res_noblank$device, transform = "log2"),
        cex = 0.5, pch = 16,
        col = paste0(col_noblank, 80), ylab = "RLA",
        border = col_noblank, boxwex = 1,
        outline = FALSE, xaxt = "n", main = "Relative log abundance",
        cex.main = 1)
axis(side = 1, at = seq_len(ncol(res_noblank)), 
    labels = colData(res_noblank)$sample_name)
grid(nx = NA, ny = NULL)
abline(h = 0, lty=3, lwd = 1, col = "black")
legend("topright", col = col_device,
       legend = names(col_device), lty=1, lwd = 2, xpd = TRUE, ncol = 3,
       cex = 0.8,  bty = "n")

```

On the RLA plot above, we can see that median is not centered around 0 after
QC07. This can be associated to the sensibility of the instrument and detector
value optimisation across the analysis (is it worthy to include "autogain" in
method). Drift correction? Technical issues? also, mainly with Plasma samples
there is more variability.

### Noise Comparison

Below we compare the noise signals between devices. We first calculate the
overall signal in the dataset and then calculate the signal that is in the
chromatographic peaks detection. We then subtract the two to get the noise
signal.

```{r Noise_comparison, echo=FALSE}
# overall signal in the dataset 
#' - for each file calculate the sum of intensities 
background  <- spectra(lcms1[!idx_blank_qc]) |>
    split(fromFile(lcms1[!idx_blank_qc])) |>
    lapply(tic) |>
    lapply(sum) |>
    unlist()

# Overall signal that is in the chromatographic peaks detection 
detected <- apply(assay(res_sample), 2, function(x) sum(x, na.rm = TRUE))
 
# substract and plot
names(background) <- names(detected) <- res_sample$device
noise <- background - detected
 
f <- factor(names(noise), levels = unique(names(noise)))
group <- split(noise, f)
 
plot(NULL, xlim = c(1, length(group)), ylim = range(unlist(group)), 
     xaxt = "n", xlab = "Devices", ylab = "Noise", 
     main = "Noise comparison between Devices")
for (i in seq_along(group)) {
  points(rep(i, length(group[[i]])), group[[i]], pch = 19, 
col = col_sampleonly[i])
}
axis(1, at = seq_along(group), labels = names(group))
```

## Between sample normalisation

Between-sample normalization use filled-in features to effectively mitigate
variations influenced by technical issues, such as differences in sample
preparation, processing and injection methods. In this instance, we will employ
a commonly used technique known as median scaling.

### Median Scaling

This method involves computing the median for each sample, followed by
determining the median of these individual sample medians. This ensures
consistent median values for each sample throughout the entire data set.
Maintaining uniformity in the average total metabolite abundance across all
samples is crucial for effective implementation.

This process aims to establish a shared baseline for the central tendency of
metabolite abundance, mitigating the impact of sample-specific technical
variations. This approach fosters a more robust and comparable analysis of the
top features across the data set. The assumption is that normalizing based on
the median, known for its lower sensitivity to extreme values, enhances the
comparability of top features and ensures a consistent average abundance across
samples.

```{r normalization-factor-all}
#' Compute median and generate normalization factor
mdns <- apply(assay(res, "raw_filled"), MARGIN = 2,
              median, na.rm = TRUE )
nf_mdn <- mdns / median(mdns)

#' divide dataset by median of median and create a new assay.
assays(res)$norm <- sweep(assay(res, "raw_filled"), MARGIN = 2, nf_mdn, '/')
assays(res)$norm_imputed <- sweep(assay(res, "raw_filled_imputed"), MARGIN = 2,
                                  nf_mdn, '/')

```

```{r normalization-factor-noblank}
#' Compute median and generate normalization factor (NO BLANKS)
mdns_noblank <- apply(assay(res_noblank, "raw_filled"), MARGIN = 2,
              median, na.rm = TRUE )
nf_mdn_noblank <- mdns_noblank / median(mdns_noblank)

#' divide dataset by median of median and create a new assay.
assays(res_noblank)$norm <- sweep(assay(res_noblank, "raw_filled"), MARGIN = 2, nf_mdn_noblank, '/')
assays(res_noblank)$norm_imputed <- sweep(assay(res_noblank, "raw_filled_imputed"), MARGIN = 2,
                                  nf_mdn_noblank, '/')
```

### Assessing overall effectiveness of the normalisation approach

#### Principal component analysis

The evaluation will be performed by comparing the distribution of the log2
transformed feature abundances before and after normalization.

```{r PCA-normalisation, fig.height=7, fig.width=7}
#| echo: false
#| fig-cap: PCA after normalization of all samples colored by device.

#' Data before normalisation
vals_st <- cbind(vals, device = res$device)
pca_raw <- autoplot(pca_res, data = vals_st,
                    color = 'device', scale = 0) +
    scale_color_manual(values = col_device) +
    labs (title = "PCA before Normalization", tag = "A")
#' Data after normalization
vals_norm <- apply(assay(res, "norm"), MARGIN = 1, na_unidis) |>
    log2() |>
    scale(center = TRUE, scale = TRUE)

pca_res_norm <- prcomp(vals_norm, scale = FALSE, center = FALSE)
vals_st_norm <- cbind(vals_norm, device = res$device)
pca_adj <- autoplot(pca_res_norm, data = vals_st_norm,
                    color = 'device', scale = 0) +
    scale_color_manual(values = col_device) +
    labs (title = "PCA before Normalization", tag = "B")
grid.arrange(pca_raw, pca_adj, ncol = 1)
```

Now all devices group together, and PCA2 change (plasma -\> devices, now devices
-\>plasma).

```{r PCA-normalisation-noblank, fig.height=7, fig.width=7}
#| echo: false
#| fig-cap: PCA after normalization of all samples without blanks colored by device.

#' Data without blanks before normalisation
vals_noblank_st<- cbind(vals_noblank, device = res_noblank$device)
pca_noblank_raw <- autoplot(pca_res_noblank, data = vals_noblank_st,
                    color = 'device', scale = 0) +
    scale_color_manual(values = col_device) + 
    labs (title = "PCA before Normalization (no blank)", tag = "A")

#' Data without blanks after normalization
vals_noblank_norm <- apply(assay(res_noblank, "norm"), MARGIN = 1, na_unidis) |>
    log2() |>
    scale(center = TRUE, scale = TRUE)
pca_res_noblank_norm <- prcomp(vals_noblank_norm, scale = FALSE, center = FALSE)


vals_noblank_st_norm <- cbind(vals_noblank_norm, device = res_noblank$device)
pca_noblank_adj <- autoplot(pca_res_noblank_norm, data = vals_noblank_st_norm,
                    color = 'device', scale = 0) +
    scale_color_manual(values = col_device) +
    labs (title = "PCA before Normalization (no blank)", tag = "B")

grid.arrange(pca_noblank_raw, pca_noblank_adj, ncol = 1)
```

Normalisation have an impact in PC1 improving the grouping of QCs, Plasma and
Blood samples.

#### Intensity Evaluation (RLA)

```{r RLA-after-norm, fig.height=7, fig.width=7}
#| fig.cap = "RLA before and after normalisation"

layout(mat = matrix(1:2, ncol = 1), height = c(0.9,0.9))

par(mfrow = c(2, 1), mar = c(2, 4.5, 2.5, 1))
# A RLA Raw data
boxplot(rowRla(assay(res_noblank, "raw_filled"), group = res_noblank$device),
        cex = 0.5, pch = 16, ylab = "RLA", border = col_device,
        col = paste0(col_device, 80), cex.main = 1, outline = FALSE,
        xaxt = "n", main = "Raw data", boxwex = 1)
grid()
abline(h = 0, lty = 3, lwd = 1, col = "black")
legend("topright", col = col_device,
       legend = names(col_device), lty=1, lwd = 2, xpd = TRUE, ncol = 3,
       cex = 0.6,  bty = "n")

# B RLA Normalized data
boxplot(rowRla(assay(res_noblank, "norm"), group = res_noblank$device),
        cex = 0.5, pch = 16, ylab = "RLA", border = col_device,
        col = paste0(col_device, 80), boxwex = 1, outline = FALSE,
        xaxt = "n", main = "Normalised data", cex.main = 1)

#axis(side = 1, at = seq_len(ncol(res_noblank)), labels = res_noblank$sample_name)

grid()
abline(h = 0, lty = 3, lwd = 1, col = "black")
legend("topright", col = col_device,
       legend = names(col_device), lty=1, lwd = 2, xpd = TRUE, ncol = 3,
       cex = 0.6,  bty = "n")


```

NOTE: the colors of this plots are not corrected. Need to fix it. 1st samples
should be yellow (QC). METABONAUT: typo normalized title. NOTE: fix legend and
colors in RLA plot

#### CV after normalisation

```{r CV-after-norm}
#' Calculate the CV values
index_samples <- res_noblank$device %in% c("WHATMAN", "MITRA", "CAPITAINER", "PLASMA", "BLOOD")
index_QC <- res_noblank$device == "QC"

sample_res <- cbind(
    QC_raw = rowRsd(assay(res_noblank, "raw_filled")[, index_QC],
                    na.rm = TRUE, mad = TRUE),
    QC_norm = rowRsd(assay(res_noblank, "norm")[, index_QC],
                     na.rm = TRUE, mad = TRUE),
    Study_raw = rowRsd(assay(res_noblank, "raw_filled")[, index_samples],
                       na.rm = TRUE, mad = TRUE),
    Study_norm = rowRsd(assay(res_noblank, "norm")[, index_samples],
                        na.rm = TRUE, mad = TRUE)
)

#' Summarize the values across features
res_noblank_df <- data.frame(
    QC_raw = quantile(sample_res[, "QC_raw"], na.rm = TRUE),
    QC_norm = quantile(sample_res[, "QC_norm"], na.rm = TRUE),
    Study_raw = quantile(sample_res[, "Study_raw"], na.rm = TRUE),
    Study_norm = quantile(sample_res[, "Study_norm"], na.rm = TRUE)
)

kable(res_noblank_df, format = "pipe")
```

NOTE: Are CVs not only useful if apply to QCs?
