---
title: "Data analysis of Pilot Study from Imperial - annotation_LPOS_CONV."
format: html
tbl-cap-location: bottom
editor: visual
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
author: Philippine Louail, Ana Sanchez Lorenzo
---

```{r setup, include=FALSE}
library(knitr)
library(quarto)
knitr::opts_knit$set(root.dir = './')
```

# Loading packages

```{r packages-used, message=FALSE, warning=FALSE}
#| code-fold: true
#| code-summary: "Show the code"
## General bioconductor package
library(BiocStyle)
library(BiocParallel)
library(knitr)

## Data Import and handling
library(readxl)
library(MsExperiment)
library(writexl)
#library(MsIO)
#library(alabaster.se)
library(SummarizedExperiment)

## Preprocessing of LC-MS data
library(xcms)
library(Spectra)
library(MetaboCoreUtils)

## Statistical analysis
library(limma) # Differential abundance
library(matrixStats) # Summaries over matrices

## Visualisation
library(pander)
library(RColorBrewer)
library(pheatmap)
library(vioplot)
library(ggfortify)   # Plot PCA
library(gridExtra)   # To arrange multiple ggplots into single plots
library(grid)
library(UpSetR)

## Annotation
library(AnnotationHub) # Annotation resources
library(CompoundDb)    # Access small compound annotation data.
library(MetaboAnnotation) # Functionality for metabolite annotation.
```

# DATA ORGANISATION

## Data import

In this case, we only import the SumExp with significant features details
(res_sts).

```{r import, message=FALSE}
# Data import
load("preprocessed_lcms1.RData")
load("res_sts.RData")
featureDefinitions(lcms1) |>head()
```

```{r indexing, message=FALSE}
# Index per sample group
idx_qc <- sampleData(lcms1)$device == "QC"
idx_blank <- sampleData(lcms1)$device == "BLANK"
idx_blank_qc <- sampleData(lcms1)$device %in% c("QC","BLANK")
idx_samples <- sampleData(lcms1)$device %in% c("WHATMAN", "MITRA", "CAPITAINER", "PLASMA", "BLOOD")
idx_bl_sp <- sampleData(lcms1)$blank_type %in% c("BLANK_W", "BLANK_C", "BLANK_M")
```

```{r par-process}#' Set up parallel processing using 3 cores}
if (.Platform$OS.type == "unix") {
  register(MulticoreParam(4)) ### chunksize has to be be the same number. 
 } else {
     register(SnowParam(4))
}
```

## Define colors

We will define colors for each of the sample groups based on their sample group
using the *RColorBrewer* package:

```{r define-colors}
#| echo: true
#' Define colors for full dataset
leg_full <- brewer.pal(12, name = "Set3")[c(1, 12, 7, 5, 10, 4, 6)]
names(leg_full) <- unique(sampleData(lcms1)$device)
leg_full["QC"] <- "#f00e0e"

#' define colors for devices only
leg_blank_BuS <- leg_full[-c(1, 2, 6, 7)]
col_blank_BuS <- leg_blank_BuS[sampleData(lcms1[!idx_bl_sp])$device]
leg_blank_BuS <- leg_blank_BuS[c("WHATMAN", "MITRA", "CAPITAINER")]
```

## Participants column addition For Differential abundance analysis

We need to specify participants. For that we need to add a new column from
sample_name column information in lcms1 object and res object.

```{r}
#' Create a new column with participants names without device specification.
sampleData(lcms1)$participants <- substr(sampleData(lcms1)$sample_name, 1, nchar(sampleData(lcms1)$sample_name) -1)
sampleData(lcms1)$participants[idx_blank_qc] <- NA_character_
colData(res_sts)$participants <- substr(colData(res_sts)$sample_name, 1, nchar(colData(res_sts)$sample_name) - 1)
```

# ANNOTATION

## MS1 Annotation

After the differential abundance analysis, the next step is to identify the
features with significant differences in abundances between plasma and the rest
of groups. As the data was acquired using the NPC LPOS method, we could target
the compounds from the NPC library. However, as LC set-up was different
(different sample manager and different tubing length), only m/z values will be
able to match, fnot rt values.

To do the matching of m/z values, MetaboAnnotation package is used. This is the
lowest level of confidence in metabolite annotation. However, it gives ideas
about of metabolites present in samples.

To start with, I imported the library of metabolites called targed_df.

```{r import-ms1-features}
# Import NPC-LPOS library from csv file
target_df <- read.csv("LPOS_ROI_V_5_1_2.csv")
target_df <- as.data.frame(target_df)  # Ensure it is a data frame
head(target_df)  
```

As we know there is a shift in the rt due to the LC set up and tube length I
would change the rt column of the library for this specific dataset. If I check
the rt of "known compounds" EIC there is a shift +30 seconds in comparison with
the rt from the library. However, I will increase the rt range in case there are
some differences between features.

```{r}
# Modify column rt ranges from library
colnames(target_df)
# fit rt ranges
target_df$rtMin <- target_df$rtMin + 30 
target_df$rtMax <- target_df$rtMax + 30 
target_df$rt <- target_df$rt + 30
target_df$rt_minutes <- target_df$rt_minutes + 0.5
#target and query need to have same rt column name
target_df$rtmed <- target_df$rt 
# check it is working
head(target_df)
```

With this modification, I am trying to do MS1 annotation with m/z and rt values.

First, I prepared the data.frame with experimental features detected:

```{r query-df}
# Prepare query data frame
rowData(res_sts)$feature_id <- rownames(rowData(res_sts))
#cont with full dataset
rowData(res_sts)
# Convert to data frame
res_sts_df <- as.data.frame(rowData(res_sts))
res_sts_df
```

Now, we do the matching in comparison with the target dataframe with reference
known compounds. As we already have the ion description in the library, we
include those adducts.

The resulting Matched object (with 10 ppm) with mz and rt values shows that 177
out of 2454 significant features were matched against at least one target
compound. Out of the 378 identify ions in NPC library, we are able to match 159
features. If we try to do MS1 annotation matching m/z abd retention time values,
we expect the match to be more accurate because rt will restrict the match and
if features are detected across the analysis it will only identify the features
that matched with biological compounds:

```{r}
parm <- Mass2MzRtParam(adducts = c("[M+H]+", "[M+Na]+", "[M+NH4]+", 
                                   "[M+H-H2O]+",  "[M+K]+", "[M]+"),
                       tolerance = 0, ppm = 10,
                       toleranceRt = 10) # here will be to change rt tolerance(sec). 
mtch <- matchValues(res_sts, target_df, param = parm,
                                rtColname = "rtmed", mzColname = "mzmed", 
     massColname = "Monoisotopic_mass")
mtch
```

```{r}
#added
parm_mzrt <- MzRtParam(tolerance = 0, ppm = 10,
                       toleranceRt = 10) # here will be to change rt tolerance(sec). 
mtch_mzrt <- matchValues(res_sts, target_df, param = parm_mzrt,
                    rtColname = "rtmed", mzColname = c("mzmed", "mz"))
mtch_mzrt
```

```{r}
#' Extracting the results
mtch_res_2 <- matchedData(mtch_mzrt, c("feature_id", "mzmed", "rtmed", 
                                "target_ion", "ppm_error", 
                                "target_chemicalFormula", 
                                "target_cpdName", "target_InChI", 
                                "coef.W", "coef.M", "coef.C", "coef.B", 
                                "adjp.W", "adjp.M", "adjp.C", "adjp.B" ))

mtch_res_2
mtch_res_2_df <- as.data.frame(mtch_res_2)
write_xlsx(mtch_res_2_df, "mtch_res_2_output.xlsx")
#filter only present in query. 
```

-   check biocrates annotation
-   MS1: manual curation. Do Ms2 individually.

After matching mz and rt, as expected, we have less matches with the target
object (159 at of 378 annotations; only mz = 226/378). However, it is finding
same compound at different rts in query object. In total, we will have 245
annotations but we will have not manually check those ones. I will continue the
annotation with this matched object. NOTE: the res object has not being filter
by only significant. I doing with all dataset. After general annotation, filter
by features of interest depending if it present or not in each device.

The result object (mtch) contains the full query data.frame and target
data.frame as well as the matching information. To visualize the info from mtch:

```{r visualisation1}
head(query(mtch))
```

```{r visualisation2}
# head(target(mtch))
```

```{r visualisation3}
# whichQuery(mtch)
```

```{r visualisation4}
# whichTarget(mtch)
```

```{r visualisation5}
# colnames(mtch)
```

```{r extraction-match}
# To extract only the object of the matched features
mtch <- mtch[whichQuery(mtch)]
matchedData(mtch)|> head()
nrow(matchedData(mtch))
```

```{r visualisation5}
mtch$target_cpdName
```

Below I extract the full result from the Matched object:

```{r extration-match}
#' Extracting the results
mtch_res <- matchedData(mtch, c("feature_id", "mzmed", "rtmed", "adduct", 
                                "target_ion", "ppm_error", 
                                "target_chemicalFormula", 
                                "target_Monoisotopic_mass", 
                                "target_cpdName", "target_InChI", 
                                "coef.W", "coef.M", "coef.C", "coef.B", 
                                "adjp.W", "adjp.M", "adjp.C", "adjp.B" ))

mtch_res
```

\[from here it is from metanonaut\]

To reduce redundancy in the result table we below iterate over the hits of a
feature and keep only matches to unique compounds (identified by their
INCHIKEY). The INCHI name combine information from compound’s chemical formula
and structure, so while different compounds can share the same chemical formula,
they should have a different structure and thus INCHI.

```{r}
rownames(mtch_res) <- NULL

#' Keep only info on features that machted - create a utility function for that
mtch_res <- split(mtch_res, mtch_res$feature_id) |>
    lapply(function(x) {
        lapply(split(x, x$target_InChI), function(z) {
            z[which.min(z$ppm_error), ]
        })
    }) |>
    unlist(recursive = FALSE) |>
    do.call(what = rbind)

#' Display the results
kable(mtch_res, format = "pipe")

nrow(mtch_res)
mtch_res <- cbind(mtch_res, rowData(res_sts)[mtch_res$feature_id, 
                                             c("coef.W"
                                               ,"coef.M", "coef.C",
                                                "coef.B",
                                               "adjp.W", "adjp.M", "adjp.C",
                                                "adjp.B")])
mtch_res
nrow(mtch_res)
```

```{r}
# Export file for better visualisation
# Make sure it is a dataframe:
# mtch_res_df <- as.data.frame(mtch_res)
# write_xlsx(mtch_res_df, "mtch_res_output.xlsx")
```

Only 195 matches were unique features. However, now adducts to not match with
the proposed adduct in the library. I am filtering again to keep only the
annotations that have the adduct from the library. For this I am using
"target_ion" to show the adduct from library and "adduct" as the m/z calculated
for the match. NOTE: Does it make sense?

```{r}
# Filter rows where 'adduct' and 'target_ion' are the same
mtch_res_filtered <- mtch_res[mtch_res$adduct == mtch_res$target_ion, ] 
nrow(mtch_res_filtered)
```

Only 77 compounds have the same ion.

```{r}
# Transform in df and export for visualisation: 
# mtch_res_filtered_df <- as.data.frame(mtch_res_filtered)
# write_xlsx(mtch_res_filtered_df, "mtch_res_filtered_output.xlsx")
```

FIX: add names of all sign features. While being of low confidence, MS1-based
annotation can provide first candidate annotations that could be confirmed or
rejected by additional analyses.

## MS2 Annotation

LC MS/MS data were recorded in data depending acquisition (DDA) in pooled
samples (and QCs samples). However, method was not fully optimised and it seems
MS2 data points will not be really abundance in those samples. I would like to
test the data to see how do I need to modify the method.

The MS2 data was acquited in 7 different samples:

-   pool sample from each device (Individual 11C, 11W, 11M)

-   pool general sample (QC)

-   pool sample from each device (QCC, QCW, QCM)

In total we have 4 measurements from each samples with different mz ranges:

-   001: mz range from 50 to 400 Da

-   002: mz range from 400 to 800 Da

-   003: mz range from 800 to 1200 Da

-   004: mz range from 50 to 1200 Da.

This was performed because we experienced a mass sensibility issue in the
instrument and by decreasing the mz range sensibility can increase.

After testing different samples, we decided to continue with all samples as we
will obtain the maximum number of possible compounds fragmentated.

```{r import-files}
#' read the sample descriptions from an xlsx sheet
pd_MS2 <- read_xlsx("metadata_msms_lcms_rpos.xlsx", sheet = "samples") |>
    as.data.frame()

#' Import MS2 files
MS2_mzml_path <- file.path("data/lpos_conv_DDA") 

lcms2 <- readMsExperiment(file.path(MS2_mzml_path, "/",pd_MS2$mzML_file),
                          sampleData = pd_MS2)
lcms2
```

```{r}
# Add mzrange data information (from file names)
sampleData(lcms2)$fragmentation_mode <- c("001", "002", "003", "004")

#' Subset only samples with mz range form 50-1200Da. 
#' NEW: Continue with full dataset MS2
# idx_mzrange <- sampleData(lcms2)$fragmentation_mode == "004"
# lcms2_004 <- lcms2[idx_mzrange]

#let's look at the updated sample data
sampleData(lcms2)
colnames(sampleData(lcms2)) 
sampleData(lcms2) |> kable (format = "pipe")
```

MS2 Data is filtered to the same RT range as LC-MS run to clean unnecessary
peaks that will not be of interest

```{r}
#' Filter the data to the same RT range as the LC-MS run
lcms2 <- filterRt(lcms2, c(60, 700))
```

Below we show the number of MS1 and MS2 spectra for each of the files.

```{r}
#' check the number of spectra per ms level
spectra(lcms2) |>
    msLevel() |>
    split(spectraSampleIndex(lcms2)) |>
    lapply(table) |>
    do.call(what = cbind)
```

Mainly all samples have similar number of spectra in MS1 and MS2, which is not
normal. We only see 4 samples with big difference (\>800 scans): 3 (MS1 2087/
MS2 1353), 15 (MS1 2538, MS2 938), 19 (MS1 2246, MS2 1060), 27 (MS1 1974, MS2
1174) . This is associated with instrument and method problem.

Actually, all samples from m/z range 800-1200Da have less scans than the others.
IMPORTANT!

LAB: modify and test MS2 method.

```{r}
sampleData(lcms2)$sample_name
```

In the plot below we can observe the different mz ranges in each sample
measurement.

```{r eval = FALSE}
#|fig: "Figure 01. Precursor ions in the LC-MS/MS data set."

dr <- "PrecursorPlots/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)

# Get number of samples
ms2_samples <- nrow(sampleData(lcms2))

# Save one PNG per sample
for (i in 1:ms2_samples) {
    png(filename = paste0(dr, "Precursor_", sampleData(lcms2)$sample_name[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plotPrecursorIons(lcms2[i, ],
                      cex = 0.5,
                      xlab = "retention time (s)",
                      ylab = "m/z ratio",
                      main = sampleData(lcms2)$sample_name[i])
    
    dev.off()
}
```

NOTE: We observed a high amount of low mass molecules and few high mass
molecules. All this precursor ions should have MSMS data or they are MS1 ions?

We next extract the Spectra object with the MS data from the data object and
assign a new spectra variable with the employed collision energy, which we
extract from the data object sampleData.

For the next steps MS data is filtered by:

-   restricting MS2 spectra

-   removing mass peaks from each spectrum with an intensity that is lower than
    5% of the highest intensity for that spectrum, assuming that these very low
    intensity peaks represent background signal.

-   removing also mass peaks with an m/z value greater or equal to the precursor
    m/z of the ion. This puts, for the later matching against reference spectra,
    more weight on the fragmentation pattern of ions and avoids hits based only
    on the precursor m/z peak (and hence a similar mass of the compared
    compounds).

-   restricting the data to spectra with at least two fragment peaks and scale
    their intensities such that their sum is 1 for each spectrum.

While similarity calculations are not affected by this scaling, it makes visual
comparison of fragment spectra easier to read.

```{r}
#' extract spectra object
ms2 <- spectra(lcms2)

#' Remove low intensity peaks
low_int <- function(x, ...) {
    x > max(x, na.rm = TRUE) * 0.05
}
ms2 <- filterMsLevel(ms2, 2L) |>
    filterIntensity(intensity = low_int) 

#' Remove precursor peaks and restrict to spectra with a minimum number of peaks
ms2 <- filterPrecursorPeaks(ms2, ppm = 50, mz = ">=")
ms2 <- ms2[lengths(ms2) > 1] |>
    scalePeaks()
```

```{r}
#' Load the MS data subset into memory
ms2 <- setBackend(ms2, MsBackendMemory())
ms2 <- applyProcessing(ms2)

#' Size of the object after loading into memory
print(object.size(ms2), units = "MB")
```

The goal of the workflow is to *identify the MS2 spectra* from the second
(LC-MS/MS) run that could represent fragments of the ions of features from the
data in the first (LC-MS) run.

Our approach is to match MS2 spectra against the *significant features*
determined earlier based on their precursor m/z and retention time (given an
acceptable tolerance) to the feature’s m/z and retention time. This can be
easily done using the featureArea() function that effectively considers the
*actual m/z and retention time ranges of the features’ chromatographic peaks*
and therefore increase the chance of finding a correct match. This, however,
also assumes that retention times between the first and second run don’t differ
by much. Alternatively, we would need to align the retention times of the second
LC-MS/MS data set to those of the first.

```{r}
#' Define the m/z and retention time ranges for the significant features
target <- featureArea(lcms1)[rownames(res_sts),] |> as.data.frame()
target$rtmin <- target$rtmin - 2
target$rtmax <- target$rtmax + 2
```

NOTE: here i am not selecting sign feature, there is only a column with info
saying if it is significant or not.

Here is were target should get all data related with sig features. We next
*identify the fragment spectra* with their precursor m/z and retention times
within these ranges. We use the filterRanges() function that allows to filter a
Spectra object using multiple ranges simultaneously. We apply this function
separately to each feature (row in the above matrix) to extract the MS2 spectra
representing fragmentation information of the presumed feature’s ions.

```{r}
#' Identify for each feature MS2 spectra with their precursor m/z and
#' retention time within the feature's m/z and retention time range
ms2_fts <- apply(target[, c("rtmin", "rtmax", "mzmin", "mzmax")],
                     MARGIN = 1, FUN = filterRanges, object = ms2,
                     spectraVariables = c("rtime", "precursorMz"))
# fid <- rep(names(ms2_fts), lengths(ms2_fts))
# 
# lengths(ms2_fts)
# ms2_fts <- concatenateSpectra(ms2_fts)
# ms2_fts$feature_id <- fid
# 
# table(ms2_fts$feature_id)
# Add feature_id to spectra

# ms2_fts_present<- ms2_fts[lengths(ms2_fts) >= 1]
# lengths(ms2_fts_present)
# count(lengths(ms2_fts_present))

```

The result from this apply() call is a *list of Spectra*, each element
representing the result for one feature. With the exception of the last feature,
multiple MS2 spectra could be identified. We next combine the list of Spectra
into a single Spectra object using the *concatenateSpectra() function* and add
an additional spectra variable containing the respective feature identifier.

```{r eval=!file.exists("spectra_significant_fts.RData")}
l <- lengths(ms2_fts)
#' Combine the individual Spectra objects
ms2_fts <- concatenateSpectra(ms2_fts) 
lengths(ms2_fts)
#' Assign the feature identifier to each MS2 spectrum
ms2_fts$feature_id <- rep(rownames(res_sts), l)

## Save for reuse in other vignettes
 save(ms2_fts, file = "spectra_significant_fts.RData") 
```

### GNPS

I first import the GNPS library for MS2 spectra matching.

```{r}
#' Load GNPS library. 
library(RSQLite)
library(MsBackendSql)
dbfile <- "MsBackendSql.GNPS.matchms.cleaned.v1.sqlite"
con <- dbConnect(SQLite(), dbfile)
mb <- Spectra(con, source = MsBackendSql())
```

```{r}
#' Filter data for positive ionisation, low intensity fragments, and fragments 
#' with higher mz than precursor peak.

ms2_GNPS_ref <- mb |>
    filterPolarity(1L) |>
    filterIntensity(intensity = low_int) |>
    filterPrecursorPeaks(ppm = 50, mz = ">=")
ms2_GNPS_ref <- ms2_GNPS_ref[lengths(ms2_GNPS_ref) > 1] |>
    scalePeaks()
# To visualise the GNPS db
# spectraVariables(ms2_GNPS_ref)
# table(ms2_GNPS_ref$ms_dissociation_method) # filter for cid. 
# table(ms2_GNPS_ref$ms_manufacturer)
#ms2_GNPS_ref_filt <- spectraData(mb)$ms_mass_analyzer == "qtof" # how can i
#filter the data? 

```

We use the matchSpectra() function from the MetaboAnnotation package with the
CompareSpectraParam to define the settings for the matching.

-   `requirePrecursor = TRUE`: Limits spectra similarity calculations to
    fragment spectra with a similar precursor *m/z*.

-   `tolerance` and `ppm`: Defines the acceptable difference between compared
    *m/z* values. These relaxed tolerance settings ensure that we find matches
    even if reference spectra were acquired on instruments with a lower
    accuracy.

-   `THRESHFUN`: Defines which matches to report. Here, we keep all matches
    resulting in a spectra similarity score (calculated with the normalized dot
    product), the default similarity function) larger than 0.6.

```{r}
register(SerialParam())
#' Define the settings for the spectra matching.
prm <- CompareSpectraParam(ppm = 40, tolerance = 0.05,
                           requirePrecursor = TRUE, 
                           THRESHFUN = function(x) which(x >= 0.6)) #similarity btw normalised dot products. Check. 

# Question: x = we changed it to 0.6? what is it doing exactly? review. 
ms2_GNPS_mtch <- matchSpectra(ms2_fts, ms2_GNPS_ref, param = prm)
ms2_GNPS_mtch
ms2_GNPS_mtch <- pruneTarget(ms2_GNPS_mtch) # ???
ms2_GNPS_mtch@target <- setBackend(ms2_GNPS_mtch@target, MsBackendMemory()) #???
# save(ms2_GNPS_mtch, file = "matchedGNPS_result.RData")
```

From the total 3115 query MS2 specta, there were 868 matching spectra with the
library. Below we restrict the results to these matching spectra and extract all
metadata of the query and target spectra as well as the similarity score.

```{r}
#' Keep only query spectra with matching reference spectra
ms2_GNPS_mtch <- ms2_GNPS_mtch[whichQuery(ms2_GNPS_mtch)]

#' Extract the results
ms2_GNPS_mtch_res <- matchedData(ms2_GNPS_mtch)
nrow(ms2_GNPS_mtch_res)
ms2_GNPS_mtch_res$target_compound_name |> head()
```

After restricting the data to the matching spectra, 279 Ms2 spectra could
matched at least one reference fragment.

```{r}
#' Complete list of available metadata information
# colnames(ms2_GNPS_mtch_res) # in GNPS column name is target_compound_name
```

The annotations seem to be repeated in most of the cases. So I will need to
filter so that I only have one annotation to each feature define.

Now, we have query-target pairs with a spectra similarity higher than 0.6.

Below we thus iterate over the feature-compound pairs and select for each the
one with the highest score. As an identifier for a compound, we use its fragment
spectra’s INCHI-key, since compound names in GNPS do not have accepted
consensus/controlled vocabularies.

```{r}
#' - split the result per feature
#' - select for each feature the best matching result for each compound
#' - combine the result again into a data frame
ms2_GNPS_mtch_res <-
    ms2_GNPS_mtch_res |>
    split(f = paste(ms2_GNPS_mtch_res$feature_id, ms2_GNPS_mtch_res$target_inchikey)) |>
    lapply(function(z) {
        z[which.max(z$score), ]
    }) |>
    do.call(what = rbind) |>
    as.data.frame()

#' List the best matching feature-compound pair
pandoc.table(ms2_GNPS_mtch_res[, c("feature_id", "target_compound_name", "score", 
                                        "target_inchikey")],
             style = "rmarkdown",
             caption = "Table 9.MS2 annotation results.",
             split.table = Inf) 
#' Export as df to manually analyse them
ms2_GNPS_result <- ms2_GNPS_mtch_res[, c("feature_id", "target_compound_name", "score", 
                                        "target_inchikey", "rtime")]
ms2_GNPS_result$rtmed <- res_sts_df[ms2_GNPS_result$feature_id, "rtmed"] 
# add rt highest spectra: rtime
write_xlsx(ms2_GNPS_result, "ms2_GNPS_results_rtime.xlsx")
```

added rtmed to check if the rtime from spectra is different. Calculate
difference NOTE: Filter consecutive features manually and same features matched
with different compounds.

We below plot the EIC for this feature highlighting the retention time at which
the highest scoring MS2 spectra was recorded and create a mirror plot comparing
this MS2 spectra to the reference fragment spectra.

```{r}
#' First, we need to add a line to do a matching between adjp < 0.05. If it is 
#' significant between two groups, it will plot plasma (ref) and those 2 groups 
#' only. 
#' Before, we need to import the columns with adjp:
ms2_GNPS_result <- cbind(ms2_GNPS_result, 
                             rowData(res_sts)[ms2_GNPS_result$feature_id, 
                                              c("coef.W", "coef.C","coef.M",
                                                "coef.B", "adjp.W", "adjp.C",
                                                "adjp.M", "adjp.B") ])
ms2_GNPS_result
```

#### EIC plot

```{r}
# x <- ms2_GNPS_result[1,]
# x
```

Next step is to subset each row by only plotting the devices that have an adjp
\< 0.05.

Each adjp means that the feature (row) is significant in M in comparison with
Plasma if the value is \< 0.05. Here is where I want to subset it somehow so
that I can filter the devices where each feature (row) is significant. For
example, if it is significant in P and M I would only plot those groups. If it
is sign in W and M in comparison with P then I will plot the 3 groups.

-   create new column: sign_in_device
-   get all adjp. columns
-   filter with \< 0.05
-   extract device names from column name: W, C, M, B
-   collapse into string (i.e. W, M, C)

```{r}
ms2_GNPS_result$sign_in_device <- apply(ms2_GNPS_result, 1, function(row) {
  # Get the names of all adjp.* columns
  adjp_cols <- grep("^adjp\\.", names(row), value = TRUE)
  
  # Filter those with value < 0.05
  sign_in_device <- adjp_cols[as.numeric(row[adjp_cols]) < 0.05]
  
  # Extract device names (remove 'adjp.' prefix)
  devices <- sub("^adjp\\.", "", sign_in_device)
  
  # Collapse into a string like "M,C"
  paste(devices, collapse = ",")
    # Map abbreviations to full names
  device_map <- c(W = "WHATMAN", M = "MITRA", C = "CAPITAINER", B = "BLOOD")
  full_names <- device_map[devices]
  paste(full_names, collapse = ",")
})
ms2_GNPS_result
```

Now, we should be able to plot EIC of significant features present only in the
devices where they are significant. For this:

-   subset/plot as if sign_in_device is only one device

-   match all colors

-   make the code for multiple devices: dev \<-
    strsplit(target\$sing_indev\[i\], ",")

-   sort colors the same order as lcms1_study_sign.

```{r}
#' Create the loop to plot and export EICs
#' Subset lcms1 to study samples
lcms1_study <- lcms1[!idx_blank_qc, keepFeatures = TRUE] # 50 samples. 

#' Get the devices names in column sign_in_device
dev <- strsplit(ms2_GNPS_result$sign_in_device[i], ",") |> unlist()

#' Subset sample groups where feature in x is present 
lcms1_study_sign <- lcms1_study[sampleData(lcms1_study)$device %in% c(dev, "PLASMA"), 
                                keepFeatures = TRUE] # 40 samples

#' Redefine colors
leg_devices <- leg_full[-c(1,2)]
tmp_leg <- leg_devices[unique(sampleData(lcms1_study_sign)$device)]
tmp_col <- tmp_leg[sampleData(lcms1_study_sign)$device]

#' Plot EIC
eic_sign <- featureChromatograms(lcms1_study_sign, features = ms2_GNPS_result$feature_id)

#' Create the loop to plot and export EICs
dr <- "EIC_sign/GNPS/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(ms2_GNPS_result))) {
    png(paste0(dr, "EIC_", ms2_GNPS_result$feature_id[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eic_sign[i, ], main = ms2_GNPS_result$feature_id[i],
         col = paste0(tmp_col, 80))
    grid()
    legend("topright", col = tmp_leg,
           legend = names(tmp_leg), lty = 1)
    dev.off()
}
```

-   FIX the colors.

-   NOTE: we should expect 279 EIC?

-   Change code. Plot in all samples, do only in specific groups per device by
    subseting each device.

-   Before plotting all groups. Now only plotting whatman and plasma.

-   We only got 128 EIC instead of 279 expected ?????

Mirror spectra: FIX: plot only mirror spectra with highest score.

```{r}
fts <- unique(ms2_GNPS_mtch$feature_id)

ppm <- 40
tolerance <- 0.05

dr <- file.path("MIRROR_SPECTRA/GNPS/", "matched_features")
dir.create(dr, showWarnings = FALSE, recursive = TRUE)

for (ft in fts) {
    ## Subset the data to all query MS2 spectra for that feature
    a <- ms2_GNPS_mtch[query(ms2_GNPS_mtch)$feature_id == ft]
    ## Identify all MS2 spectra that could be matched against reference
    for (i in whichQuery(a)) {
        ai <- a[i]
        ## For each query MS2 spectrum, identify all matching target spectra
        for (j in whichTarget(ai)) {
            ## Get the query and the target spectrum.
            q <- query(ai)
            t <- target(ai)[j]
            png(paste0(dr, "/", ft, "-q", i, "-t", j, ".png"),
                width = 10, height = 8, units = "cm", res = 600, pointsize = 6)
            plotSpectraMirror(q, t, ppm = ppm,
                              tolerance = tolerance, main = ft)
            grid()
            legend("topleft", legend = paste0("RT: ", format(rtime(q), digits = 4)))
            legend("bottomleft", legend = t$compound_name)
            dev.off()
        }
    }
}
```

we got \~3000 features with mirror spectra. How can I choose between q and t
repeated mirror spectra.

Next steps:

-   manually detect consecutive features annotated and repeated annotations
-   Filter those features
-   Change EIC code: split annotation sign per device and then plot it.
-   Create final list of annotation to start working with and reviewing MS2
    spectra
-   Heatmap per device (Ms1 data first)
