---
title: "Data analysis of Pilot Study from Imperial - Full data."
format: html
tbl-cap-location: bottom
editor: visual
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
author: Philippine Louail, Ana Sanchez Lorenzo
---

```{r setup, include=FALSE}
library(knitr)
library(quarto)
knitr::opts_knit$set(root.dir = './')
```

# Loading packages

```{r packages-used, message=FALSE, warning=FALSE}
#| code-fold: true
#| code-summary: "Show the code"
## General bioconductor package
library(BiocStyle)
library(BiocParallel)
library(knitr)

## Data Import and handling
library(readxl)
library(MsExperiment)
library(SummarizedExperiment)

## Preprocessing of LC-MS data
library(xcms)
library(Spectra)
library(MetaboCoreUtils)

## Statistical analysis
library(limma) # Differential abundance
library(matrixStats) # Summaries over matrices

## Visualisation
library(pander)
library(RColorBrewer)
library(pheatmap)
library(vioplot)
library(ggfortify)   # Plot PCA
library(gridExtra)   # To arrange multiple ggplots into single plots
library(grid)
library(UpSetR)

## Annotation
library(AnnotationHub) # Annotation resources
library(CompoundDb)    # Access small compound annotation data.
library(MetaboAnnotation) # Functionality for metabolite annotation.
```

# DATA ORGANISATION

## Data import

```{r import, message=FALSE}
# Data import
load("preprocessed_lcms1.RData")
load("res.RData")
featureDefinitions(lcms1) |>head()
```

```{r indexing, message=FALSE}
# Index per sample group
idx_qc <- sampleData(lcms1)$device == "QC"
idx_blank <- sampleData(lcms1)$device == "BLANK"
idx_blank_qc <- sampleData(lcms1)$device %in% c("QC","BLANK")
idx_samples <- sampleData(lcms1)$device %in% c("WHATMAN", "MITRA",
                                               "CAPITAINER", "PLASMA", "BLOOD")
idx_bl_sp <- sampleData(lcms1)$blank_type %in% c("BLANK_W", "BLANK_C", "BLANK_M")
```

```{r par-process}
#' Set up parallel processing using 3 cores
if (.Platform$OS.type == "unix") {
  register(MulticoreParam(4)) ### chunksize has to be be the same number. 
 } else {
     register(SnowParam(4))
}
```

## Define colors

We will define colors for each of the sample groups based on their sample group
using the *RColorBrewer* package:

```{r define-colors}
#| echo: true
#' Define colors for full dataset
leg_full <- brewer.pal(12, name = "Set3")[c(1, 12, 7, 5, 10, 4, 6)]
names(leg_full) <- unique(sampleData(lcms1)$device)
leg_full["QC"] <- "#f00e0e"
col_sample <- leg_full[sampleData(lcms1)$device]

#' define colors for devices only
leg_blank_BuS <- leg_full[-c(1, 2, 6, 7)]
col_blank_BuS <- leg_blank_BuS[sampleData(lcms1[!idx_bl_sp])$device]
leg_blank_BuS <- leg_blank_BuS[c("WHATMAN", "MITRA", "CAPITAINER")]
```

## Participants column addition For Differential abundance analysis

Linear model will include need to specify participants. For that we need to add
a new column from sample_name.

```{r participants}
#' Create a new column with participants names without device specification.
sampleData(lcms1)$participants <- substr(sampleData(lcms1)$sample_name, 1,
                                         nchar(sampleData(lcms1)$sample_name) -1)
sampleData(lcms1)$participants[idx_blank_qc] <- NA_character_
colData(res)$participants <- substr(colData(res)$sample_name, 1,
                                    nchar(colData(res)$sample_name) - 1)
colData(res)$participants[idx_blank_qc] <- NA_character_
```

# DATA NORMALISATION

## Initial quality assesment

Before normalization, PCA plot detect possible biases in data. Principal
Component Analysis is an unsupervised visualization that enables data
dimensionality reduction in order to visualise batch effect, injection index
variation and clustering of group samples.

Imputation of missing values is need by replacing by random numbers from a
uniform distribution from half the smallest signal measured for a specific
feature to its minimal value.

```{r Impute_MV}
#' Impute missing values using an uniform distribution
na_unidis <- function(z) {
    na <- is.na(z)
    if (any(na)) {
        min = min(z, na.rm = TRUE)
        z[na] <- runif(sum(na), min = min/2, max = min)
    }
    z
}

#' Row-wise impute missing values and add the data as a new assay
tmp <- apply(assay(res, "raw_filled"), MARGIN = 1, na_unidis)
assays(res)$raw_filled_imputed <- t(tmp)
```

### subset sumExp

In this line, blanks are removed from the we are subsetting the res object to do
not take into account blanks and/or QCs.

```{r PCA_indexing}
# PCA without blank
## Data exploration and comparison
### We remove blank and QC samples from the study samples

res_blank <- res[, res$device == "BLANK"] #only blank
res <- res[, !idx_blank]
 
#remove features that only have NA in the study samples
na_idx <- apply(assay(res, "raw_filled"), MARGIN = 1, function(x) all(is.na(x)))
res <- res[!na_idx, ]

## redefine color without blanks
leg_full <- brewer.pal(12, name = "Set3")[c(12, 7, 5, 10, 4, 6)]
names(leg_full) <- unique(res$device)
leg_full["QC"] <- "#f00e0e"
col_sample <- leg_full[res$device]
```

For better representation, data is log2 transformed, centered and scaled in 
order to stabilize the variance and remove dependence on absolute abundances,
respectively.

Plot all the PCA to observe the data structure and grouping of the different
samples.

```{r PCA_noblank, fig.height=5, fig.width=10}
#| echo: false
#| fig-cap: "Figure 1. PCA of all samples and QCs colored by sample group"
#' Log2 transform and scale data
vals <- assay(res, "raw_filled_imputed") |>
    log2() |>
    t() |>
    scale(center = TRUE, scale = TRUE)
 
#' Perform the PCA
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)
 
#' Plot the results
vals_st <- cbind(vals, device = res$device)
pca_12 <- autoplot(pca_res, data = vals_st , colour = 'device', scale = 0) +
    scale_color_manual(values = leg_full) +
    theme_minimal()
pca_34 <- autoplot(pca_res , data = vals_st, colour = 'device',
                   x = 3, y = 4, scale = 0) +
    scale_color_manual(values = leg_full) +
    theme_minimal()
grid.arrange(pca_12, pca_34, ncol = 2)
```

In the first PCA, study samples from each sample group are clustered separated
form QCs, with biggest representation of variance by PC1 (37.88%) and PC2
(12.89%). Plasma is more separation from blood and devices, and devices seem to
be close to each other. PC3 and PC4 do not contribute that much in the variance
between sample groups.

However, QC samples are spread across PC2, highlighting a possible technical
issue during the run. In the plot below, injection order correlation is analysed
in case injection order correction is needed.

```{r PCA_inj_order_samples_QCs, echo=TRUE, fig.height=5, fig.width=10}
#| fig-cap: "Figure 2. PCA of all samples and QCs colored by injection order"

#' check injection index samples and QCs
vals_st <- cbind(vals, injection_index = res$injection_index)

pca_injidx_12 <- autoplot(pca_res, data = vals_st, 
                                  colour = 'injection_index', scale = 0) + 
                                    theme_minimal()
pca_injidx_34 <- autoplot(pca_res, data = vals_st, colour = 'injection_index',
                   x = 3, y = 4, scale = 0) +   
    labs (title = "PCA coloured by injection order") +
                    theme_minimal()
grid.arrange(pca_injidx_12, pca_injidx_34, ncol = 2)

```

The PCA above shows a injection order correlation with the variance of QC
samples. To correct it and decrease variability due to technical bias, median
scaling will be performed.

### Intensity evaluation

```{r Intensity_evaluation_all}
#| fig-cap: "Figure 3. Number of detected peaks and feature abundances (SAMPLES AND QCs)."
layout(mat = matrix(1:3, ncol = 1), height = c(0.2, 0.2, 0.8))
par(mar = c(0.2, 4.5, 0.2, 3))

barplot(apply(assay(res, "raw"), MARGIN = 2, function(x) sum(!is.na(x))),
        col = paste0(col_sample, 80), border = col_sample,
        ylab = "# detected peaks", xaxt = "n", space = 0.012)
grid(nx = NA, ny = NULL)
barplot(apply(assay(res, "raw_filled"), MARGIN = 2, function(x) sum(!is.na(x))),
        col = paste0(col_sample, 80), border = col_sample,
        ylab = "# detected + filled peaks", xaxt = "n",
        space = 0.012)
grid(nx = NA, ny = NULL)
vioplot(log2(assay(res, "raw_filled")), xaxt = "n",
        ylab = expression(log[2]~feature~abundance),
        col = paste0(col_sample, 80), border = col_sample)
points(colMedians(log2(assay(res, "raw_filled")), na.rm = TRUE), type = "b",
       pch = 1)
grid(nx = NA, ny = NULL)
legend("topright", col = leg_full,
       legend = names(leg_full), lty=1, lwd = 2, xpd = TRUE, ncol = 3,
       cex = 0.8,  bty = "n")
```

The representation above of total intensities and log2 feature abundance per
sample show a decrease of number of features detected alog the retention time
due to instrumental sensibility decreased.The standard deviation is higher and
the number of features decreased in the end of the analysis. This change is
shown in all samples and is particularly obvious in QC samples. As samples were
randomized, the injection order correction will only affect the instrumental
variability and not biological variability. The comparison between raw data and
raw filled data showed that gap filling step rescue those features that have not
being detected in samples along the analysis.

### CV values

```{r CV_values}
#' Calculate the CV values
index_QC <- res$device == "QC"
cv_qc <- rowRsd(assay(res, "raw_filled")[, index_QC],
                    na.rm = TRUE, mad = TRUE)

#' Summarize the values across features
res_df <- quantile(cv_qc, na.rm = TRUE)

kable(res_df, format = "pipe")
```

A preliminary calculation of CV values show that 75% of features have CV\<30%
(23%), which indicates that less than 25% of features will be filtered after the
normalization.

### RLA

An RLA (Relative Log Abundance) value is the abundance of a feature in a sample
relative to the median abundance of the same feature across multiple samples.

-   Within group RLA plots: assess tightness of replicates within groups should
    have a median close to zero and low variation around it.
-   Across groups RLAs: they allow to compare behavior between group.

In this case, we apply calculate RLAs across groups to compare the RLA between
different groups and detect possible between-samples differences across the
analysis.

```{r RLA-log-abundance}
#| fig-cap: "Figure 4. RLAs plot for the filled data across the analysis"
par(mfrow = c(1, 1), mar = c(3.5, 4.5, 2.5, 1))
boxplot(MsCoreUtils::rowRla(assay(res, "raw_filled"),
                            f = res$device, transform = "log2"),
        cex = 0.5, pch = 16,
        col = paste0(col_sample, 80), ylab = "RLA",
        border = col_sample, boxwex = 1,
        outline = FALSE, xaxt = "n", main = "Relative log abundance",
        cex.main = 1)
axis(side = 1, at = seq_len(ncol(res)), 
    labels = colData(res)$sample_name)
grid(nx = NA, ny = NULL)
abline(h = 0, lty=3, lwd = 1, col = "black")
legend("topright", col = leg_full,
       legend = names(leg_full), lty=1, lwd = 2, xpd = TRUE, ncol = 3,
       cex = 0.8,  bty = "n")

```

On the RLA plot above, we can see that median is not centered around 0 and
decrease after QC07 (red samples). This can be associated to the sensibility of
the instrument and detector value optimisation across the analysis (is it worthy
to include "autogain" in method? Technical issues?). Also, mainly with Plasma
samples there is more variability, which is also shown in the PCA. This requires
samples need to be centered.

## Between sample normalisation

Between-sample normalization use filled-in features to effectively mitigate
variations influenced by technical issues, such as differences in sample
preparation, processing and injection methods. In this instance, we will employ
a commonly used technique known as median scaling.

### Median Scaling

This method involves computing the median for each sample, followed by
determining the median of these individual sample medians. This ensures
consistent median values for each sample throughout the entire data set.
Maintaining uniformity in the average total metabolite abundance across all
samples is crucial for effective implementation.

This process aims to establish a shared baseline for the central tendency of
metabolite abundance, mitigating the impact of sample-specific technical
variations. This approach fosters a more robust and comparable analysis of the
top features across the data set. The assumption is that normalizing based on
the median, known for its lower sensitivity to extreme values, enhances the
comparability of top features and ensures a consistent average abundance across
samples.

```{r normalization-factor-all}
#' Compute median and generate normalization factor
mdns <- apply(assay(res, "raw_filled"), MARGIN = 2,
              median, na.rm = TRUE )
nf_mdn <- mdns / median(mdns)

#' divide dataset by median of median and create a new assay.
assays(res)$norm <- sweep(assay(res, "raw_filled"), MARGIN = 2, nf_mdn, '/')
assays(res)$norm_imputed <- sweep(assay(res, "raw_filled_imputed"), MARGIN = 2,
                                  nf_mdn, '/')

```

The median scaling is calculated for imputed and non-imputed data, with each set
stored separately within the summarizedExperiment object. This approach
facilitates testing various nromalisation strategies while maintaining the
record of all processing steps undertaken, enabling easy regression to previous
stages if necessary.

```{r normalization-factor-all-2}
#' Compute median and generate normalization factor
mdns <- apply(assay(res_blank, "raw_filled"), MARGIN = 2,
              median, na.rm = TRUE )
nf_mdn <- mdns / median(mdns)

#' divide dataset by median of median and create a new assay.
assays(res_blank)$norm <- sweep(assay(res_blank, "raw_filled"), MARGIN = 2, nf_mdn, '/')
assays(res_blank)$norm_imputed <- sweep(assay(res_blank, "raw_filled_imputed"), MARGIN = 2,
                                  nf_mdn, '/')

```

Same step is repeating in blank samples set as all processing needs to be done
for Qualitative Analysis and Blank samples.

## Assessing overall effectiveness of the normalisation approach

### Principal component analysis

The evaluation will be performed by comparing the distribution of the log2
transformed feature abundances before and after normalization.

```{r PCA-normalisation, fig.height=7, fig.width=7}
#| echo: false
#| fig-cap: "Figure 5. PCA after normalization of all samples colored by groups. "

#' General PCA: QCs and devices
#' Log2 transform and scale data
vals <- assay(res, "raw_filled_imputed") |>
    log2() |>
    t() |>
    scale(center = TRUE, scale = TRUE)

#' Perform the PCA
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

#' Data before normalisation
vals_st <- cbind(vals, device = res$device)
pca_raw <- autoplot(pca_res, data = vals_st,
                    color = 'device', scale = 0) +
    scale_color_manual(values = leg_full) +
    labs (title = "PCA before Normalization", tag = "A") +
    theme_minimal()+
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
#' Data after normalization
vals_norm <- apply(assay(res, "norm"), MARGIN = 1, na_unidis) |>
    log2() |>
    scale(center = TRUE, scale = TRUE)

pca_res_norm <- prcomp(vals_norm, scale = FALSE, center = FALSE)
vals_st_norm <- cbind(vals_norm, device = res$device)
pca_adj <- autoplot(pca_res_norm, data = vals_st_norm,
                    color = 'device', scale = 0) +
    scale_color_manual(values = leg_full) +
    labs (title = "PCA after Normalization", tag = "B") +
    theme_minimal() + 
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
grid.arrange(pca_raw, pca_adj, ncol = 1)
```

Normalization have an impact in PC1 improving the grouping of QCs, Plasma and
Blood samples. Regarding QC samples, they cluste more tightly compared to the
prenormalisation PCA suggesting improved non-biological variation effect in the
experiment.

In terms of similarities between sample groups, microsampling devices
(Capitainer, Mitra and Whatman) cluster closely together and are clearly
separated from the whole blood and plasma samples along the PC1, which represent
44.26% of variance. This suggests that microsampling devices present similar
profiles, which differ from conventional matrices. Capitainer could be more
similar to blood and Mitra more similar to plasma but it is important to do a
quantitative analysis to determine if there are significant differences.

### Intensity Evaluation (RLA)

Relative log abundance plots (RLA) are useful to compare differences in
abundance between samples.

```{r RLA-after-norm2, fig.height=7, fig.width=7}
#| fig-cap: "Figure 6. RLA before and after normalisation across analysis."

layout(mat = matrix(1:2, ncol = 1), height = c(0.9,0.9))
par(mfrow = c(2, 1), mar = c(2, 4.5, 2.5, 1))

# A) RLA Raw data
boxplot(rowRla(assay(res, "raw_filled"), group = res$device),
        cex = 0.5, pch = 16,  border = leg_full, 
        ylab = "RLA", xlab = "injection order", 
        col = paste0(leg_full, 80), cex.main = 1, outline = FALSE,
        xaxt = "n", main = "Non-normalised data", boxwex = 1)
grid()
abline(h = 0, lty = 3, lwd = 1, col = "black")
legend("topright", col = leg_full,
       legend = names(leg_full), lty=1, lwd = 2, xpd = TRUE, ncol = 3,
       cex = 0.6,  bty = "n")

# B) RLA Normalised data
boxplot(rowRla(assay(res, "norm"), group = res$device),
        cex = 0.5, pch = 16,  border = leg_full, 
        ylab = "RLA", xlab = "injection order", 
        col = paste0(leg_full, 80), boxwex = 1, outline = FALSE,
        xaxt = "n", main = "Normalised data", cex.main = 1)
axis(side = 1, at = seq_len(ncol(res)), 
     labels = colData(res)$sample_name)
grid(nx = NA, ny = NULL)
abline(h = 0, lty = 3, lwd = 1, col = "black")
legend("topright", col = leg_full,
       legend = names(leg_full), lty=1, lwd = 2, xpd = TRUE, ncol = 3,
       cex = 0.6,  bty = "n")

```

After normalization, relative log abundance plot shows that all QCs abundance
have been centered around the median. This make the medians of each sample close
to zero. Also, all boxes are centered around 0 suggesting all samples being
comparable.

RLA plot is performed again only focussing on QC samples for better
visualisation.

```{r RLA-after-norm1, fig.height=7, fig.width=7}
#| fig-cap: "Figure 7. RLA of QCs before and after normalisation."

colors <- ifelse(res$device == "QC", "#f00e0e", "grey")

layout(mat = matrix(1:2, ncol = 1), height = c(0.8,0.8))
par(mfrow = c(2, 1), mar = c(4, 4.5, 3.5, 1))

# A RLA Raw data
boxplot(rowRla(assay(res, "raw_filled"), group = res$device),
        cex = 0.5, pch = 16, ylab = "RLA", border = colors,
        col = paste0(colors, 80), cex.main = 1, outline = FALSE,
        xaxt = "n", main = "Non-normalised data", boxwex = 1)
grid()
abline(h = 0, lty = 3, lwd = 1, col = "black")
legend("topright", col = colors,
       legend = names(leg_full), lty=1, lwd = 2, xpd = TRUE, ncol = 3,
       cex = 0.6,  bty = "n")
mtext("Injection Order", side = 1, line = 2.5, cex = 0.9)
# B RLA Normalised data
boxplot(rowRla(assay(res, "norm"), group = res$device),
        cex = 0.5, pch = 16, ylab = "RLA", border = colors,
        col = paste0(colors, 80), boxwex = 1, outline = FALSE,
        xaxt = "n", main = "Normalised data", cex.main = 1)
mtext("Injection Order", side = 1, line = 2.5, cex = 0.9)
grid()
abline(h = 0, lty = 3, lwd = 1, col = "black")
legend("topright", col = colors,
       legend = names(leg_full), lty=1, lwd = 2, xpd = TRUE, ncol = 3,
       cex = 0.6,  bty = "n")
```

By only plotting the QCs we can visualise the effect of normalisation based on
QC samples are centered around 0. 


### CV after normalisation

```{r CV-after-norm}
#' Calculate the CV values
index_samples <- res$device %in% c("WHATMAN", "MITRA", "CAPITAINER",
                                   "PLASMA", "BLOOD")
index_QC <- res$device == "QC"

sample_res <- cbind(
    QC_raw = rowRsd(assay(res, "raw_filled")[, index_QC],
                    na.rm = TRUE, mad = TRUE),
    QC_norm = rowRsd(assay(res, "norm")[, index_QC],
                     na.rm = TRUE, mad = TRUE),
    Study_raw = rowRsd(assay(res, "raw_filled")[, index_samples],
                       na.rm = TRUE, mad = TRUE),
    Study_norm = rowRsd(assay(res, "norm")[, index_samples],
                        na.rm = TRUE, mad = TRUE)
)

#' Summarize the values across features
res_df <- data.frame(
    QC_raw = quantile(sample_res[, "QC_raw"], na.rm = TRUE),
    QC_norm = quantile(sample_res[, "QC_norm"], na.rm = TRUE),
    Study_raw = quantile(sample_res[, "Study_raw"], na.rm = TRUE),
    Study_norm = quantile(sample_res[, "Study_norm"], na.rm = TRUE)
)

kable(res_df, format = "pipe")
```

The coefficient of variation is clearly reduced by the normalization for the QC
samples suggesting that indeed technical variance was removed. For study
samples, CV is comparable suggesting that normalization indeed just reduced the
technical, but not the biological variance in the data set.


#### Density Plots with the distribution of CV in the whole dataset.

```{r densityplot-all}
#| fig-cap: "Figure 8. Density plot for Normalisation evaluation with CV values distribution"

## Density Plot to evaluation normalization evaluation
dens <- apply(sample_res, 2, density)
# Calculate the density estimates in the CV values and subset it only for the QCs

# Create density plots
plot_density_full <- plot(NA, xlim=range(sapply(dens, "[", "x")), ylim=range(sapply(dens, "[", "y")),
     xlab = "CV Values", ylab = "Density", main = "Density Plot for Normalisation Evaluation")

    # Define a color palette with transparency
    cols <- rainbow(length(dens))  # You can use any color palette you prefer
    alpha <- 0.6  # Set the transparency level (0 is fully transparent, 1 is fully opaque)
    transparent_cols <- sapply(cols, function(col) rgb(t(col2rgb(col)) / 255, alpha = alpha))

    # Plot each density with a different color and transparency
    mapply(function(density, col) {
    polygon(density, col = col, border = NA)
    }, dens, col = transparent_cols)
    
    # Add a legend to the plot
    legend("topright", legend = names(dens), fill = transparent_cols, title = "Variables")
```

In the density plot above, normalisation effect on the coefficient of variation
(CV) values is evaluated. After applying median scaling, the variance among QC
samples is reduced (from 23% to 16%), indicating effective correction of
technical variability. In contrast, the distribution of CVs in the study samples
remains, suggesting biological variability has been preserved. However, the
variance in the biological samples stay without changes. This demonstrates that
the normalization method successfully minimizes technical noise without
compromising the biological signal.

```{r densityplot-qc}
#| fig-cap: "Figure 9. Density plot for Normalisation evaluation with CV values distribuiton from QC samples."

# Subset dens to include only QC_raw and QC_norm

dens_subset <- dens[c("QC_raw", "QC_norm")]

# Plot setup
plot(NA, xlim=range(sapply(dens_subset, "[", "x")), ylim=range(sapply(dens_subset, "[", "y")),
     xlab = "Value", ylab = "Density", main = "Density Plot for Normalisation Evaluation", xaxt = "n")

# Define specific colors with transparency for the subset
colors <- c("QC_raw" = "red", "QC_norm" = "green")
alpha <- 0.6  # Set the transparency level (0 is fully transparent, 1 is fully opaque)
transparent_cols_subset <- sapply(colors, function(col) rgb(t(col2rgb(col)) / 255, alpha = alpha))

# Plot each density in the subset with the specified color and transparency
mapply(function(density, col) {
  polygon(density, col = col, border = NA)
}, dens_subset, col = transparent_cols_subset)

# Add a vertical line at x = 0.3, red and dashed
abline(v = 0.3, col = "red", lty = 2)

# Add custom x-axis with values from 0 to 1.5 every 0.1
axis(1, at = seq(0, 1.5, by = 0.1), labels = seq(0, 1.5, by = 0.1))

# Add a legend to the plot for the subset
legend("topright", legend = names(dens_subset), fill = transparent_cols_subset)

```

This density plot shows the distribution of CV values for QC samples before
(red) and after (green) normalization. The dashed vertical line at 30%
represents a commonly used threshold in metabolomics for filtering out features
with high technical variability. The plot allows visual assessment of how many
features exceed this cutoff, and how normalization reduces the number of
features above this threshold, thereby improving data quality while retaining
more reliable variables.

# FILTERS:

## BlankFlag features filtering

After the normalization, I removed all possible contaminants from the blank to
avoid including features in the qualitative analysis. In this case, the
'possible_contaminant' column has NA values which need to be transformed in
FALSE based.

```{r blankflag-filter}
# Keep an original object
res_orig <- res
nrow(res_orig)
# fix for NA presence in column.
rowData(res)$possible_contaminants[is.na(rowData(res)$possible_contaminants)] <- FALSE

# subsetting res removing possible contaminants from blank. 
## NOTE: FROM NOW ON NROWs is has less features. Do not mix. 
res <- res[!rowData(res)$possible_contaminants, ]
nrow(res)
#Percentage of features removed
nrow(res)/nrow(res_orig) * 100
```

The original data set consisted of `r nrow(res_orig)` features. Removal of
possible contaminants reduced the data set to `r nrow(res)` features.

Number of possible_contaminants removed: 338 Number features remaining 2564
Percentage remain features : 88%

## Filter RSD filter:

RSD, defined as the st. dev of QC samples divided by st deviation for biological
samples, for each feature.

```{r unfiltered-object}
# Keep unfiltered object
res_unfilt <- res
```

```{r RSD-filter}
#' Compute and filter based on the RsdFilter
filter_Rsd <- RsdFilter(threshold = 0.3,
                        qcIndex = res$device == "QC",
                        mad = FALSE) # any particular reason why mad = FALSE was used here and not above?
res <- filterFeatures(res, filter = filter_Rsd, assay = "norm_imputed")

```

After RSD (CV) filtering, the data set contains `r nrow(res)` features.

NOTE: check filtering. before 198 features were removed. Why now 200 features
are removed?

```{r features-number}
#' Number of features after analysis
nrow(res)
#' Percentage left: end/beginning
nrow(res)/nrow(res_orig) * 100
```

As we need to assess number of features in blank we need to do all process of
normalisation and filtering with res_blank object.

```{r features-number-2}
# Keep an original object
res_blank_orig <- res_blank
nrow(res_blank_orig)
# fix for NA presence in column.
rowData(res_blank)$possible_contaminants[is.na(rowData(res_blank)$possible_contaminants)] <- FALSE
# subsetting res removing possible contaminants from blank. 

## FROM NOW ON NROWs is different. 
res_blank <- res_blank[!rowData(res_blank)$possible_contaminants,]
nrow(res_blank)
#Percentage of features removed
nrow(res_blank)/nrow(res_blank_orig) * 100

# Keep unfiltered object
res_blank_unfilt <- res_blank
```

*RsdFilter*: enables features filtering based on relative standard deviation
(RSD or CV) for a specified threshold. In this case, features with CV \> 30%
will be removed, as it is te upper limit accepted for biomarker discovery in
conventional bioanalysis (doi:10.1038/nprot.2010.50).

After RSD filtering, 198 features were removed, which represents the 92.2% of
total number of features. Only `r nrow(assay(res))` features will remain. After
the normalisation we want to remove QCs samples to evaluate the quality data.
Also, the data is used to see the clustering depending on cofactors as age and
sex.

```{r removing-QCs}
res_QC <- res[, res$device == "QC"] #only QCs
res <- res[, !res$device == "QC"]

# redefine colors without QCs
leg_full <- brewer.pal(12, name = "Set3")[c(7, 5, 10, 4, 6)]
names(leg_full) <- unique(res$device)
col_sample <- leg_full[res$device]
```

After filter, Age and Sex PCA are plotting to asses if they could be cofactors
for linear model.

```{r PCA-age}
#| fig-cap: "Figure 10. PCA after normalization of all samples colored by age"
#' Add age to the PCA plot depending on age

# subset res object to remove qcs to assess age. 
#' Log2 transform and scale data
vals <- assay(res, "raw_filled_imputed") |>
    log2() |>
    t() |>
    scale(center = TRUE, scale = TRUE)
#' Perform the PCA
pca_res <- prcomp(vals, scale = FALSE, center = FALSE)

vals_st_age <- cbind(vals, age = res$age)
pca_age_12 <- autoplot(pca_res, data = vals_st_age , 
                              colour = 'age', scale = 0)  + labs (title = "PCA coloured by age")
    theme_minimal()
pca_age_12

```

```{r PCA_age, echo=TRUE, fig.height=5, fig.width=10}
#' Add age to the PCA plot depending on sex
vals_st_sex <- cbind(vals, sex = res$sex)
pca_sample_sex_12 <- autoplot(pca_res, data = vals_st_sex , colour = 'sex', scale = 0) + labs (title = "PCA coloured by sex") +
    theme_minimal()
grid.arrange(pca_sample_sex_12)
```

PCAs do not show any clustering related with age or sex.

# QUALITATIVE ANALYSIS

In order to do a qualitative comparison between the devices and reference
samples, number of features, percentage of missing values, and distribution of
intensities will be analysed in bar plots. Unique features will be assessed in
an upset plot.

## Device specific blanks

Apart from possible contaminants from double blanks, it would be interesting to
assess which device is more prone to contamination. That would be translate in
which device has more chrompeak detected, as it is expected to have the less as
possible. In this case, blank from whatman, mitra and capitainer were subsetted
from he raw sumExp object (res). All intensities values will be transformed into
a binary matrix where 1 is feature present, 0 is feature not present. Using the
colSum() function all features found in each device are sum per sample and then
ploted in a barplot.

(We realized that with raw-filled assay it doesn't bring new information, as
after the gap-filling almost all features will have intensity values and
therefor will have been detected).

Same data is ploted with raw assay to avoid filling process.

```{r test-function-int}
#x <- res[, 1]
small_fct <-function (x){
    as.integer(!is.na(x))
}
#small_fct(x)
```

```{r features-blanks}
idx_bl_sp_1<-which(idx_bl_sp)
b_fts <- apply(assay(res_blank[,idx_bl_sp_1], "raw"), 2, small_fct)
colnames(b_fts) <- c("Whatman Blank", "Mitra Blank", "Capitainer Blank")

#Bar Plot to identify NA per device
barplot(colSums(b_fts), col = leg_blank_BuS, ylab = "Number of features", 
        space = 0.25, ylim = c(0, 600), main = "Number of features per Blank Type (raw data)")
```

Blank from Mitra seems to have less features (226) in the blank than Whatman
(178) and Capitainer (301), being Capitainer blank the one with highest number
of features, which could be related with some possible contaminants due to the
device. NOTE: we do not remove blank_sp features because there are no
triplicates.

### Upset Plot Blanks

```{r upset-plot-blank, eval=FALSE}
#' Upset plot from device specific blanks
upset(as.data.frame(b_fts), sets.bar.color = c("#80B1D3","#B3DE69", "#BC80BD"))
```

If we look specifically in a Upset plot, it can be identified the unique devices
features and common features to all devices. In this case, there is a big
overlap between blank devices. However, we found unique features present only in
specific devices (86 unique features in Capitainer, 63 in whatman and 6 in
mitra).

However, we cannot identify them as there are no MSMS data from blanks. Only we
could compare if they are statistical signigicant during the statistical
analysis.

## Number of features and missing values

To assess the number of features and missing values per device and reference
samples (blood and plasma) it is required to create a binary matrix from the
sumExp object where 0 is non detected and 1 is detected. As this time, we have
10 samples per group, there will be a criteria: if it is detected in less than 5
samples per group it will be consider as non detected (0).

Why do we use unfilled data for qualitative analysis?

During the gap filling, the algorithm is rescuing intensities of NA features in
some samples from the unfilled data. The reason why we can rescue that data is
because those features are already present in at least one sample.\
However, if we want to do an experiment only with one type of device or
reference samples, the algorithm won't know those missing features. So, it is
important to keep raw data to measure the number of features. However, for
quantitative analysis, as intensity will be taken into account and imputation
will need to be done, it would be better to do gap filling beforehand.

```{r upset-plot}
# Indexing per device to sum per rows
idx_W <- res$device == "WHATMAN"
idx_C <- colData(res)$device == "CAPITAINER"
idx_M <- colData(res)$device == "MITRA"
idx_B <- colData(res)$device == "BLOOD"
idx_P <- colData(res)$device == "PLASMA"

# Create a data frames per groups for the BAR plot with binary data
fts_W <- apply(assay(res[,idx_W], "raw"), 2, small_fct)
fts_C <- apply(assay(res[,idx_C], "raw"), 2, small_fct)
fts_M <- apply(assay(res[,idx_M], "raw"), 2, small_fct)
fts_B <- apply(assay(res[,idx_B], "raw"), 2, small_fct)
fts_P <- apply(assay(res[,idx_P], "raw"), 2, small_fct)
ncol(res)
# Create a binary matrix with condition: <5 = 0
binary_all <- cbind(
    ifelse(rowSums(fts_W) < 5, 0, 1),
    ifelse(rowSums(fts_C) < 5, 0, 1),
    ifelse(rowSums(fts_M) < 5, 0, 1),
    ifelse(rowSums(fts_B) < 5, 0, 1),
    ifelse(rowSums(fts_P) < 5, 0, 1)
    )

colnames(binary_all) <- c("WHATMAN", "CAPITAINER", "MITRA", "BLOOD", "PLASMA")


# #Bar Plot to identify NA per device
# barplot(colSums(binary_all), col = leg_full, ylab = "Number of features", 
#         space = 0.25, ylim = c(0, 2500), main = "Number of features per sample group", 
#         xlab = colSums(binary_all))

# save positions of each bar
bar_positions <- barplot(colSums(binary_all), 
                         col = leg_full, 
                         ylab = "Number of features", 
                         space = 0.25, 
                         ylim = c(0, 2500), 
                         main = "Number of features per sample group")

# Add number of features on top of each bar
text(x = bar_positions,
     y = colSums(binary_all) + 100,  # +100 to center text
     labels = colSums(binary_all),
     cex = 0.9, font = 2)


```

Once error solved, replace binary_all by logical_all and delete logical all from
below. As we were expecting plasma has less number of features than the rest and
all devices have a similar number of features, being Whatman the device with
higher number of features:

```{r}
colSums(binary_all)
```

### Upset Plot samples

In our analysis, we aim to study deeper the distribution and overlap of features
across various sample groups. Specifically, we are interested in identifying the
number of features that are common to all samples, those that are shared
exclusively among specific devices, and the features that are unique to each
sample group. To achieve this, an upset plot provide a clear visualisation of
all sets of intersections:

```{r up-set-all}
#' Upset plot from device specific blanks
upset(as.data.frame(binary_all), sets.bar.color = c("#B3DE69", "#BC80BD", "#80B1D3","#FB8072",   "#FDB462"))
```

Set size: plasma has the smallest set size, while Whatman has the largest. Mitra
and capitainer seems to have a similar set size but as we created a binary
matrix.

The upset plot shows there is 990 features common to all sample groups. However,
there are 235 common to blood and devices and 291 common to devices. Plasma is
the group with more unique features (167). It would be interesting to identify
number of unique features per sample group and features common to all devices.

```{r NMissing-values}
#| fig-height: 8
#| fig-width: 5
# log 2 intensities
intensity <- cbind(
    Whatman = log2(as.numeric(assay(res[,idx_W], "raw_filled"))),
    Capitainer = log2(as.numeric(assay(res[,idx_C], "raw_filled"))),
    Mitra = log2(as.numeric(assay(res[,idx_M], "raw_filled"))),
    Blood = log2(as.numeric(assay(res[,idx_B], "raw_filled"))),
    Plasma = log2(as.numeric(assay(res[,idx_P], "raw_filled")))
)
#percentage of missing values with unfilled data. 
percent_missing <- cbind(
    Whatman = sum(is.na(assay(res[,idx_W], "raw")))/length(assay(res[,idx_W], "raw")) * 100, 
    Capitainer = sum(is.na(assay(res[,idx_C], "raw")))/length(assay(res[,idx_C], "raw")) * 100,
    Mitra = sum(is.na(assay(res[,idx_M], "raw")))/length(assay(res[,idx_M], "raw")) * 100,
    Blood = sum(is.na(assay(res[,idx_B], "raw")))/length(assay(res[,idx_B], "raw")) * 100,
    Plasma = sum(is.na(assay(res[,idx_P], "raw")))/length(assay(res[,idx_P], "raw")) * 100
)
# Plot
layout(mat = matrix(1:3, ncol = 1), height = c(0.3, 0.3, 0.8))
par(mar = c(1, 4.5, 1, 3))
barplot(colSums(binary_all), col = leg_full, ylab = "Number of features", 
        space = 0.05)
barplot(c(percent_missing), ylab = "% of missing values", 
        col = leg_full, space = 0.05, ylim = c(0, 60))
vioplot(intensity, ylab = "Log2 intensity", col = leg_full,
        space = 0.10, ylim = c(4, 25))
```

If I used the unfilled and filtered dataset I see the distribution of low
intensity features is cut. (question: Should I keep the original assay
(assay(res)\$raw) without gapfilling? median scaling? noise threshold?)

## Distribution across rt

```{r}
#' Bin features per RT slices
vc <- rowData(res)$rtmed
breaks <- seq(60, max(vc, na.rm = TRUE) + 1, length.out = 15) |> 
    round(0) 
cuts <- cut(vc, breaks = breaks, include.lowest = TRUE)
table(cuts)

num_ft <- apply(binary_all, MARGIN = 2, 
                function(x) unlist(lapply(split(x, cuts), sum)))

#' Create the dataframe with TRUE/FALSE values per sample group. 
#conditioning: if rowsum is more or igual than 5 it is TRUE, otherwise FALSE.
logical_all <- data.frame(
    WHATMAN = rowSums(fts_W) >= 5,
    CAPITAINER = rowSums(fts_C) >= 5,
    MITRA = rowSums(fts_M) >= 5,
    BLOOD = rowSums(fts_B) >= 5,
    PLASMA = rowSums(fts_P) >= 5
    ) 

#' Subset Res object per sample group
res_w <- res[logical_all$WHATMAN, res$device == "WHATMAN"]
res_m <- res[logical_all$MITRA, res$device == "MITRA"]
res_c <- res[logical_all$CAPITAINER, res$device == "CAPITAINER"]
res_b <- res[logical_all$BLOOD, res$device == "BLOOD"]
res_p <- res[logical_all$PLASMA, res$device == "PLASMA"]

#' Create funtion to sum intensities per groups and then per rt slices. 
ftc <-function(res_devices, fts_idx) {
    tmp <- rowMedians(assay(res_devices, "raw"), na.rm = TRUE) # get median of intensity per feature
    cuts_tmp <- cuts[fts_idx] #split into the rt ranges/slices
    t <- split(tmp, cuts_tmp) |>
        lapply(sum, na.rm = TRUE, simplify = TRUE) # summing per slices
    unlist(t)
}

ftc(res_devices = res_w, fts_idx = logical_all$WHATMAN)
ftc(res_devices = res_m, fts_idx = logical_all$MITRA)
ftc(res_devices = res_c, fts_idx = logical_all$CAPITAINER)
ftc(res_devices = res_b, fts_idx = logical_all$BLOOD)
ftc(res_devices = res_p, fts_idx = logical_all$PLASMA)

int <- list(
    WHATMAN = log2(ftc(res_w, logical_all$WHATMAN)),
    MITRA = log2(ftc(res_m, logical_all$MITRA)),
    CAPITAINER = log2(ftc(res_c, logical_all$CAPITAINER)),
    BLOOD = log2(ftc(res_b, logical_all$BLOOD)),
    PLASMA = log2(ftc(res_p, logical_all$PLASMA))
    )
```

```{r}
# Plot
layout(mat = matrix(1:2, ncol = 1), heights = c(0.5, 0.5))
par(mar = c(0.5, 4.5, 2, 3))

## Plot number of features
ylim_features <- c(0, max(unlist(num_ft)))
plot(x = seq_len(nrow(num_ft)), y = num_ft[,1], xaxt = "n", col = leg_full[1], ylab = "Number of features",
     xlab = "", type = "b", pch = 16,  ylim = ylim_features,
     main = "Analysis along the RT axis for each devices")
for (i in 2:ncol(num_ft)) {
   points(x = seq_len(nrow(num_ft)), y = num_ft[,i], col = leg_full[i], type = "b", pch = 16)
}
axis(1, at = 1:length(num_ft[,1]), labels = FALSE)
grid(lwd = 0.5)
legend("top", legend = names(logical_all), col = leg_full, 
       pch = 16, cex = 0.8, horiz = TRUE, bty = "n")
grid()
# Plot intensity
par(mar = c(4.5, 4.5, 2, 3))
ylim_intensity <- range(unlist(int))
plot(int[[1]], type = "b", pch = 16, ylab = "Log2 intensity",
     col = leg_full[[1]], xaxt = "n", xlab = "Retention time bins [s]",
     ylim = ylim_intensity)
for (i in 2:length(int)) {
  lines(int[[i]], type = "b", pch = 16, col = leg_full[i])
}
axis(side = 1, at = seq_len(nrow(num_ft)), labels = rownames(num_ft))
grid()
```

The number of features across the rt is similar between the devices. However,
blood and plasma present less number of features in a between 3-6 min and 8-10
(aproximately). These are areas of chromatogram that do not match with specific
biological lipids groups, so we will expect differences in contaminants
features. In general, in the regions where we found biological features plasma
and blood present less intensity than the devices.

## Distribution across m/z

```{r}
#' Bin features per m/z slices
vc <- rowData(res)$mzmed
breaks <- seq(60, max(vc, na.rm = TRUE) + 1, length.out = 15) |> 
    round(0) 
cuts <- cut(vc, breaks = breaks, include.lowest = TRUE)
table(cuts)

num_ft <- apply(binary_all, MARGIN = 2, 
                function(x) unlist(lapply(split(x, cuts), sum)))

#' Create the dataframe with TRUE/FALSE values per sample group. 
#conditioning: if rowsum is more or igual than 5 it is TRUE, otherwise FALSE.
logical_all <- data.frame(
    WHATMAN = rowSums(fts_W) >= 5,
    CAPITAINER = rowSums(fts_C) >= 5,
    MITRA = rowSums(fts_M) >= 5,
    BLOOD = rowSums(fts_B) >= 5,
    PLASMA = rowSums(fts_P) >= 5
    ) 

#' Subset Res object per sample group
res_w <- res[logical_all$WHATMAN, res$device == "WHATMAN"]
res_m <- res[logical_all$MITRA, res$device == "MITRA"]
res_c <- res[logical_all$CAPITAINER, res$device == "CAPITAINER"]
res_b <- res[logical_all$BLOOD, res$device == "BLOOD"]
res_p <- res[logical_all$PLASMA, res$device == "PLASMA"]

#' Create funtion to sum intensities per groups and then per rt slices. 
ftc <-function(res_devices, fts_idx) {
    tmp <- rowMedians(assay(res_devices, "raw"), na.rm = TRUE) # get median of intensity per feature
    cuts_tmp <- cuts[fts_idx] #split into the rt ranges/slices
    t <- split(tmp, cuts_tmp) |>
        lapply(sum, na.rm = TRUE, simplify = TRUE) # summing per slices
    unlist(t)
}

ftc(res_devices = res_w, fts_idx = logical_all$WHATMAN)
ftc(res_devices = res_m, fts_idx = logical_all$MITRA)
ftc(res_devices = res_c, fts_idx = logical_all$CAPITAINER)
ftc(res_devices = res_b, fts_idx = logical_all$BLOOD)
ftc(res_devices = res_p, fts_idx = logical_all$PLASMA)

int <- list(
    WHATMAN = log2(ftc(res_w, logical_all$WHATMAN)),
    MITRA = log2(ftc(res_m, logical_all$MITRA)),
    CAPITAINER = log2(ftc(res_c, logical_all$CAPITAINER)),
    BLOOD = log2(ftc(res_b, logical_all$BLOOD)),
    PLASMA = log2(ftc(res_p, logical_all$PLASMA))
    )
```

```{r}
# Plot
layout(mat = matrix(1:2, ncol = 1), heights = c(0.5, 0.5))
par(mar = c(0.5, 4.5, 2, 3))

## Plot number of features
ylim_features <- c(0, max(unlist(num_ft)))
plot(x = seq_len(nrow(num_ft)), y = num_ft[,1], xaxt = "n", col = leg_full[1], ylab = "Number of features",
     xlab = "", type = "b", pch = 16,  ylim = ylim_features,
     main = "Analysis along the m/z axis for each devices")
for (i in 2:ncol(num_ft)) {
   points(x = seq_len(nrow(num_ft)), y = num_ft[,i], col = leg_full[i], type = "b", pch = 16)
}
axis(1, at = 1:length(num_ft[,1]), labels = FALSE)
grid(lwd = 0.5)
legend("top", legend = names(logical_all), col = leg_full, 
       pch = 16, cex = 0.8, horiz = TRUE, bty = "n")
grid()
# Plot intensity
par(mar = c(4.5, 4.5, 2, 3))
ylim_intensity <- range(unlist(int))
plot(int[[1]], type = "b", pch = 16, ylab = "Log2 intensity",
     col = leg_full[[1]], xaxt = "n", xlab = "m/z bins [s]",
     ylim = ylim_intensity)
for (i in 2:length(int)) {
  lines(int[[i]], type = "b", pch = 16, col = leg_full[i])
}
axis(side = 1, at = seq_len(nrow(num_ft)), labels = rownames(num_ft))
grid()
```



# DIFFERENCIAL ABUNDANCE ANALYSIS (lm)

The next step is to identify features that are differential abundant between the
devices and reference samples. This allows to identify potential metabolites
that are extracted with significant differences within different devices. In
this case, a multiple linear regression analysis is used to identify the
features with significantly difference in abundances between Whatman, Mitra,
capitainer, Blood or plasma. First, it is important to evaluate the similarities
between study samples using PCA, after excluding QC samples to avoid them to
bring variance.

We want to see variantion within the devices so first we will see how they
cluster.

```{r PCA-only-samples-before-lm}
#| echo: false
#| fig-cap: "PCA after normalization of all samples colored by device"

#' Data without blanks and QCs after normalization
vals <- apply(assay(res, "norm_imputed"), MARGIN = 1, na_unidis) |>
    log2() |>
    scale(center = TRUE, scale = TRUE)
pca_res_norm <- prcomp(vals, scale = FALSE, 
                       center = FALSE)

vals_st <- cbind(vals, device = res$device)
pca_sample_adj <- autoplot(pca_res_norm, data = vals_st,
                    color = 'device', scale = 0) +
    scale_color_manual(values = leg_full) +
    labs (title = "PCA after Normalization") +
    theme_minimal()

grid.arrange(pca_sample_adj, ncol = 1)
```

For our analysis we use a linear model y ~ device + sex + participant, hence
explaining the abundances of one metabolite accounting for the study group
assignment and sex of each individual.

## Plasma vs devices

### Linear model

```{r lm plasma}
#' Define the linear model to be applied to the data
p.cut <- 0.05     # cut-off for significance.
m.cut <- 0.5      # cut-off for log2 fold change

# The algorithm takes the first group as reference, so levels must be changed: 
# levels(res$device) <- c("PLASMA", "WHATMAN", "MITRA", "CAPITAINER", "BLOOD")

#' Define the linear model to be applied to the data
device <- factor(x = res$device, levels =  c("PLASMA", "WHATMAN", "MITRA",
                                             "CAPITAINER", "BLOOD"))
sex <- factor(res$sex)
# design <- model.matrix(~ device + sex + participants)
# As there are NA values in one of the participants we need to continue without participants. 

design <- model.matrix(~ device + sex)

#' Fit the linear model to the data, explaining metabolite
#' concentrations by device and sex.
fit <- lmFit(log2(assay(res, "norm_imputed")), design = design) # fit model to log2-transformed abundances
fit <- eBayes(fit) # calculate p-value
fit
```

Extract the results:

```{r result-df-plasma}
#' Compile a result data frame, do we do one data.frame per device ? or for the full thing ? 
tmp <- data.frame(
    coef.W = fit$coefficients[, "deviceWHATMAN"],
    pvalue.W = fit$p.value[, "deviceWHATMAN"],
    adjp.W = p.adjust(fit$p.value[, "deviceWHATMAN"], method = "BH"), # Benjamini-Hochberg correction
    avg.W = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "WHATMAN"])),
    avg.P = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "PLASMA"])), 
    coef.M = fit$coefficients[, "deviceMITRA"],
    pvalue.M = fit$p.value[, "deviceMITRA"],
    adjp.M = p.adjust(fit$p.value[, "deviceMITRA"], method = "BH"), # Benjamini-Hochberg correction
    avg.M = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "MITRA"])),
    coef.C = fit$coefficients[, "deviceCAPITAINER"],
    pvalue.C = fit$p.value[, "deviceCAPITAINER"],
    adjp.C = p.adjust(fit$p.value[, "deviceCAPITAINER"], method = "BH"), # Benjamini-Hochberg correction 
    avg.C = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "CAPITAINER"])),
    coef.B = fit$coefficients[, "deviceBLOOD"],
    pvalue.B = fit$p.value[, "deviceBLOOD"],
    adjp.B = p.adjust(fit$p.value[, "deviceBLOOD"], method = "BH"), # Benjamini-Hochberg correction
    avg.B = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "BLOOD"])),
    coef.S = fit$coefficients[, "sexW"],
    pvalue.S = fit$p.value[, "sexW"],
    adjp.S = p.adjust(fit$p.value[, "sexW"], method = "BH"), # Benjamini-Hochberg correction
    avg.S = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "SEX"]))
)

tmp$significant.W <- tmp$adjp.W < 0.05
tmp$significant.M <- tmp$adjp.M < 0.05
tmp$significant.C <- tmp$adjp.C < 0.05
tmp$significant.B <- tmp$adjp.B < 0.05
tmp$significant.S <- tmp$adjp.S < 0.05
#' Add differential abundance results to the object's rowData
rowData(res) <- cbind(rowData(res), tmp)
```

Now we visualise the distribution of the raw and adjusted p-values:

### Distribution pvalues

```{r distribution-pvalues-plasma}
#' Plot the distribution of p-values in Whatman
par(mfrow = c(1, 2))
hist(rowData(res)$pvalue.W, breaks = 64, xlab = "p value",
     main = "Distribution of raw p-values - Whatman",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
hist(rowData(res)$adjp.W, breaks = 64, xlab = expression(p[BH]~value),
     main = "Distribution of adjusted p-values - Whatman",
     cex.main = 1, cex.lab = 1, cex.axis = 1)

#' Plot the distribution of p-values in Mitra
par(mfrow = c(1, 2))
hist(rowData(res)$pvalue.M, breaks = 64, xlab = "p value",
     main = "Distribution of raw p-values - Mitra",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
hist(rowData(res)$adjp.M, breaks = 64, xlab = expression(p[BH]~value),
     main = "Distribution of adjusted p-values - Mitra",
     cex.main = 1, cex.lab = 1, cex.axis = 1)

#' Plot the distribution of p-values in Capitainer
par(mfrow = c(1, 2))
hist(rowData(res)$pvalue.C, breaks = 64, xlab = "p value",
     main = "Distribution of raw p-values - Capitainer",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
hist(rowData(res)$adjp.C, breaks = 64, xlab = expression(p[BH]~value),
     main = "Distribution of adjusted p-values - Capitainer",
     cex.main = 1, cex.lab = 1, cex.axis = 1)

#' Plot the distribution of p-values in Blood
par(mfrow = c(1, 2))
hist(rowData(res)$pvalue.B, breaks = 64, xlab = "p value",
     main = "Distribution of raw p-values - Blood",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
hist(rowData(res)$adjp.B, breaks = 64, xlab = expression(p[BH]~value),
     main = "Distribution of adjusted p-values - Blood",
     cex.main = 1, cex.lab = 1, cex.axis = 1)

#' Plot the distribution of p-values in Sex
par(mfrow = c(1, 2))
hist(rowData(res)$pvalue.S, breaks = 64, xlab = "p value",
     main = "Distribution of raw p-values - Sex",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
hist(rowData(res)$adjp.S, breaks = 64, xlab = expression(p[BH]~value),
     main = "Distribution of adjusted p-values - Sex",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
```

Histograms shows high percentage of features present a very low p-value,
obtaining around 2000 significant features (\~1500 in Blood). This could be
because of the matrix difference between blood samples and plasma samples.

Below p-value are ploted against the log2 fold change of (average) abundances in
a Volcano plot. This volcano plot will allow us to visualize the features that
are significantly different between the two study groups.

### Volcano plot

```{r volcano-plot-plasma}
#' Plot volcano plot of the statistical results - Whatman
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res)$coef.W, -log10(rowData(res)$adjp.W),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - Plasma vs Whatman",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#0000ffcc")
if (any(rowData(res)$significant.W)) {
    points(rowData(res)$coef.W[rowData(res)$significant.W],
           -log10(rowData(res)$adjp.W[rowData(res)$significant.W]),
           col = "#0000ffcc")
}
#' Plot volcano plot of the statistical results - Mitra
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res)$coef.M, -log10(rowData(res)$adjp.M),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - Plasma vs Mitra",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#0000ffcc")
if (any(rowData(res)$significant.M)) {
    points(rowData(res)$coef.M[rowData(res)$significant.M],
           -log10(rowData(res)$adjp.M[rowData(res)$significant.M]),
           col = "#0000ffcc")
}
#' Plot volcano plot of the statistical results - Capitainer
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res)$coef.C, -log10(rowData(res)$adjp.C),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - Plasma vs Capitainer",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#0000ffcc")
if (any(rowData(res)$significant.C)) {
    points(rowData(res)$coef.C[rowData(res)$significant.C],
           -log10(rowData(res)$adjp.C[rowData(res)$significant.C]),
           col = "#0000ffcc")
}
#' Plot volcano plot of the statistical results - Blood
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res)$coef.B, -log10(rowData(res)$adjp.B),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - Plasma vs Blood",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#0000ffcc")
if (any(rowData(res)$significant.B)) {
    points(rowData(res)$coef.B[rowData(res)$significant.B],
           -log10(rowData(res)$adjp.B[rowData(res)$significant.B]),
           col = "#0000ffcc")
}
#' Plot volcano plot of the statistical results - sex
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res)$coef.S, -log10(rowData(res)$adjp.S),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - SEX",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#0000ffcc")
if (any(rowData(res)$significant.S)) {
    points(rowData(res)$coef.S[rowData(res)$significant.S],
           -log10(rowData(res)$adjp.S[rowData(res)$significant.S]),
           col = "#0000ffcc")
}
```

The interesting features would be on top corners of the volcano plot. However,
high amount of features seems to be significant in both groups.\
There features are listed, along with their average difference in log(2)
abundance between the compared groups, their adjusted p-value, their average
(log2) abundance in each sample group and their RSD in QC samples in the table
bellow.

NOTE: modify table adding columns from p-value in each device.

Compared to plasma samples, very large differences in feature abundances can be
observed. We next directly compare the coefficients of the different devices to
evaluate whether there might be differences also between devices.

```{r}
par(mfrow = c(1, 3))
plot(rowData(res)$coef.M, rowData(res)$coef.W, pch = 21, col = "#00000060",
     bg = "#00000020", xlab = expression(coef[Mitra]),
     ylab = expression(coef[Whatman]))
grid()
abline(0, 1)
plot(rowData(res)$coef.M, rowData(res)$coef.C, pch = 21, col = "#00000060",
     bg = "#00000020", xlab = expression(coef[Mitra]),
     ylab = expression(coef[Capitainer]))
grid()
abline(0, 1)
plot(rowData(res)$coef.W, rowData(res)$coef.C, pch = 21, col = "#00000060",
     bg = "#00000020", xlab = expression(coef[Whatman]),
     ylab = expression(coef[Capitainer]))
grid()
abline(0, 1)

```

While over and above the abundances between the microsampling devices are
similar/correlated, there is a distinct set of features with relatively low
abundances in Mitra, while in the other two microsampling devices the abundances
are higher. A similar observation can be made in the comparison between Whatman
and Capitainer.


### Table of significant features

```{r significant-features-plasma}
# Table of significant features in all groups
tab <- rowData(res)[rowData(res)$significant.W,
                    c("mzmed", "rtmed", 
                      "coef.W","pvalue.W","adjp.W", "avg.W",
                      "avg.P", "avg.M","avg.C", "avg.B", "avg.S",
                      "coef.M","pvalue.M","adjp.M",
                      "coef.C","pvalue.C","adjp.C", 
                      "coef.B","pvalue.B","adjp.B", 
                      "coef.S","pvalue.S","adjp.S")] |>
    as.data.frame()
tab <- tab[order(abs(tab$coef.W), decreasing = TRUE), ]
kable(tab, format = "pipe")
```

And a heatmap of the features that are significant for at least one
microsampling device:

```{r}
amap <- log2(assay(res, "norm_imputed"))
#' center the abundances by row (features) as we're only interested in
#' relative abundance differences between devices
amap <- amap - rowMeans(amap)
#' restrict to features that are significant for at leat one device
amap <- amap[rowData(res)$significant.M | rowData(res)$significant.W |
             rowData(res)$significant.B | rowData(res)$significant.C, ]

ann <- data.frame(device = colData(res)$device)
rownames(ann) <- colnames(amap)

pheatmap(amap, show_rownames = FALSE, show_colnames = FALSE,
         annotation_col = ann, annotation_colors = list(device = leg_full))

```

There is a set of features with consistently higher abundances in microsampling
devices compared to blood or plasma. Another set of features shows similarly
high abundances in Capitainer and Whatman, while lower abundances are present
for Mitra, Blood and Plasma. Yet another set of features has lower
concentrations in Plasma, while higher abundances are present for all other
sample types.

Below we visualise the EICs of the significant features to evaluate their
signal. We restrict the MS data set to study samples. Parameters: - keepFeatures
= TRUE: ensures that identified features are retained in the subset object. -
peakBg: defines the (background) color for each individual chromatographic peak
in the EIC object.

```{r eval=FALSE}
#FIX

#' Restrict the raw data to study samples.
lcms1_study <- lcms1[!idx_blank_qc, keepFeatures = TRUE]
# save("lcms1_study.RData")

#' Extract EICs for the significant features
eic_sign_test <- featureChromatograms(
    lcms1_study, features = "FT1714", expandRt = 5, filled = TRUE, chunksize = 4)
# PLOT ONLY ONE FEATURE TO CHECK
#' Plot the EICs.
par(mar = c(4, 4, 2, 1))  # Adjust bottom, left, top, right margins
plot(eic_sign_test, col = col_sample,
     peakBg = paste0(col_sample[chromPeaks(eic_sign_test)[, "sample"]], 60))
legend("topright", col = col_sample, legend = names(col_sample), lty = 1)
dev.off()  # Closes any active plotting devices

# save("lcms1_study.RData")
```

NOTE: fix. Error in plot.new() : figure margins too large. Is it related with
number of significant features?

In order to export the res object to perform the annotation in a different file,
I save it as res_sign_fts.

```{r res-save}
# Save res object
res_sts <- res
save(res_sts, file = "res_sts.RData")
```

#### Comparison with linear model participants

##### Linear Model with participants

The same way we compared in a linear model all the devices adn references
samples taking into account the sex variable, now we include participants as a
cofactor. In that case, Limma is returning NA value for the 10th participant and
giving error, but for the rest, I will compare the coef.W and -log10 pvalue
using both models.

```{r lm-participants}
#' Define the linear model to be applied to the data
p.cut <- 0.05     # cut-off for significance.
m.cut <- 0.5      # cut-off for log2 fold change

#' Define the linear model to be applied to the data using participants as a factor
device <- factor(x = res$device, levels =  c("PLASMA", "WHATMAN", "MITRA",
                                             "CAPITAINER", "BLOOD"))
sex <- factor(res$sex)
participants <- factor(res$participants)
design_part <- model.matrix(~ device + sex + participants)

#' Fit the linear model to the data, explaining metabolite
#' concentrations by device and sex.
fit_part <- lmFit(log2(assay(res, "norm_imputed")), design = design_part) # fit model to log2-transformed abundances
fit_part <- eBayes(fit_part) # calculate p-value
fit_part
```

```{r result-df-participants}
#' Compile a result data frame. 

tmp_part <- data.frame(
    coef.W.p = fit_part$coefficients[, "deviceWHATMAN"],
    pvalue.W.p = fit_part$p.value[, "deviceWHATMAN"],
    adjp.W.p = p.adjust(fit_part$p.value[, "deviceWHATMAN"], method = "BH"), # Benjamini-Hochberg correction
    avg.W.p = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "WHATMAN"])),
    avg.P.p = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "PLASMA"])), 
    coef.M.p = fit_part$coefficients[, "deviceMITRA"],
    pvalue.M.p = fit_part$p.value[, "deviceMITRA"],
    adjp.M.p = p.adjust(fit_part$p.value[, "deviceMITRA"], method = "BH"), # Benjamini-Hochberg correction
    avg.M.p = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "MITRA"])),
    coef.C.p = fit_part$coefficients[, "deviceCAPITAINER"],
    pvalue.C.p = fit_part$p.value[, "deviceCAPITAINER"],
    adjp.C.p = p.adjust(fit_part$p.value[, "deviceCAPITAINER"], method = "BH"), # Benjamini-Hochberg correction 
    avg.C.p = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "CAPITAINER"])),
    coef.B.p = fit_part$coefficients[, "deviceBLOOD"],
    pvalue.B.p = fit_part$p.value[, "deviceBLOOD"],
    adjp.B.p = p.adjust(fit_part$p.value[, "deviceBLOOD"], method = "BH"), # Benjamini-Hochberg correction
    avg.B.p = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "BLOOD"])),
    coef.S.p = fit_part$coefficients[, "sexW"],
    pvalue.S.p = fit_part$p.value[, "sexW"],
    adjp.S.p = p.adjust(fit_part$p.value[, "sexW"], method = "BH"), # Benjamini-Hochberg correction
    avg.S.p = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "SEX"]))
)

tmp_part$significant.W.p <- tmp_part$adjp.W.p < 0.05
tmp_part$significant.M.p <- tmp_part$adjp.M.p < 0.05
tmp_part$significant.C.p <- tmp_part$adjp.C.p < 0.05
tmp_part$significant.B.p <- tmp_part$adjp.B.p < 0.05
tmp_part$significant.S.p <- tmp_part$adjp.S.p < 0.05
#' Add differential abundance results to the object's rowData
rowData(res) <- cbind(rowData(res), tmp_part)
```


##### Plot Coef values with and without participant factor

```{r}
# Comparison of Coef in WHATMAN between two models. 

par(mfrow = c(1, 2))
#' Plot the values against each other (in log2 scale)
plot(rowData(res)$coef.W, rowData(res)$coef.W.p,
     xlim = range(c(rowData(res)$coef.W, rowData(res)$coef.W.p)),
     ylim = range(c(rowData(res)$coef.W, rowData(res)$coef.W.p)), 
     pch = 21, bg = "#00000020", col = "#00000080",  
     xlab = "y ~ device + sex",
     ylab = "y ~ device + sex + participants", 
     main = "Whatman Coefficients comparison"
     )
grid()
abline(0, 1)

plot(-log10(rowData(res)$pvalue.W), -log10(rowData(res)$pvalue.W.p),
     pch = 21, bg = "#00000020", col = "#00000080",  
     xlab = "y ~ device + sex",
     ylab = "y ~ device + sex + participants", 
     main = "Whatman -log10 p-values comparison"
     )
grid()
abline(0, 1)

```

The coefficients are identical while the p-values differ between the two models.

```{r}
# Comparison of Coef in MITRA between two models. 

par(mfrow = c(1, 2))
#' Plot the values against each other (in log2 scale)
plot(rowData(res)$coef.M, rowData(res)$coef.M.p,
     xlim = range(c(rowData(res)$coef.M, rowData(res)$coef.M.p)),
     ylim = range(c(rowData(res)$coef.M, rowData(res)$coef.M.p)), 
     pch = 21, bg = "#00000020", col = "#00000080",  
     xlab = "y~ device + sex",
     ylab = "y~ device + sex + participants", 
     main = "Mitra Coeficients comparison"
     )
grid()
abline(0, 1)

plot(-log10(rowData(res)$pvalue.M), -log10(rowData(res)$pvalue.M.p),
     pch = 21, bg = "#00000020", col = "#00000080",  
     xlab = "y ~ device + sex",
     ylab = "y ~ device + sex + participants", 
     main = "Mitra -log10 p-values comparison"
     )
grid()
abline(0, 1)

```

```{r}
# Comparison of Coef in CAPITAINER between two models. 

par(mfrow = c(1, 2))
#' Plot the values against each other (in log2 scale)
plot(rowData(res)$coef.C, rowData(res)$coef.C.p,
     xlim = range(c(rowData(res)$coef.C, rowData(res)$coef.C.p)),
     ylim = range(c(rowData(res)$coef.C, rowData(res)$coef.C.p)), 
     pch = 21, bg = "#00000020", col = "#00000080",  
     xlab = "y~ device + sex",
     ylab = "y~ device + sex + participants", 
     main = "Capitainer Coeficients comparison"
     )
grid()
abline(0, 1)

plot(-log10(rowData(res)$pvalue.C), -log10(rowData(res)$pvalue.C.p),
     pch = 21, bg = "#00000020", col = "#00000080",  
     xlab = "y ~ device + sex",
     ylab = "y ~ device + sex + participants", 
     main = "Capitainer -log10 p-values comparison"
     )
grid()
abline(0, 1)
```

```{r}
# Comparison of Coef in BLOOD between two models. 

par(mfrow = c(1, 2))
#' Plot the values against each other (in log2 scale)
plot(rowData(res)$coef.B, rowData(res)$coef.B.p,
     xlim = range(c(rowData(res)$coef.B, rowData(res)$coef.B.p)),
     ylim = range(c(rowData(res)$coef.B, rowData(res)$coef.B.p)), 
     pch = 21, bg = "#00000020", col = "#00000080",  
     xlab = "y~ device + sex",
     ylab = "y~ device + sex + participants", 
     main = "Blood Coeficients comparison"
     )
grid()
abline(0, 1)

plot(-log10(rowData(res)$pvalue.B), -log10(rowData(res)$pvalue.B.p),
     pch = 21, bg = "#00000020", col = "#00000080",  
     xlab = "y ~ device + sex",
     ylab = "y ~ device + sex + participants", 
     main = "Blood -log10 p-values comparison"
     )
grid()
abline(0, 1)
```

## Blood vs devices

since microsampling devices contains whole blood samples, it is useful to
include whole blood as a reference group when comparing microsampling devices.
By suing whole blood as the "intercept" in the linear model, we cab assess how
each microsampling device (Whatman, Capitainer, Mitra) differs in terms of
significant features or detection performance relative to traditional sampling.

### Linear model

```{r lm-blood}
#' Define the linear model to be applied to the data
p.cut <- 0.05     # cut-off for significance.
m.cut <- 0.5      # cut-off for log2 fold change
# The algorithm takes the first group as reference, so levels must be changed: 

#' Define the linear model to be applied to the data
device <- factor(x = res$device, levels = c("BLOOD", "WHATMAN", "MITRA",
                                            "CAPITAINER", "PLASMA"))
sex <- factor(res$sex)
# design <- model.matrix(~ device + sex + participants)
# As there are NA values in one of the participants we need to continue without participants. 
design <- model.matrix(~ device + sex)

#' Fit the linear model to the data, explaining metabolite
#' concentrations by device and sex.
fit_B <- lmFit(log2(assay(res, "norm_imputed")), design = design) # fit model to log2-transformed abundances
fit_B <- eBayes(fit_B) # calculate p-value
fit_B
```

Extract the results:

```{r result-df-blood}
#' Compile a result data frame, do we do one data.frame per device ? or for the full thing ? 
tmp_B <- data.frame(
    coef.W = fit_B$coefficients[, "deviceWHATMAN"],
    pvalue.W = fit_B$p.value[, "deviceWHATMAN"],
    adjp.W = p.adjust(fit_B$p.value[, "deviceWHATMAN"], method = "BH"), # Benjamini-Hochberg correction  
    avg.W = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "WHATMAN"])),
    avg.B = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "BLOOD"])),
    coef.M = fit_B$coefficients[, "deviceMITRA"],
    pvalue.M = fit_B$p.value[, "deviceMITRA"],
    adjp.M= p.adjust(fit_B$p.value[, "deviceMITRA"], method = "BH"), # Benjamini-Hochberg correction   
    avg.M = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "MITRA"])),
    coef.C = fit_B$coefficients[, "deviceCAPITAINER"],
    pvalue.C = fit_B$p.value[, "deviceCAPITAINER"],
    adjp.C = p.adjust(fit_B$p.value[, "deviceCAPITAINER"], method = "BH"), # Benjamini-Hochberg correction 
    avg.C = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "CAPITAINER"])),
    coef.S = fit_B$coefficients[, "sexW"],
    pvalue.S = fit_B$p.value[, "sexW"],
    adjp.S = p.adjust(fit_B$p.value[, "sexW"], method = "BH"), # Benjamini-Hochberg correction
    avg.S = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "SEX"]))
)
tmp$significant.W <- tmp_B$adjp.W < 0.05
tmp$significant.M <- tmp_B$adjp.M < 0.05
tmp$significant.C <- tmp_B$adjp.C < 0.05
tmp$significant.S <- tmp_B$adjp.S < 0.05

#' Add differential abundance results to the object's rowData
rowData(res) <- cbind(rowData(res), tmp_B)
```

Now we visualise the distribution of the raw and adjusted p-values:

### Distribution pvalues

```{r distribution-pvalues-blood}
#' Plot the distribution of p-values in Whatman
par(mfrow = c(1, 2))
hist(rowData(res)$pvalue.W, breaks = 64, xlab = "p value",
     main = "Distribution of raw p-values - Whatman",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
hist(rowData(res)$adjp.W, breaks = 64, xlab = expression(p[BH]~value),
     main = "Distribution of adjusted p-values - Whatman",
     cex.main = 1, cex.lab = 1, cex.axis = 1)

#' Plot the distribution of p-values in Mitra
par(mfrow = c(1, 2))
hist(rowData(res)$pvalue.M, breaks = 64, xlab = "p value",
     main = "Distribution of raw p-values - Mitra",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
hist(rowData(res)$adjp.M, breaks = 64, xlab = expression(p[BH]~value),
     main = "Distribution of adjusted p-values - Mitra",
     cex.main = 1, cex.lab = 1, cex.axis = 1)

#' Plot the distribution of p-values in Capitainer
par(mfrow = c(1, 2))
hist(rowData(res)$pvalue.C, breaks = 64, xlab = "p value",
     main = "Distribution of raw p-values - Capitainer",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
hist(rowData(res)$adjp.C, breaks = 64, xlab = expression(p[BH]~value),
     main = "Distribution of adjusted p-values - Capitainer",
     cex.main = 1, cex.lab = 1, cex.axis = 1)

#' Plot the distribution of p-values in Sex
par(mfrow = c(1, 2))
hist(rowData(res)$pvalue.S, breaks = 64, xlab = "p value",
     main = "Distribution of raw p-values - Sex",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
hist(rowData(res)$adjp.S, breaks = 64, xlab = expression(p[BH]~value),
     main = "Distribution of adjusted p-values - Sex",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
```

### Volcano plot

```{r volcano-plot-blood}
#' Plot volcano plot of the statistical results - Whatman
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res)$coef.W, -log10(rowData(res)$adjp.W),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - Blood vs Whatman",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#EB7F2C")
if (any(rowData(res)$significant.W)) {
    points(rowData(res)$coef.W[rowData(res)$significant.W],
           -log10(rowData(res)$adjp.W[rowData(res)$significant.W]),
           col = "#EB7F2C")
}
#' Plot volcano plot of the statistical results - Mitra
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res)$coef.M, -log10(rowData(res)$adjp.M),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - Blood vs Mitra",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#EB7F2C")
if (any(rowData(res)$significant.M)) {
    points(rowData(res)$coef.M[rowData(res)$significant.M],
           -log10(rowData(res)$adjp.M[rowData(res)$significant.M]),
           col = "#EB7F2C")
}
#' Plot volcano plot of the statistical results - Capitainer
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res)$coef.C, -log10(rowData(res)$adjp.C),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - Blood vs Capitainer",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#EB7F2C")
if (any(rowData(res)$significant.C)) {
    points(rowData(res)$coef.C[rowData(res)$significant.C],
           -log10(rowData(res)$adjp.C[rowData(res)$significant.C]),
           col = "#EB7F2C")
}

#' Plot volcano plot of the statistical results - Sex
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res)$coef.S, -log10(rowData(res)$adjp.S),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - SEX",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#EB7F2C")
if (any(rowData(res)$significant.S)) {
    points(rowData(res)$coef.S[rowData(res)$significant.S],
           -log10(rowData(res)$adjp.S[rowData(res)$significant.S]),
           col = "#EB7F2C")
}
```

The interesting features would be on top corners of the volcano plot. All
significant features have a negative coefficient (log 2 fold change value)
indicating that their abundance is lower in Whatman compared to Plasma samples.
There features are listed, along with their average difference in log(2)
abundance between the compared groups, their adjusted p-value, their average
(log2) abundance in each sample group and their RSD in QC samples in the table
bellow.

### Table of signigicant features

```{r significant-features-plasma}
# Table of significant features in all groups
tab_B <- rowData(res)[rowData(res)$significant.W,
                    c("mzmed", "rtmed", 
                      "coef.W","pvalue.W","adjp.W", "avg.W",
                      "avg.B", "avg.M","avg.C", "avg.S",
                      "coef.M","pvalue.M","adjp.M",
                      "coef.C","pvalue.C","adjp.C", 
                      "coef.S","pvalue.S","adjp.S")] |>
    as.data.frame()
tab_B <- tab_B[order(abs(tab_B$coef.W), decreasing = TRUE), ]
kable(tab_B, format = "pipe")
```

## Whatman vs Mitra

I also try to compare between devices as an example:

### Linear model

```{r lm-whatman}
#' Define the linear model to be applied to the data
p.cut <- 0.05     # cut-off for significance.
m.cut <- 0.5      # cut-off for log2 fold change
# The algorithm takes the first group as reference, so levels must be changed: 

#' Define the linear model to be applied to the data
device <- factor(x = res$device, levels = c("WHATMAN", "MITRA", "CAPITAINER", "BLOOD", "PLASMA"))
sex <- res$sex
# design <- model.matrix(~ device + sex + participants)
# As there are NA values in one of the participants we need to continue without participants. 

design <- model.matrix(~ device + sex)

#' Fit the linear model to the data, explaining metabolite
#' concentrations by device and sex.
fit_W <- lmFit(log2(assay(res, "norm_imputed")), design = design) # fit model to log2-transformed abundances
fit_W <- eBayes(fit_W) # calculate p-value
fit_W
```

Extract the results:

```{r result-df-whatman}
#' Compile a result data frame, do we do one data.frame per device ? or for the full thing ? 
tmp_W <- data.frame(
    coef.M = fit_W$coefficients[, "deviceMITRA"],
    pvalue.M = fit_W$p.value[, "deviceMITRA"],
    adjp.M= p.adjust(fit_W$p.value[, "deviceMITRA"], method = "BH"), # Benjamini-Hochberg correction 
    avg.M = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "MITRA"])),
    avg.W = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "WHATMAN"])),
    coef.C = fit_W$coefficients[, "deviceCAPITAINER"],
    pvalue.C = fit_W$p.value[, "deviceCAPITAINER"],
    adjp.C = p.adjust(fit_W$p.value[, "deviceCAPITAINER"], method = "BH"), # Benjamini-Hochberg correction
    avg.C = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "CAPITAINER"])),
    coef.S = fit_W$coefficients[, "sexW"],
    pvalue.S = fit_W$p.value[, "sexW"],
    adjp.S = p.adjust(fit_W$p.value[, "sexW"], method = "BH"), # Benjamini-Hochberg correction
    avg.S = rowMeans(
        log2(assay(res, "norm_imputed")[,res$device == "SEX"]))
)
tmp_W$significant.M <- tmp_W$adjp.M < 0.05
tmp_W$significant.C <- tmp_W$adjp.C < 0.05
tmp_W$significant.S <- tmp_W$adjp.S < 0.05

#' Add differential abundance results to the object's rowData
rowData(res) <- cbind(rowData(res), tmp_W)
```

Now we visualise the distribution of the raw and adjusted p-values:

### Distribution pvalues

```{r distribution-pvalues-3-whatman}

#' Plot the distribution of p-values in Mitra
par(mfrow = c(1, 2))
hist(rowData(res)$pvalue.M, breaks = 64, xlab = "p value",
     main = "Distribution of raw p-values - Mitra",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
hist(rowData(res)$adjp.M, breaks = 64, xlab = expression(p[BH]~value),
     main = "Distribution of adjusted p-values - Mitra",
     cex.main = 1, cex.lab = 1, cex.axis = 1)

#' Plot the distribution of p-values in Capitainer
par(mfrow = c(1, 2))
hist(rowData(res)$pvalue.C, breaks = 64, xlab = "p value",
     main = "Distribution of raw p-values - Capitainer",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
hist(rowData(res)$adjp.C, breaks = 64, xlab = expression(p[BH]~value),
     main = "Distribution of adjusted p-values - Capitainer",
     cex.main = 1, cex.lab = 1, cex.axis = 1)

```

### Volcano plot

```{r volcano-plot-whatman}
#' Plot volcano plot of the statistical results - Mitra
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res)$coef.M, -log10(rowData(res)$adjp.M),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - Whatman vs Mitra",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#1BE959")
if (any(rowData(res)$significant.M)) {
    points(rowData(res)$coef.M[rowData(res)$significant.M],
           -log10(rowData(res)$adjp.M[rowData(res)$significant.M]),
           col = "#1BE959")
}
#' Plot volcano plot of the statistical results - Capitainer
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res)$coef.C, -log10(rowData(res)$adjp.C),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - Whatman vs Capitainer",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#1BE959")
if (any(rowData(res)$significant.C)) {
    points(rowData(res)$coef.C[rowData(res)$significant.C],
           -log10(rowData(res)$adjp.C[rowData(res)$significant.C]),
           col = "#1BE959")
}

#' Plot volcano plot of the statistical results - Sex
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res)$coef.S, -log10(rowData(res)$adjp.S),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - SEX",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#1BE959")
if (any(rowData(res)$significant.S)) {
    points(rowData(res)$coef.S[rowData(res)$significant.S],
           -log10(rowData(res)$adjp.S[rowData(res)$significant.S]),
           col = "#1BE959")
}
```

The comparison between plasma vs devices and blood vs devices, look the same.
The number of significant features seems to be quite high taking into account
that there are from same patients. Is there any error in the code? Are we
overfitting the lm?

### Table of signigicant features

```{r significant-features-plasma}
# Table of significant features in all groups
tab_devices <- rowData(res)[rowData(res)$significant.W,
                    c("mzmed", "rtmed", 
                      "coef.W","pvalue.W","adjp.W", "avg.W",
                      "avg.P", "avg.M","avg.C", "avg.B", "avg.S",
                      "coef.M","pvalue.M","adjp.M",
                      "coef.C","pvalue.C","adjp.C", 
                      "coef.B","pvalue.B","adjp.B", 
                      "coef.S","pvalue.S","adjp.S")] |>
    as.data.frame()
tab_devices <- tab_devices[order(abs(tab_devices$coef.W), decreasing = TRUE), ]
kable(tab_devices, format = "pipe")
```

## LINEAR MODEL - Whatman ~ sex

In this case, I create a linear model where y is the log 2 abundance of each metabolite in each sample, using sex as a binary variable with women/men. The aim is to study if each device of reference sample is able to find significant differences related with the sex. 

```{r lm plasma}
#' Define the linear model to be applied to the data
p.cut <- 0.05     # cut-off for significance.
m.cut <- 0.5      # cut-off for log2 fold change

#' Define the linear model to be applied to the data
sex <- res_w$sex
design <- model.matrix(~ + sex)

#' Fit the linear model to the data, explaining metabolite
#' concentrations by device and sex.
fit <- lmFit(log2(assay(res_w, "norm_imputed")), design = design) # fit model to log2-transformed abundances
fit <- eBayes(fit) # calculate p-value
fit
```

Extract the results:

```{r result-df-plasma}
#' Compile a result data frame, do we do one data.frame per device ? or for the full thing ? 
tmp <- data.frame(
    coef.S = fit$coefficients[, "sexW"],
    pvalue.S = fit$p.value[, "sexW"],
    adjp.S = p.adjust(fit$p.value[, "sexW"], method = "BH"), # Benjamini-Hochberg correction
    avg.S = rowMeans(
        log2(assay(res_w, "norm_imputed")[,res_w$device == "SEX"]))
)

# tmp$significant.W <- tmp$adjp.W < 0.05
tmp$significant.S <- tmp$adjp.S < 0.05
#' Add differential abundance results to the object's rowData
rowData(res_w) <- cbind(rowData(res_w), tmp)
```

Now we visualise the distribution of the raw and adjusted p-values:

### Distribution pvalues - Whatman ~ sex

```{r distribution-pvalues-plasma}
#' Plot the distribution of p-values in Sex
par(mfrow = c(1, 2))
hist(rowData(res_w)$pvalue.S, breaks = 64, xlab = "p value",
     main = "Distribution of raw p-values - Sex",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
hist(rowData(res_w)$adjp.S, breaks = 64, xlab = expression(p[BH]~value),
     main = "Distribution of adjusted p-values - Sex",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
```

### Volcano plot - Whatman ~ sex

```{r volcano-plot-plasma}
#' Plot volcano plot of the statistical results - sex
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res_w)$coef.S, -log10(rowData(res_w)$adjp.S),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - Whatman (SEX)",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#82db16")
if (any(rowData(res_w)$significant.S)) {
    points(rowData(res_w)$coef.S[rowData(res_w)$significant.S],
           -log10(rowData(res_w)$adjp.S[rowData(res_w)$significant.S]),
           col = "#82db16")
}
```

### Table of signigicant features - Whatman ~ sex

```{r significant-features-plasma}
# Table of significant features in all groups
tab <- rowData(res_w)[rowData(res_w)$significant.S,
                    c("mzmed", "rtmed", 
                      "coef.S","pvalue.S","adjp.S")] |>
    as.data.frame()
tab <- tab[order(abs(tab$coef.S), decreasing = TRUE), ]
kable(tab, format = "pipe")
```

## LINEAR MODEL - Mitra ~ sex


```{r lm plasma}
#' Define the linear model to be applied to the data
p.cut <- 0.05     # cut-off for significance.
m.cut <- 0.5      # cut-off for log2 fold change

#' Define the linear model to be applied to the data
sex <- res_m$sex
design <- model.matrix(~ + sex)

#' Fit the linear model to the data, explaining metabolite
#' concentrations by device and sex.
fit <- lmFit(log2(assay(res_m, "norm_imputed")), design = design) # fit model to log2-transformed abundances
fit <- eBayes(fit) # calculate p-value
fit
```

Extract the results:

```{r result-df-plasma}
#' Compile a result data frame, do we do one data.frame per device ? or for the full thing ? 
tmp <- data.frame(
    coef.S = fit$coefficients[, "sexW"],
    pvalue.S = fit$p.value[, "sexW"],
    adjp.S = p.adjust(fit$p.value[, "sexW"], method = "BH"), # Benjamini-Hochberg correction
    avg.S = rowMeans(
        log2(assay(res_m, "norm_imputed")[,res_m$device == "SEX"]))
)

# tmp$significant.W <- tmp$adjp.W < 0.05
tmp$significant.S <- tmp$adjp.S < 0.05
#' Add differential abundance results to the object's rowData
rowData(res_m) <- cbind(rowData(res_m), tmp)
```

Now we visualise the distribution of the raw and adjusted p-values:

### Distribution pvalues - Mitra ~ sex

```{r distribution-pvalues-plasma}
#' Plot the distribution of p-values in Sex
par(mfrow = c(1, 2))
hist(rowData(res_m)$pvalue.S, breaks = 64, xlab = "p value",
     main = "Distribution of raw p-values - Sex",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
hist(rowData(res_m)$adjp.S, breaks = 64, xlab = expression(p[BH]~value),
     main = "Distribution of adjusted p-values - Sex",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
```

### Volcano plot - Mitra ~ sex

```{r volcano-plot-plasma}
#' Plot volcano plot of the statistical results - sex
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res_m)$coef.S, -log10(rowData(res_m)$adjp.S),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - Mitra (SEX)",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#a90cdf")
if (any(rowData(res_m)$significant.S)) {
    points(rowData(res_m)$coef.S[rowData(res_m)$significant.S],
           -log10(rowData(res_m)$adjp.S[rowData(res_m)$significant.S]),
           col = "#a90cdf")
}
```

### Table of signigicant features - Mitra ~ sex

```{r significant-features-plasma}
# Table of significant features in all groups
tab <- rowData(res_m)[rowData(res_m)$significant.S,
                    c("mzmed", "rtmed", 
                      "coef.S","pvalue.S","adjp.S")] |>
    as.data.frame()
tab <- tab[order(abs(tab$coef.S), decreasing = TRUE), ]
kable(tab, format = "pipe")
```

## LINEAR MODEL - Capitainer ~ sex

```{r lm plasma}
#' Define the linear model to be applied to the data
p.cut <- 0.05     # cut-off for significance.
m.cut <- 0.5      # cut-off for log2 fold change

#' Define the linear model to be applied to the data
sex <- res_c$sex
design <- model.matrix(~ + sex)

#' Fit the linear model to the data, explaining metabolite
#' concentrations by device and sex.
fit <- lmFit(log2(assay(res_c, "norm_imputed")), design = design) # fit model to log2-transformed abundances
fit <- eBayes(fit) # calculate p-value
fit
```

Extract the results:

```{r result-df-plasma}
#' Compile a result data frame, do we do one data.frame per device ? or for the full thing ? 
tmp <- data.frame(
    coef.S = fit$coefficients[, "sexW"],
    pvalue.S = fit$p.value[, "sexW"],
    adjp.S = p.adjust(fit$p.value[, "sexW"], method = "BH"), # Benjamini-Hochberg correction
    avg.S = rowMeans(
        log2(assay(res_c, "norm_imputed")[,res_c$device == "SEX"]))
)

# tmp$significant.W <- tmp$adjp.W < 0.05
tmp$significant.S <- tmp$adjp.S < 0.05
#' Add differential abundance results to the object's rowData
rowData(res_c) <- cbind(rowData(res_c), tmp)
```

Now we visualise the distribution of the raw and adjusted p-values:

### Distribution pvalues - Capitainer ~ sex

```{r distribution-pvalues-plasma}
#' Plot the distribution of p-values in Sex
par(mfrow = c(1, 2))
hist(rowData(res_c)$pvalue.S, breaks = 64, xlab = "p value",
     main = "Distribution of raw p-values - Sex",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
hist(rowData(res_c)$adjp.S, breaks = 64, xlab = expression(p[BH]~value),
     main = "Distribution of adjusted p-values - Sex",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
```

### Volcano plot - Capitainer ~ sex

```{r volcano-plot-plasma}
#' Plot volcano plot of the statistical results - sex
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res_w)$coef.S, -log10(rowData(res_w)$adjp.S),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - Capitainer (SEX)",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#57c1f7")
if (any(rowData(res_w)$significant.S)) {
    points(rowData(res_w)$coef.S[rowData(res_w)$significant.S],
           -log10(rowData(res_w)$adjp.S[rowData(res_w)$significant.S]),
           col = "#57c1f7")
}
```

### Table of signigicant features - Capitainer ~ sex

```{r significant-features-plasma}
# Table of significant features in all groups
tab <- rowData(res_w)[rowData(res_w)$significant.S,
                    c("mzmed", "rtmed", 
                      "coef.S","pvalue.S","adjp.S")] |>
    as.data.frame()
tab <- tab[order(abs(tab$coef.S), decreasing = TRUE), ]
kable(tab, format = "pipe")
```

## LINEAR MODEL - Plasma ~ sex

```{r lm plasma}
#' Define the linear model to be applied to the data
p.cut <- 0.05     # cut-off for significance.
m.cut <- 0.5      # cut-off for log2 fold change

#' Define the linear model to be applied to the data
sex <- res_p$sex
design <- model.matrix(~ + sex)

#' Fit the linear model to the data, explaining metabolite
#' concentrations by device and sex.
fit <- lmFit(log2(assay(res_p, "norm_imputed")), design = design) # fit model to log2-transformed abundances
fit <- eBayes(fit) # calculate p-value
fit
```

Extract the results:

```{r result-df-plasma}
#' Compile a result data frame, do we do one data.frame per device ? or for the full thing ? 
tmp <- data.frame(
    coef.S = fit$coefficients[, "sexW"],
    pvalue.S = fit$p.value[, "sexW"],
    adjp.S = p.adjust(fit$p.value[, "sexW"], method = "BH"), # Benjamini-Hochberg correction
    avg.S = rowMeans(
        log2(assay(res_p, "norm_imputed")[,res_p$device == "SEX"]))
)

# tmp$significant.W <- tmp$adjp.W < 0.05
tmp$significant.S <- tmp$adjp.S < 0.05
#' Add differential abundance results to the object's rowData
rowData(res_p) <- cbind(rowData(res_p), tmp)
```

Now we visualise the distribution of the raw and adjusted p-values:

### Distribution pvalues - Plasma ~ sex

```{r distribution-pvalues-plasma}
#' Plot the distribution of p-values in Sex
par(mfrow = c(1, 2))
hist(rowData(res_p)$pvalue.S, breaks = 64, xlab = "p value",
     main = "Distribution of raw p-values - Sex",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
hist(rowData(res_p)$adjp.S, breaks = 64, xlab = expression(p[BH]~value),
     main = "Distribution of adjusted p-values - Sex",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
```

### Volcano plot - Plasma ~ sex

```{r volcano-plot-plasma}
#' Plot volcano plot of the statistical results - sex
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res_p)$coef.S, -log10(rowData(res_p)$adjp.S),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - Plasma (SEX)",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#dfdc16")
if (any(rowData(res_p)$significant.S)) {
    points(rowData(res_p)$coef.S[rowData(res_p)$significant.S],
           -log10(rowData(res_p)$adjp.S[rowData(res_p)$significant.S]),
           col = "#dfdc16")
}
```

### Table of signigicant features - Plasma ~ sex

```{r significant-features-plasma}
# Table of significant features in all groups
tab <- rowData(res_p)[rowData(res_p)$significant.S,
                    c("mzmed", "rtmed", 
                      "coef.S","pvalue.S","adjp.S")] |>
    as.data.frame()
tab <- tab[order(abs(tab$coef.S), decreasing = TRUE), ]
kable(tab, format = "pipe")
```


## LINEAR MODEL - Blood ~ sex

```{r lm plasma}
#' Define the linear model to be applied to the data
p.cut <- 0.05     # cut-off for significance.
m.cut <- 0.5      # cut-off for log2 fold change

#' Define the linear model to be applied to the data
sex <- res_b$sex
design <- model.matrix(~ + sex)

#' Fit the linear model to the data, explaining metabolite
#' concentrations by device and sex.
fit <- lmFit(log2(assay(res_b, "norm_imputed")), design = design) # fit model to log2-transformed abundances
fit <- eBayes(fit) # calculate p-value
fit 
```

Extract the results:

```{r result-df-plasma}
#' Compile a result data frame, do we do one data.frame per device ? or for the full thing ? 
tmp <- data.frame(
    coef.S = fit$coefficients[, "sexW"],
    pvalue.S = fit$p.value[, "sexW"],
    adjp.S = p.adjust(fit$p.value[, "sexW"], method = "BH"), # Benjamini-Hochberg correction
    avg.S = rowMeans(
        log2(assay(res_b, "norm_imputed")[,res_b$device == "SEX"]))
)

# tmp$significant.W <- tmp$adjp.W < 0.05
tmp$significant.S <- tmp$adjp.S < 0.05
#' Add differential abundance results to the object's rowData
rowData(res_b) <- cbind(rowData(res_b), tmp)
```

Now we visualise the distribution of the raw and adjusted p-values:

### Distribution pvalues - Blood ~ sex

```{r distribution-pvalues-plasma}
#' Plot the distribution of p-values in Sex
par(mfrow = c(1, 2))
hist(rowData(res_b)$pvalue.S, breaks = 64, xlab = "p value",
     main = "Distribution of raw p-values - Sex",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
hist(rowData(res_b)$adjp.S, breaks = 64, xlab = expression(p[BH]~value),
     main = "Distribution of adjusted p-values - Sex",
     cex.main = 1, cex.lab = 1, cex.axis = 1)
```

### Volcano plot - Blood ~ sex

```{r volcano-plot-plasma}
#' Plot volcano plot of the statistical results - sex
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res_b)$coef.S, -log10(rowData(res_b)$adjp.S),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - Blood (SEX)",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#f25800")
if (any(rowData(res_b)$significant.S)) {
    points(rowData(res_b)$coef.S[rowData(res_b)$significant.S],
           -log10(rowData(res_b)$adjp.S[rowData(res_b)$significant.S]),
           col = "#f25800")
}
```

### Table of signigicant features - Blood ~ sex

```{r significant-features-plasma}
# Table of significant features in all groups
tab <- rowData(res_b)[rowData(res_b)$significant.S,
                    c("mzmed", "rtmed", 
                      "coef.S","pvalue.S","adjp.S")] |>
    as.data.frame()
tab <- tab[order(abs(tab$coef.S), decreasing = TRUE), ]
kable(tab, format = "pipe")
```

We can confirm that devices are no able to bring significant differences between
men and women. Only one feature in plasma was significantly different between
women and men with a adjusted pvalue > 0.05.


