---
title: "Data analysis of Pilot Study from Imperial - annotation_LPOS_CONV."
format: html
tbl-cap-location: bottom
editor: visual
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
author: Philippine Louail, Ana Sanchez Lorenzo
---

```{r setup, include=FALSE}
library(knitr)
library(quarto)
knitr::opts_knit$set(root.dir = './')
```

# Loading packages

```{r packages-used, message=FALSE, warning=FALSE}
#| code-fold: true
#| code-summary: "Show the code"
## General bioconductor package
library(BiocStyle)
library(BiocParallel)
library(knitr)

## Data Import and handling
library(readxl)
library(MsExperiment)
#library(MsIO)
#library(alabaster.se)
library(SummarizedExperiment)

## Preprocessing of LC-MS data
library(xcms)
library(Spectra)
library(MetaboCoreUtils)

## Statistical analysis
library(limma) # Differential abundance
library(matrixStats) # Summaries over matrices

## Visualisation
library(pander)
library(RColorBrewer)
library(pheatmap)
library(vioplot)
library(ggfortify)   # Plot PCA
library(gridExtra)   # To arrange multiple ggplots into single plots
library(grid)
library(UpSetR)

## Annotation
library(AnnotationHub) # Annotation resources
library(CompoundDb)    # Access small compound annotation data.
library(MetaboAnnotation) # Functionality for metabolite annotation.
```

# DATA ORGANISATION

## Data import

In this case, we only import the SumExp with significant features details
(res_sts).

```{r import, message=FALSE}
# Data import
load("preprocessed_lcms1.RData")
load("res_sts.RData")
featureDefinitions(lcms1) |>head()
```

```{r indexing, message=FALSE}
# Index per sample group
idx_qc <- sampleData(lcms1)$device == "QC"
idx_blank <- sampleData(lcms1)$device == "BLANK"
idx_blank_qc <- sampleData(lcms1)$device %in% c("QC","BLANK")
idx_samples <- sampleData(lcms1)$device %in% c("WHATMAN", "MITRA", "CAPITAINER", "PLASMA", "BLOOD")
idx_bl_sp <- sampleData(lcms1)$blank_type %in% c("BLANK_W", "BLANK_C", "BLANK_M")
```

```{r par-process}#' Set up parallel processing using 3 cores}
if (.Platform$OS.type == "unix") {
  register(MulticoreParam(4)) ### chunksize has to be be the same number. 
 } else {
     register(SnowParam(4))
}
```

## Define colors

We will define colors for each of the sample groups based on their sample group
using the *RColorBrewer* package:

```{r define-colors}
#| echo: true
#' Define colors for full dataset
leg_full <- brewer.pal(12, name = "Set3")[c(1, 12, 7, 5, 10, 4, 6)]
names(leg_full) <- unique(sampleData(lcms1)$device)
leg_full["QC"] <- "#f00e0e"
col_sample <- leg_full[sampleData(lcms1)$device]

#' define colors for devices only
leg_blank_BuS <- leg_full[-c(1, 2, 6, 7)]
col_blank_BuS <- leg_blank_BuS[sampleData(lcms1[!idx_bl_sp])$device]
leg_blank_BuS <- leg_blank_BuS[c("WHATMAN", "MITRA", "CAPITAINER")]
```

## Participants column addition For Differential abundance analysis

We need to specify participants. For that we need to add a new column from
sample_name column information in lcms1 object and res object.

```{r}
#' Create a new column with participants names without device specification.
sampleData(lcms1)$participants <- substr(sampleData(lcms1)$sample_name, 1, nchar(sampleData(lcms1)$sample_name) -1)
sampleData(lcms1)$participants[idx_blank_qc] <- NA_character_
colData(res_sts)$participants <- substr(colData(res_sts)$sample_name, 1, nchar(colData(res_sts)$sample_name) - 1)
```

# ANNOTATION

## MS1 Annotation

After the differential abundance analysis, the next step is to identify the
features with significant differences in abundances between plasma and the rest
of groups. As the data was acquired using the NPC LPOS method, we could target
the compounds from the NPC library. However, as LC set-up was different
(different sample manager and different tubing length), only m/z values will be
able to match, fnot rt values.

To do the matching of m/z values, MetaboAnnotation package is used. This is the
lowest level of confidence in metabolite annotation. However, it gives ideas
about of metabolites present in samples.

To start with, I imported the library of metabolites called targed_df.

```{r import-ms1-features}
# Import NPC-LPOS library from csv file
target_df <- read.csv("LPOS_ROI_V_5_1_2.csv")
target_df <- as.data.frame(target_df)  # Ensure it is a data frame
head(target_df)  
```

As we know there is a shift in the rt due to the LC set up and tube length I
would change the rt column of the library for this specific dataset. If I check
the rt of "known compounds" EIC there is a shift +30 seconds in comparison with
the rt from the library. However, I will increase the rt range in case there are
some differences between features.

```{r}
# Modify column rt ranges from library
colnames(target_df)
# fit rt ranges
target_df$rtMin <- target_df$rtMin + 25
target_df$rtMax <- target_df$rtMax + 35
target_df$rt <- target_df$rt + 30
target_df$rt_minutes <- target_df$rt_minutes + 0.5
#target and query need to have same rt column name
target_df$rtmed <- target_df$rt 
# check it is working
head(target_df)
```

With this modification, I am trying to do MS1 annotation with m/z and rt values.

First, I prepared the data.frame with experimental features detected:

```{r query-df}
# Prepare query data frame
rowData(res_sts)$feature_id <- rownames(rowData(res_sts))
#cont with full datset
rowData(res_sts)
# Convert to data frame
res_sts_df <- as.data.frame(rowData(res_sts))
res_sts_df
```

Now, we do the matching in comparison with the target dataframe with reference
known compounds. As we already have the ion description in the library, we
include those adducts.

NOTE: Metabonaut: tolerance = 0, ppm = 5; MetaboAnnotation: tolerance = 0.005,
ppm = 0. No need to be that restrictive as I am using my own library with
metabolites already identified with standards in my lab with same method.

The resulting Matched object (with 10 ppm) shows that 362 out of 1623
significant features were matched against at least one target compound. Out of
the 378 identify ions in NPC library, we are able to match 226 metabolites. If
we try to do MS1 annotation matching m/z abd retention time values, we expect
the match to be more accurate because rt will restrict the match and if features
are detected across the analysis it will only identify the features that matched
with biological compounds:

```{r}
parm <- Mass2MzRtParam(adducts = c("[M+H]+", "[M+Na]+", "[M+NH4]+", "[M+H-H2O]+",  "[M+K]+", "[M]+"),
                       tolerance = 0, ppm = 10,
                       toleranceRt = 10)
mtch <- matchValues(res_sts, target_df, param = parm,
                                rtColname = "rtmed", mzColname = "mzmed",
                                massColname = "Monoisotopic_mass")
mtch
```

After matching mz and rt, as expected, we have less matches with the target
object (159 at of 378 annotations; only mz = 226/378). However, it is finding
same compound at different rts in query object. In total, we will have 245
annotations but we will have not manually check those ones. I will continue the
annotation with this matched object.

The result object (mtch) contains the full query data.frame and target
data.frame as well as the matching information. To visualize the info from mtch:

```{r visualisation1}
# head(query(mtch))
```

```{r visualisation2}
# head(target(mtch))
```

```{r visualisation3}
# whichQuery(mtch)
```

```{r visualisation4}
# whichTarget(mtch)
```

```{r visualisation5}
# colnames(mtch)
```

```{r extraction-match}
# To extract only the object of the matched features
mtch <- mtch[whichQuery(mtch)]
matchedData(mtch)|> head()
nrow(matchedData(mtch))
```

```{r visualisation5}
mtch$target_cpdName
```

Below I extract the full result from the Matched object:

```{r extration-match}
#' Extracting the results
mtch_res <- matchedData(mtch, c("feature_id", "mzmed", "rtmed", "adduct", 
                                "target_ion", "ppm_error", 
                                "target_chemicalFormula", 
                                "target_Monoisotopic_mass", 
                                "target_cpdName", "target_InChI", 
                                "coef.W", "coef.M", "coef.C", "coef.B", 
                                "adjp.W", "adjp.M", "adjp.C", "adjp.B" ))

mtch_res
```

\[from here it is from metanonaut\]

To reduce redundancy in the result table we below iterate over the hits of a
feature and keep only matches to unique compounds (identified by their
INCHIKEY). The INCHI name combine information from compound’s chemical formula
and structure, so while different compounds can share the same chemical formula,
they should have a different structure and thus INCHI.

```{r}
rownames(mtch_res) <- NULL

#' Keep only info on features that machted - create a utility function for that
mtch_res <- split(mtch_res, mtch_res$feature_id) |>
    lapply(function(x) {
        lapply(split(x, x$target_InChI), function(z) {
            z[which.min(z$ppm_error), ]
        })
    }) |>
    unlist(recursive = FALSE) |>
    do.call(what = rbind)

#NOTE WHAT MEANS THIS FUNCTION?

#' Display the results
kable(mtch_res, format = "pipe")

nrow(mtch_res)
mtch_res <- cbind(mtch_res, rowData(res_sts)[mtch_res$feature_id, 
                                             c("coef.W", "adjp.B")])
mtch_res

```

```{r}
# Add all coef and adjp from res_sts. 
nrow(mtch_res)
mtch_res <- cbind(mtch_res, rowData(res_sts)[mtch_res$feature_id, 
                                             c("coef.W"
                                               ,"coef.M", "coef.C",
                                                "coef.B",
                                               "adjp.W", "adjp.M", "adjp.C",
                                                "adjp.B"
                                               )])
mtch_res
```

```{r}
mtch_res_df <- as.data.frame(mtch_res)
library(writexl)
write_xlsx(mtch_res_df, "mtch_res_output.xlsx")
```

Only 195 match. Let see how to make sure the ion matched is the expected.

```{r}
# Filter rows where 'adduct' and 'target_ion' are the same
mtch_res_filtered <- mtch_res[mtch_res$adduct == mtch_res$target_ion, ] 
nrow(mtch_res_filtered)
```

Only 77 compounds have the same ion.

```{r}
mtch_res_filtered_df <- as.data.frame(mtch_res_filtered)
write_xlsx(mtch_res_filtered_df, "mtch_res_filtered_output.xlsx")
```
FIX: add names of all sign features. 
While being of low confidence, MS1-based annotation can provide first candidate
annotations that could be confirmed or rejected by additional analyses.

## MS2 Annotation

LC MS/MS data were recorded in data depending acquisition (DDA) in pooled
samples (and QCs samples). However, method was not fully optimised and it seems
MS2 data points will not be really abundance in those samples. I would like to
test the data to see how do I need to modify the method.

The MS2 data was acquited in 7 different samples:

-   pool sample from each device (Individual 11C, 11W, 11M)

-   pool general sample (QC)

-   pool sample from each device (QCC, QCW, QCM)

In total we have 4 measurements from each samples with different mz ranges:

-   001: mz range from 50 to 400 Da

-   002: mz range from 400 to 800 Da

-   003: mz range from 800 to 1200 Da

-   004: mz range from 50 to 1200 Da.

This was performed because we experienced a mass sensibility issue in the
instrument and by decreasing the mz range sensibility can increase.

First step is to load LC-MS/MS data and subset it now only for general QC
samples.

```{r import-files}
#' read the sample descriptions from an xlsx sheet
pd_MS2 <- read_xlsx("metadata_msms_lcms_rpos.xlsx", sheet = "samples") |>
    as.data.frame()

#' Import MS2 files
MS2_mzml_path <- file.path("data/lpos_conv_DDA") 

lcms2 <- readMsExperiment(file.path(MS2_mzml_path, "/",pd_MS2$mzML_file),
                          sampleData = pd_MS2)
lcms2
```

```{r}
# Add mzrange data information (from file names)
sampleData(lcms2)$fragmentation_mode <- c("001", "002", "003", "004")

#' Subset only samples with mz range form 50-1200Da. 
idx_mzrange <- sampleData(lcms2)$fragmentation_mode =="004"
lcms2_004 <- lcms2[idx_mzrange]

#let's look at the updated sample data
sampleData(lcms2_004)
colnames(sampleData(lcms2_004)) 
sampleData(lcms2_004) |> kable (format = "pipe")
```

MS2 Data is filtered to the same RT range as LC-MS run to clean unnecessary
peaks that will not be of interest

```{r}
#' Filter the data to the same RT range as the LC-MS run
lcms2_004 <- filterRt(lcms2_004, c(60, 700))
```

Below we show the number of MS1 and MS2 spectra for each of the files.

```{r}
#' check the number of spectra per ms level
spectra(lcms2_004) |>
    msLevel() |>
    split(spectraSampleIndex(lcms2_004)) |>
    lapply(table) |>
    do.call(what = cbind)
```

NOTE: same number of spectra in MS1 and MS2 (suspicious?) sample 6 is the only
one with one spectrum differents.

```{r}
sampleData(lcms2_004)$sample_name
```

In the plot below we can observe the different mz ranges in each sample
measurement.

```{r}
#|fig: "Figure XX. Precursor ions in the LC-MS/MS data set."

# par (mfrow = c(7, 1), mar = c(2, 2.5, 2, 4))
plotPrecursorIons(lcms2_004[1,], cex = 0.5, 
                  xlab = "retention time (s)",
                  ylab = "m/z ratio",
                  main = sampleData(lcms2_004)$sample_name[1])
plotPrecursorIons(lcms2_004[2,], cex = 0.5, 
                  xlab = "retention time (s)",
                  ylab = "m/z ratio",
                  main = sampleData(lcms2_004)$sample_name[2])
plotPrecursorIons(lcms2_004[3,], cex = 0.5, 
                  xlab = "retention time (s)",
                  ylab = "m/z ratio",
                  main = sampleData(lcms2_004)$sample_name[3])
plotPrecursorIons(lcms2_004[4,], cex = 0.5, 
                  xlab = "retention time (s)",
                  ylab = "m/z ratio",
                  main = sampleData(lcms2_004)$sample_name[4])
plotPrecursorIons(lcms2_004[5,], cex = 0.5, 
                  xlab = "retention time (s)",
                  ylab = "m/z ratio",
                  main = sampleData(lcms2_004)$sample_name[5])
plotPrecursorIons(lcms2_004[6,], cex = 0.5, 
                  xlab = "retention time (s)",
                  ylab = "m/z ratio",
                  main = sampleData(lcms2_004)$sample_name[6])
plotPrecursorIons(lcms2_004[7,], cex = 0.5, 
                  xlab = "retention time (s)",
                  ylab = "m/z ratio",
                  main = sampleData(lcms2_004)$sample_name[7])
 
```

We observed a high amount of low mass molecules and few high mass molecules. All
this precursor ions should have MSMS data or they are MS1 ions? How the If they
are only precursor ions,

We next extract the Spectra object with the MS data from the data object and
assign a new spectra variable with the employed collision energy, which we
extract from the data object sampleData.

For the next steps MS data is filtered by:

-   restricting MS2 spectra

-   removing mass peaks from each spectrum with an intensity that is lower than
    5% of the highest intensity for that spectrum, assuming that these very low
    intensity peaks represent background signal.

-   removing also mass peaks with an m/z value greater or equal to the precursor
    m/z of the ion. This puts, for the later matching against reference spectra,
    more weight on the fragmentation pattern of ions and avoids hits based only
    on the precursor m/z peak (and hence a similar mass of the compared
    compounds).

-   restricting the data to spectra with at least two fragment peaks and scale
    their intensities such that their sum is 1 for each spectrum.

While similarity calculations are not affected by this scaling, it makes visual
comparison of fragment spectra easier to read.

```{r}
#' extract spectra object
ms2_004 <- spectra(lcms2_004)

#' Remove low intensity peaks
low_int <- function(x, ...) {
    x > max(x, na.rm = TRUE) * 0.05
}
ms2_004 <- filterMsLevel(ms2_004, 2L) |>
    filterIntensity(intensity = low_int) 

#' Remove precursor peaks and restrict to spectra with a minimum number of peaks
ms2_004 <- filterPrecursorPeaks(ms2_004, ppm = 50, mz = ">=")
ms2_004 <- ms2_004[lengths(ms2_004) > 1] |>
    scalePeaks()
```

```{r}
#' Load the MS data subset into memory
ms2_004 <- setBackend(ms2_004, MsBackendMemory())
ms2_004 <- applyProcessing(ms2_004)

#' Size of the object after loading into memory
print(object.size(ms2_004), units = "MB")
```
NOTE: how is the preprocessing affecting the next code? Why if I do not perform
it it gives error related with dimention. 

The goal of the workflow is to *identify the MS2 spectra* from the second
(LC-MS/MS) run that could represent fragments of the ions of features from the
data in the first (LC-MS) run.

Our approach is to match MS2 spectra against the *significant features*
determined earlier based on their precursor m/z and retention time (given an
acceptable tolerance) to the feature’s m/z and retention time. This can be
easily done using the featureArea() function that effectively considers the
*actual m/z and retention time ranges of the features’ chromatographic peaks*
and therefore increase the chance of finding a correct match. This, however,
also assumes that retention times between the first and second run don’t differ
by much. Alternatively, we would need to align the retention times of the second
LC-MS/MS data set to those of the first.

```{r}
#' Define the m/z and retention time ranges for the significant features
target <- featureArea(lcms1)[rownames(res_sts), ] |> as.data.frame()
target$rtmin <- target$rtmin - 2
target$rtmax <- target$rtmax + 2
```

We next *identify the fragment spectra* with their precursor m/z and retention
times within these ranges. We use the filterRanges() function that allows to
filter a Spectra object using multiple ranges simultaneously. We apply this
function separately to each feature (row in the above matrix) to extract the MS2
spectra representing fragmentation information of the presumed feature’s ions.

```{r}
#' Identify for each feature MS2 spectra with their precursor m/z and
#' retention time within the feature's m/z and retention time range
ms2_004_fts <- apply(target[, c("rtmin", "rtmax", "mzmin", "mzmax")],
                     MARGIN = 1, FUN = filterRanges, object = ms2_004,
                     spectraVariables = c("rtime", "precursorMz"))
# lengths(ms2_004_fts)

ms2_004_fts_test<- ms2_004_fts[lengths(ms2_004_fts) > 0]
lengths(ms2_004_fts_test)
```

The result from this apply() call is a *list of Spectra*, each element
representing the result for one feature. With the exception of the last feature,
multiple MS2 spectra could be identified. We next combine the list of Spectra
into a single Spectra object using the *concatenateSpectra() function* and add
an additional spectra variable containing the respective feature identifier.

```{r}
l <- lengths(ms2_004_fts)
#' Combine the individual Spectra objects
ms2_004_fts <- concatenateSpectra(ms2_004_fts) 
lengths(ms2_004_fts)
#' Assign the feature identifier to each MS2 spectrum
ms2_004_fts$feature_id <- rep(rownames(res_sts), l)
# fix
## Save for reuse in other vignettes
# save(ms2_004_fts, file = "objects/spectra_significant_fts.RData") ## Error in gzfile(file, "wb") : cannot open the connection
```
### MassBank

We next build our reference data which we need to process the same way as our
query spectra. We extract all fragment spectra from the MassBank database,
restrict to positive polarity data (since our experiment was acquired in
positive polarity) and perform the same processing to the fragment spectra from
the MassBank database.

```{r}
#' load reference data
ah <- AnnotationHub()
#' List available MassBank data sets
query(ah, "MassBank")
```

```{r}
#' Load one MassBank release: choose last release. 
mb <- ah[["AH119519"]]

ms2_ref <- Spectra(mb) |>
    filterPolarity(1L) |>
    filterIntensity(intensity = low_int) |>
    filterPrecursorPeaks(ppm = 50, mz = ">=")
ms2_ref <- ms2_ref[lengths(ms2_ref) > 1] |>
    scalePeaks()
```

We use the matchSpectra() function from the MetaboAnnotation package with the
CompareSpectraParam to define the settings for the matching.

```{r}
register(SerialParam())

#' Define the settings for the spectra matching.
prm <- CompareSpectraParam(ppm = 40, tolerance = 0.05,
                           requirePrecursor = TRUE,
                           THRESHFUN = function(x) which(x >= 0.6))
#Note: same parameters?

ms2_mtch <- matchSpectra(ms2_004_fts, ms2_ref, param = prm)
ms2_mtch

```

From 691 query MS2 spectra, only 13 could be matched to one reference fragment
spectrum.

Below we restrict the results to these matching spectra and extract all metadata
of the query and target spectra as well as the similarity score

```{r}
#' Keep only query spectra with matching reference spectra
ms2_mtch <- ms2_mtch[whichQuery(ms2_mtch)]

#' Extract the results
ms2_mtch_res <- matchedData(ms2_mtch)
nrow(ms2_mtch_res)
ms2_mtch_res$target_name
```
The annotations seems to be the same in most of the cases. So it seems to be
only 4 metabolites (lipids). NOTE: do I need to modify the parameters?

Now, we have query-target pairs with a spectra similarity higher than 0.7. Below
we thus iterate over the feature-compound pairs and select for each the one with
the highest score. As an identifier for a compound, we use its fragment
spectra’s INCHI-key, since compound names in MassBank do not have accepted
consensus/controlled vocabularies.

```{r}
#' - split the result per feature
#' - select for each feature the best matching result for each compound
#' - combine the result again into a data frame
ms2_mtch_res <-
    ms2_mtch_res |>
    split(f = paste(ms2_mtch_res$feature_id, ms2_mtch_res$target_inchikey)) |>
    lapply(function(z) {
        z[which.max(z$score), ]
    }) |>
    do.call(what = rbind) |>
    as.data.frame()

#' List the best matching feature-compound pair
pandoc.table(ms2_mtch_res[, c("feature_id", "target_name", "score",
                              "target_inchikey")],
             style = "rmarkdown",
             caption = "Table 9.MS2 annotation results.",
             split.table = Inf) 
#' Export as df to manually analyse them
ms2_MassBank_result <- ms2_mtch_res[, c("feature_id", "target_name", "score", "target_inchikey")]
write_xlsx(ms2_MassBank_result, "ms2_MassBank_results.xlsx")
```

From all 35 significant features, only 10 could be annotated to a compound based 
on the MS2-based approach. However, annotations show 5 out 10 annotations is the 
same compound. Thus, combining the information from the MS1- and MS2 based 
annotation we can only annotate 5 feature with a considerable confidence. 
(how is the score calculated? Threshold)

MS 2 data is not good at the moment? GNPS could be better?
NOTE: what is the score based on?

We below plot the EIC for this feature highlighting the retention time at which 
the highest scoring MS2 spectra was recorded and create a mirror plot comparing 
this MS2 spectra to the reference fragment spectra.

```{r}
#' Define colors for samples only 
leg_devices <- leg_full[-c(1,2)]
col_devices <- leg_devices[sampleData(lcms1[!idx_blank_qc])$device]

#' Extract and plot EIC for the annotated feature
eic <- featureChromatograms(lcms1, features = ms2_mtch_res$feature_id[1])
plot(eic, col = col_sample, peakCol = col_sample[chromPeaks(eic)[, "sample"]],
     peakBg = paste0(col_sample[chromPeaks(eic)[, "sample"]], 20))
legend("topright", col = leg_devices, legend = names(leg_devices), lty = 1)

#' Identify the best matching query-target spectra pair
idx <- which.max(ms2_mtch_res$score)

#' Indicate the retention time of the MS2 spectrum in the EIC plot
abline(v = ms2_mtch_res$rtime[idx])
```
ADD: only plot features that are significant in each comparison. 
- add columns with coef and adjp in the matrix
- only plot features in that row give me the devices that have an adjp >0.05. 
- plot only devices that are sinficantly different. 
- export it: for loop, png with info to save it, plot, dev off (finish initial 
graphical device). 

FIX: export the EIC of each peack annotated with MassBank. 

Mirror spectra:
```{r}
#' Get the index of the MS2 spectrum in the query object
query_idx <- which(query(ms2_mtch)$.original_query_index ==
                                  ms2_mtch_res$.original_query_index[idx])
query_ms2 <- query(ms2_mtch)[query_idx]
#' Get the index of the MS2 spectrum in the target object
target_idx <- which(target(ms2_mtch)$spectrum_id ==
                                    ms2_mtch_res$target_spectrum_id[idx])
target_ms2 <- target(ms2_mtch)[target_idx]

#' Create a mirror plot comparing the two best matching spectra
plotSpectraMirror(query_ms2, target_ms2)
legend("topleft",
       legend = paste0("precursor m/z: ", format(precursorMz(query_ms2), 3)))
```
FIX: Johannes?
Scores?

### GNPS

I first import the GNPS library for MS2 spectra matching.

```{r}
#' Load GNPS library. 
library(RSQLite)
library(MsBackendSql)
dbfile <- "MsBackendSql.GNPS.matchms.cleaned.v1.sqlite"
con <- dbConnect(SQLite(), dbfile)
mb <- Spectra(con, source = MsBackendSql())
```

```{r}
#' Filter data for positive ionisation, low intensity fragments, and fragments with higher mz than precursepeak
ms2_GNPS_ref <- mb |>
    filterPolarity(1L) |>
    filterIntensity(intensity = low_int) |>
    filterPrecursorPeaks(ppm = 50, mz = ">=")
ms2_GNPS_ref <- ms2_GNPS_ref[lengths(ms2_GNPS_ref) > 1] |>
    scalePeaks()
```

We use the matchSpectra() function from the MetaboAnnotation package with the
CompareSpectraParam to define the settings for the matching.

```{r}
register(SerialParam())

#' Define the settings for the spectra matching.
prm <- CompareSpectraParam(ppm = 40, tolerance = 0.05,
                           requirePrecursor = TRUE,
                           THRESHFUN = function(x) which(x >= 0.6))
#Note: same parameters?

ms2_GNPS_mtch <- matchSpectra(ms2_004_fts, ms2_GNPS_ref, param = prm)
ms2_GNPS_mtch

```
FIX: if bug appear ask again. 
From 691 query MS2 spectra, only 13 could be matched to one reference fragment
spectrum.

Object of class MatchedSpectra 
Total number of matches: 5219 
Number of query objects: 773 (294 matched)
Number of target objects: 270686 (1060 matched)

Below we restrict the results to these matching spectra and extract all metadata
of the query and target spectra as well as the similarity score

```{r}
#' Keep only query spectra with matching reference spectra
ms2_GNPS_mtch <- ms2_GNPS_mtch[whichQuery(ms2_GNPS_mtch)]

#' Extract the results
ms2_GNPS_mtch_res <- matchedData(ms2_GNPS_mtch)
nrow(ms2_GNPS_mtch_res)
ms2_GNPS_mtch_res$target_name
```
COMMENT: 

Now, we have query-target pairs with a spectra similarity higher than 0.7. Below
we thus iterate over the feature-compound pairs and select for each the one with
the highest score. As an identifier for a compound, we use its fragment
spectra’s INCHI-key, since compound names in MassBank do not have accepted
consensus/controlled vocabularies.

```{r}
#' - split the result per feature
#' - select for each feature the best matching result for each compound
#' - combine the result again into a data frame
ms2_GNPS_mtch_res <-
    ms2_GNPS_mtch_res |>
    split(f = paste(ms2_GNPS_mtch_res$feature_id, ms2_GNPS_mtch_res$target_inchikey)) |>
    lapply(function(z) {
        z[which.max(z$score), ]
    }) |>
    do.call(what = rbind) |>
    as.data.frame()

#' List the best matching feature-compound pair
pandoc.table(ms2_GNPS_mtch_res[, c("feature_id", "target_name", "score",
                              "target_inchikey")],
             style = "rmarkdown",
             caption = "Table 9.MS2 annotation results.",
             split.table = Inf) 
#' Export as df to manually analyse them
ms2_MassBank_result <- ms2_GNPS_mtch_res[, c("feature_id", "target_name", "score", "target_inchikey")]
write_xlsx(ms2_MassBank_result, "ms2_MassBank_results.xlsx")
```
NOTE: read in metaboannotation how score is calculated. 

From all 35 significant features, only 10 could be annotated to a compound based 
on the MS2-based approach. However, annotations show 5 out 10 annotations is the 
same compound. Thus, combining the information from the MS1- and MS2 based 
annotation we can only annotate 5 feature with a considerable confidence. 
(how is the score calculated? Threshold)

GNPS could be better?
NOTE: what is the score based on?

We below plot the EIC for this feature highlighting the retention time at which 
the highest scoring MS2 spectra was recorded and create a mirror plot comparing 
this MS2 spectra to the reference fragment spectra.

```{r}
#' Define colors for samples only 
leg_devices <- leg_full[-c(1,2)]
col_devices <- leg_devices[sampleData(lcms1[!idx_blank_qc])$device]

#' Extract and plot EIC for the annotated feature
eic <- featureChromatograms(lcms1, features = ms2_GNPS_mtch_res$feature_id[1:6])
plot(eic, col = col_sample, peakCol = col_sample[chromPeaks(eic)[, "sample"]],
     peakBg = paste0(col_sample[chromPeaks(eic)[, "sample"]], 20))
legend("topright", col = leg_devices, legend = names(leg_devices), lty = 1)

#' Identify the best matching query-target spectra pair
idx <- which.max(ms2_GNPS_mtch_res$score)

#' Indicate the retention time of the MS2 spectrum in the EIC plot
abline(v = ms2_GNPS_mtch_res$rtime[idx])
```

FIX: export the EIC of each peack annotated with MassBank. 

# REMOVEME

This is just some test code to see how we can create mirror plots for all
results.

```{r}
library(MetaboAnnotation)
library(Spectra)
load("matchedSpectra_result.RData")
ms2_mtch
```

We can next iterate over the features. For each feature we plot the matching MS2
spectra.

```{r}
fts <- unique(ms2_mtch$feature_id)

ppm <- 20
tolerance <- 0.01

dr <- file.path("images", "matched_features")
dir.create(dr, showWarnings = FALSE, recursive = TRUE)

for (ft in fts) {
    ## Subset the data to all query MS2 spectra for that feature
    a <- ms2_mtch[query(ms2_mtch)$feature_id == ft]
    ## Identify all MS2 spectra that could be matched against reference
    for (i in whichQuery(a)) {
        ai <- a[i]
        ## For each query MS2 spectrum, identify all matching target spectra
        for (j in whichTarget(ai)) {
            ## Get the query and the target spectrum.
            q <- query(ai)
            t <- target(ai)[j]
            png(paste0(dr, "/", ft, "-q", i, "-t", j, ".png"),
                width = 10, height = 8, units = "cm", res = 600, pointsize = 6)
            plotSpectraMirror(q, t, ppm = ppm,
                              tolerance = tolerance, main = ft)
            grid()
            legend("topleft", legend = paste0("RT: ", format(rtime(q), digits = 4)))
            legend("bottomleft", legend = t$name)
            dev.off()
        }
    }
}
```

To get the full result table.

```{r, results = "asis"}
res <- matchedData(
    ms2_mtch, columns = c("feature_id", "rtime", "target_name",
                          "target_spectrum_name", "target_inchikey", "score"))


library(pander)
pandoc.table(as.data.frame(res), split.table = Inf, style = "rmarkdown")

```

And only the ones with matches:

```{r, results = "asis"}
res_found <- res[!is.na(res$score), ]
pandoc.table(as.data.frame(res_found), split.table = Inf, style = "rmarkdown",
             caption = "Features with matched MS2 spectra")
```
