---
title: "Data analysis of Pilot Study from Imperial - annotation_LPOS_CONV."
format: html
tbl-cap-location: bottom
editor: visual
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
author: Philippine Louail, Ana Sanchez Lorenzo
---

```{r setup, include=FALSE}
library(knitr)
library(quarto)
knitr::opts_knit$set(root.dir = './')
```

# Loading packages

```{r packages-used, message=FALSE, warning=FALSE}
#| code-fold: true
#| code-summary: "Show the code"
## General bioconductor package
library(BiocStyle)
library(BiocParallel)
library(knitr)

## Data Import and handling
library(readxl)
library(MsExperiment)
library(writexl)
#library(MsIO)
#library(alabaster.se)
library(SummarizedExperiment)

## Preprocessing of LC-MS data
library(xcms)
library(Spectra)
library(MetaboCoreUtils)

## Statistical analysis
library(limma) # Differential abundance
library(matrixStats) # Summaries over matrices

## Visualisation
library(pander)
library(RColorBrewer)
library(pheatmap)
library(vioplot)
library(ggfortify)   # Plot PCA
library(gridExtra)   # To arrange multiple ggplots into single plots
library(grid)
library(UpSetR)

## Annotation
library(AnnotationHub) # Annotation resources
library(CompoundDb)    # Access small compound annotation data.
library(MetaboAnnotation) # Functionality for metabolite annotation.
```

# DATA ORGANISATION

## Data import

In this case, we only import the SumExp with significant features details
(res_sts).

```{r import, message=FALSE}
# Data import
load("preprocessed_lcms1.RData")
load("res_sts.RData")
featureDefinitions(lcms1) |>head()
```

```{r indexing, message=FALSE}
# Index per sample group
idx_qc <- sampleData(lcms1)$device == "QC"
idx_blank <- sampleData(lcms1)$device == "BLANK"
idx_blank_qc <- sampleData(lcms1)$device %in% c("QC","BLANK")
idx_samples <- sampleData(lcms1)$device %in% c("WHATMAN", "MITRA", "CAPITAINER", "PLASMA", "BLOOD")
idx_bl_sp <- sampleData(lcms1)$blank_type %in% c("BLANK_W", "BLANK_C", "BLANK_M")
```

```{r par-process}#' Set up parallel processing using 3 cores}
if (.Platform$OS.type == "unix") {
  register(MulticoreParam(4)) ### chunksize has to be be the same number. 
 } else {
     register(SnowParam(4))
}
```

## Define colors

We will define colors for each of the sample groups based on their sample group
using the *RColorBrewer* package:

```{r define-colors}
#| echo: true
#' Define colors for full dataset
leg_full <- brewer.pal(12, name = "Set3")[c(1, 12, 7, 5, 10, 4, 6)]
names(leg_full) <- unique(sampleData(lcms1)$device)
leg_full["QC"] <- "#f00e0e"

#' define colors for devices only
leg_blank_BuS <- leg_full[-c(1, 2, 6, 7)]
col_blank_BuS <- leg_blank_BuS[sampleData(lcms1[!idx_bl_sp])$device]
leg_blank_BuS <- leg_blank_BuS[c("WHATMAN", "MITRA", "CAPITAINER")]
```

## Participants column addition For Differential abundance analysis

We need to specify participants. For that we need to add a new column from
sample_name column information in lcms1 object and res object.

```{r}
#' Create a new column with participants names without device specification.
sampleData(lcms1)$participants <- substr(sampleData(lcms1)$sample_name, 1, nchar(sampleData(lcms1)$sample_name) -1)
sampleData(lcms1)$participants[idx_blank_qc] <- NA_character_
colData(res_sts)$participants <- substr(colData(res_sts)$sample_name, 1, nchar(colData(res_sts)$sample_name) - 1)
```

# ANNOTATION

## MS1 Annotation

After the differential abundance analysis, the next step is to identify the
features with significant differences in abundances between plasma and the rest
of groups. As the data was acquired using the NPC LPOS method, we could target
the compounds from the NPC library. However, as LC set-up was different
(different sample manager and different tubing length), only m/z values will be
able to match, fnot rt values.

To do the matching of m/z values, MetaboAnnotation package is used. This is the
lowest level of confidence in metabolite annotation. However, it gives ideas
about of metabolites present in samples.

To start with, I imported the library of metabolites called targed_df.

```{r import-ms1-features}
# Import NPC-LPOS library from csv file
target_df <- read.csv("LPOS_ROI_V_5_1_2.csv")
target_df <- as.data.frame(target_df)  # Ensure it is a data frame
head(target_df)  
```

As we know there is a shift in the rt due to the LC set up and tube length I
would change the rt column of the library for this specific dataset. If I check
the rt of "known compounds" EIC there is a shift +30 seconds in comparison with
the rt from the library. However, I will increase the rt range in case there are
some differences between features.

```{r}
# Modify column rt ranges from library
colnames(target_df)
# fit rt ranges
target_df$rtMin <- target_df$rtMin + 30 
target_df$rtMax <- target_df$rtMax + 30 
target_df$rt <- target_df$rt + 30
target_df$rt_minutes <- target_df$rt_minutes + 0.5
#target and query need to have same rt column name
target_df$rtmed <- target_df$rt 
# check it is working
head(target_df)
```

With this modification, I am trying to do MS1 annotation with m/z and rt values.

First, I prepared the data.frame with experimental features detected:

```{r query-df}
# Prepare query data frame
rowData(res_sts)$feature_id <- rownames(rowData(res_sts))
#cont with full dataset
rowData(res_sts)
# Convert to data frame
res_sts_df <- as.data.frame(rowData(res_sts))
res_sts_df
```

Now, we do the matching in comparison with the target dataframe with reference
known compounds. As we already have the ion description in the library, we
include those adducts.

The resulting Matched object (with 10 ppm) with mz and rt values shows that 177
out of 2454 significant features were matched against at least one target
compound. Out of the 378 identify ions in NPC library, we are able to match 159
features. If we try to do MS1 annotation matching m/z and retention time values,
we expect the match to be more accurate because rt will restrict the match and
if features are detected across the analysis it will only identify the features
that matched with biological compounds. The reference library provides already
the m/z of the (validated) ions of the detected compounds. We will thus directly
match the *experimental* m/z of the features against these *reference* m/z
values.

```{r}
# We do the  matching deleting adduct type from parameters. 
# Also, function changes from Mass2MzRtParam() to MzRtParam(). 

parm_mzrt <- MzRtParam(tolerance = 0, ppm = 10,
                       toleranceRt = 10) # here will be to change rt tolerance(sec). 
mtch_mzrt <- matchValues(res_sts, target_df, param = parm_mzrt,
                    rtColname = "rtmed", mzColname = c("mzmed", "mz"))
mtch_mzrt
```

With this approach we identified however **only** features matching the ions
reported in the reference library. As an alternative, we could also use the
exact (monoisotopic) mass of the reference compounds, calculate all possible
(potential) ions/adducts from that and use the m/z values of these to annotate
features. The code for that is shown (but not evaluated) below.

```{r, eval = FALSE}
parm <- Mass2MzRtParam(adducts = c("[M+H]+", "[M+Na]+", "[M+NH4]+", 
                                   "[M+H-H2O]+",  "[M+K]+", "[M]+"),
                       tolerance = 0, ppm = 10,
                       toleranceRt = 10) # here will be to change rt tolerance(sec). 
mtch <- matchValues(res_sts, target_df, param = parm,
                    rtColname = "rtmed", mzColname = "mzmed", 
                    massColname = "Monoisotopic_mass")
mtch
```

-   check biocrates annotation
-   MS1: manual curation. Do MS2 individually.

After matching mz and rt, as expected, we have less matches with the target
object (159 at of 378 annotations; only mz = 226/378). However, it is finding
same compound at different rts in query object. In total, we will have 245
annotations but we will have not manually check those ones. I will continue the
annotation with this matched object. NOTE: the res object has not being filter
by only significant. I doing with all dataset. After general annotation, filter
by features of interest depending if it present or not in each device.

The result object (mtch) contains the full query data.frame and target
data.frame as well as the matching information. To visualize the info from mtch:

```{r visualisation1}
head(query(mtch_mzrt))
```

```{r visualisation2}
# head(target(mtch_mzrt))
```

```{r visualisation3}
# whichQuery(mtch_mzrt)
```

```{r visualisation4}
# whichTarget(mtch_mzrt)
```

```{r visualisation5}
# colnames(mtch_mzrt)
```

The `mtch_mzrt` variable contains the matching result, but also the full query
and target objects used for the matching. The data is organized along the query
object, i.e. the length of the `Matched` object is the same as the query object
had rows:

```{r}
length(mtch_mzrt)
nrow(res_sts)
```

We can restrict the results to specific rows in query, i.e. if we want to know
the matching results for e.g. the row (feature) with index 13:

```{r}
mtch_mzrt[13]
```

This feature could however not be annotated (we have 0 matches against the
target library). The `whichQuery()` function can be used to get the indices of
features in *query* for which a match could be found:

```{r}
whichQuery(mtch_mzrt)
```

We next use this function to subset the result object to features that could be
annotated to at least one compound in the target library.

```{r}
mtch_mzrt <- mtch_mzrt[whichQuery(mtch_mzrt)]
```

The table below shows all matches.

```{r, results = "asis"}
mtch_res <- matchedData(mtch_mzrt, c("feature_id", "mzmed", "ppm_error",
                                     "rtmed", "score_rt", "target_cpdID", 
                                     "target_cpdName", "target_ion", 
                                     "ppm_error", "target_chemicalFormula"))
   
mtch_res_df <- as.data.frame(mtch_res)
rownames(mtch_res) <- NULL
pandoc.table(mtch_res_df, style = "rmarkdown", split.table = Inf) 
nrow(mtch_res)
```

There are thus some features matching multiple compounds, but most of them seem
to be isomers - and with the chosen retention time tolerance (10 seconds) we can
not reliably assign them.

We next extract the ion chromatograms for the annotated features and indicate in
the plot also the retention time from the target reference library. An example
for such a plot is shown below while all plots are exported to individuals png
files.

```{r}
mtch_mzrt_eic <- featureChromatograms(
    lcms1, features = rowData(query(mtch_mzrt))$feature_id, expandRt = 10)
col_sample <- leg_full[sampleData(lcms1)$device]

#' Create an example plot
i <- 2
a <- mtch_mzrt_eic[i, ]
col_peaks <- col_sample[chromPeaks(a)[, "sample"]]
plot(a, col = paste0(col_sample, 40), peakBg = paste0(col_peaks, 20),
     peakCol = paste0(col_peaks, 80))
# What is maxy indicating?
maxy <- max(chromPeaks(a)[, "maxo"]) 
#' indicate the retention time of the feature
abline(v = mtch_mzrt[i]$rtmed, col = "grey", lty = 2)
#' indicate the retention time of the matched target compounds
points(mtch_mzrt[i]$target_rtmed, maxy * 0.75, lty = 2, type = "h")
text(mtch_mzrt[i]$target_rtmed, maxy * 0.75,
     labels = mtch_mzrt[i]$target_cpdID, pos = 3)
grid()

# dr <- file.path("images", "ms1_matched_features")
# dir.create(dr, recursive = TRUE, showWarnings = FALSE)
# 
# #' Run that code within a loop to export all peaks
# for (i in seq_along(mtch_mzrt)) {
#     a <- mtch_mzrt_eic[i, ]
#     png(paste0(dr, "/", mtch_mzrt[i]$feature_id[1L], ".png"),
#         width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
#     col_peaks <- col_sample[chromPeaks(a)[, "sample"]]
#     plot(a, col = paste0(col_sample, 40), peakBg = paste0(col_peaks, 20),
#          peakCol = paste0(col_peaks, 80))
#     maxy <- max(chromPeaks(a)[, "maxo"])
#     abline(v = mtch_mzrt[i]$rtmed, col = "grey", lty = 2)
#     xpos <- mtch_mzrt[i]$target_rtmed
#     points(xpos, rep(maxy * 0.75, length(xpos)), lty = 2, type = "h")
#     text(xpos, rep(maxy * 0.75, length(xpos)),
#          labels = mtch_mzrt[i]$target_cpdID, pos = 3)
#     grid()
#     dev.off()
# }
```

NOTE: Even if target_rtmed is correct, there is still some difference to where
peak elutes. How this is affecting to annotation and what should be a criteria
to accept the anntoation or not? Should be this discriminant?

While we could try to select a single, most likely, compound for such
multi-matching (e.g. by using the information reported in the `"ppm_error"` or
`"score_rt"` columns above, which provide the absolute m/z difference between
query and target, in ppm, and the difference between the query and target
retention time), we below simply collapse and merge this information.

```{r}
#' Split the data by feature_id
mtch_res <- split(mtch_res, mtch_res$feature_id) 
# tmp is a CompressedSplitDFrameList of feature_id

#' Paste/merge the content of the annotation columns per feature
mtch_res <- lapply(mtch_res, function(z) {
    data.frame(feature_id = z$feature_id[1L],
               cpdID = paste0(z$target_cpdID, collapse = "; "),
               cpdName = paste0(z$target_cpdName, collapse = "; "))
}) |>
    do.call(what = rbind)
mtch_res
```

**Note**: this is just a quick, easy, automated processing of the data. More in
depth analysis could be done by manually checking all features.

Since we have now a single annotation for each feature we can subset the result
object to these features and add this annotation to this object.

```{r}
#' Subset the result object to annotated features
res_sts_ms1 <- res_sts[mtch_res$feature_id, ]
#' Add the annotations.
rowData(res_sts_ms1) <- cbind(
    rowData(res_sts_ms1), mtch_res[, colnames(mtch_res) != "feature_id"])
rowData(res_sts_ms1)
res_sts_ms1
```

### Heatmap

And a heatmap of the features annotated with MS1 that are significant for at
least one microsampling device in comparison with Plasma samples:

```{r}
amap <- log2(assay(res_sts_ms1, "norm_imputed"))
#' center the abundances by row (features) as we're only interested in
#' relative abundance differences between devices
amap <- amap - rowMeans(amap)
#' restrict to features that are significant for at least one device
amap <- amap[rowData(res_sts_ms1)$significant.M | 
                 rowData(res_sts_ms1)$significant.W |
                 rowData(res_sts_ms1)$significant.B | 
                 rowData(res_sts_ms1)$significant.C, ]
# amap: replace in amap rownames by name. 
#nrow(amap) # matix array
# NOTE significant = adjpvalue < 0.05
ann <- data.frame(device = colData(res_sts_ms1)$device)
rownames(ann) <- colnames(amap)
# met <- data.frame(cpdID = rowData(res_sts_ms1)$cpdID)

# redefine colors without QCs and blanks
leg_full <- brewer.pal(12, name = "Set3")[c(7, 5, 10, 4, 6)]
names(leg_full) <- unique(res_sts_ms1$device)
col_sample <- leg_full[res_sts_ms1$device]

# plot heatmap
pheatmap(amap, show_rownames = FALSE, show_colnames = FALSE,
         annotation_col = ann, annotation_colors = list(device = leg_full)) 
         # annotation_row = rowData(res_sts_ms1)$cpdID) 
# find a way to create labels with metabolites annotations (63 lipids sign+ann)

```

AMAP: There are 63 features annotated that are significant for at least one
device.

In the heatmap, we represent the RLA of significant feature that could be
annotated with MS1 data in each sample. Samples were grouped based on the
Pearson correlation and euclidean distance.

After analysis, first 3 features are found in higher abundance and plasma and
blood samples, while it is low abundance in devices. A few samples from mitra
and capitainer are grouped together blood group. The rest of the features do not
seem to have a clear intensity difference. Can it be because of the scale?

NEXT STEPS:

Below I extract the full result from the Matched object:

```{r extration-match}
#' Extracting the results
# mtch_res <- matchedData(mtch_mzrt, c("feature_id", "mzmed", "rtmed", "adduct",
#                                 "target_ion", "ppm_error",
#                                 "target_chemicalFormula",
#                                 "target_Monoisotopic_mass",
#                                 "target_cpdName", "target_InChI",
#                                 "coef.W", "coef.M", "coef.C", "coef.B",
#                                 "adjp.W", "adjp.M", "adjp.C", "adjp.B" ))
# 
# mtch_res
```


### VOLCANO PLOT

After the annotation, we would like to visualise again in the volcano plot which
are the annotation the affect the most to the significant difference between
plasma and the rest of groups. 

For these, we replot the volcano from linear model coloring just the ones that
have an annnotations. 

```{r}
#' Plot volcano plot of the statistical results - Whatman
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res_sts_ms1)$coef.W, -log10(rowData(res_sts_ms1)$adjp.W),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - Annotated lipids significant Plasma vs Whatman",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#0000ffcc")
if (any(rowData(res_sts_ms1)$significant.W)) {
    points(rowData(res_sts_ms1)$coef.W[rowData(res_sts_ms1)$significant.W],
           -log10(rowData(res_sts_ms1)$adjp.W[rowData(res_sts_ms1)$significant.W]),
           col = "#0000ffcc")
}
#' Plot volcano plot of the statistical results - Mitra
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res_sts_ms1)$coef.M, -log10(rowData(res_sts_ms1)$adjp.M),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - Annotated lipids significant Plasma vs Mitra",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#0000ffcc")
if (any(rowData(res_sts_ms1)$significant.M)) {
    points(rowData(res_sts_ms1)$coef.M[rowData(res_sts_ms1)$significant.M],
           -log10(rowData(res_sts_ms1)$adjp.M[rowData(res_sts_ms1)$significant.M]),
           col = "#0000ffcc")
}
#' Plot volcano plot of the statistical results - Capitainer
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res_sts_ms1)$coef.C, -log10(rowData(res_sts_ms1)$adjp.C),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - Annotated lipids significant Plasma vs  Capitainer",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#0000ffcc")
if (any(rowData(res_sts_ms1)$significant.C)) {
    points(rowData(res_sts_ms1)$coef.C[rowData(res_sts_ms1)$significant.C],
           -log10(rowData(res_sts_ms1)$adjp.C[rowData(res_sts_ms1)$significant.C]),
           col = "#0000ffcc")
}
#' Plot volcano plot of the statistical results - Blood
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res_sts_ms1)$coef.B, -log10(rowData(res_sts_ms1)$adjp.B),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - Annotated lipids significant Plasma vs  Blood",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#0000ffcc")
if (any(rowData(res_sts_ms1)$significant.B)) {
    points(rowData(res_sts_ms1)$coef.B[rowData(res_sts_ms1)$significant.B],
           -log10(rowData(res_sts_ms1)$adjp.B[rowData(res_sts_ms1)$significant.B]),
           col = "#0000ffcc")
}
#' Plot volcano plot of the statistical results - sex
par(mfrow = c(1, 1), mar = c(5, 5, 5, 1))
plot(rowData(res_sts_ms1)$coef.S, -log10(rowData(res_sts_ms1)$adjp.S),
     xlab = expression(log[2]~difference),
     ylab = expression(-log[10]~p[BH]), pch = 16, col = "#00000060",
     main = "Volcano Plot - Annotated lipids significant Plasma vs SEX",
     cex.main = 1.5, cex.lab = 1.5, cex.axis = 1.3)
grid()
abline(h = -log10(0.05), col = "#0000ffcc")
if (any(rowData(res_sts_ms1)$significant.S)) {
    points(rowData(res_sts_ms1)$coef.S[rowData(res_sts_ms1)$significant.S],
           -log10(rowData(res_sts_ms1)$adjp.S[rowData(res_sts_ms1)$significant.S]),
           col = "#0000ffcc")
}
```
#### Table of significant features

```{r significant-features-plasma}
# Table of significant features in all groups
tab <- rowData(res_sts_ms1)[rowData(res_sts_ms1)$significant.W,
                    c("feature_id", "cpdID", "cpdName",
                      "mzmed", "rtmed", 
                      "coef.W","pvalue.W","adjp.W", "avg.W",
                      "avg.P", "avg.M","avg.C", "avg.B", "avg.S",
                      "coef.M","pvalue.M","adjp.M",
                      "coef.C","pvalue.C","adjp.C", 
                      "coef.B","pvalue.B","adjp.B", 
                      "coef.S","pvalue.S","adjp.S")] |>
    as.data.frame()
tab <- tab[order(abs(tab$coef.W), decreasing = TRUE), ]
kable(tab, format = "pipe")
```
suspecious that adj pvalue = 0? 

### QUALITATIVE ANALYSIS: heatmap binary

NEXT STEPS: Ms2 data

-   concatenate annotationes in repeated features

-   Add rtime (spectrum), rtmed(ms1), dif rt, cpdID (Ms1) and feature ID (MS1).

-   Manual filtering
