---
title: "Data analysis of Pilot Study from Imperial - Full data."
format: html
tbl-cap-location: bottom
editor: visual
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
author: Philippine Louail, Ana Sanchez Lorenzo
---

```{r setup, include=FALSE}
library(knitr)
library(quarto)
knitr::opts_knit$set(root.dir = './')
```

# Data description

The aim of this pilot study is to compare three microsampling devices profiles
with conventional liquid samples, plasma and whole blood, in an untargeted LCMS
analysis. The final objective is to determine which device perform a similar
profile to plasma, gold standard sample, generating less artefacts. For this,
there will be a qualitative comparison (number of features, metabolites
annotated) and a semi-quantitative comparison (intensity and statistical
differences).

[**Samples of study**]{.underline}

The samples used in this study were venous blood samples collected from 10
healthy individuals ( 5 men and 5 women) in EDTA tubes. Part of this blood was
used as whole blood sample, part was used for plasma extraction and part was
spiked in devices to simulate sample collection through fingerprick.

Total number of samples: 66

double blanks:

-   BLANK_001(beginning)

-   BLANK_002 (beginning)

-   BLANK_003 (beginning)

-   BLANK_END

Blanks from devices:

-   Blank_W

-   Blank_M

-   Blank_C

Randomise Samples:

-   10 x WHATMAN

-   10 x MITRA

-   10 x CAPITAINER

-   10 x PLASMA

-   10 x BLOOD

-   10 x QCs

[**Sample preparation and acquisition**]{.underline}

Samples were treated H2O and IPA for protein precipitation and lipids
extraction. Then extracts were analysed in a UPLC-ESI-QTOF system (Synapt-G2S)
(Waters Corporation) using resolution mode for MS1 and FastDDA for MS2 data.

## Loading packages

```{r packages-used, message=FALSE, warning=FALSE}
#| code-fold: true
#| code-summary: "Show the code"
## General bioconductor package
library(BiocStyle)
library(BiocParallel)
library(knitr)

## Data Import and handling
library(readxl)
library(MsExperiment)
library(SummarizedExperiment)

## Preprocessing of LC-MS data
library(xcms)
library(Spectra)
library(MetaboCoreUtils)

## Statistical analysis
library(limma) # Differential abundance
library(matrixStats) # Summaries over matrices

## Visualisation
library(pander)
library(RColorBrewer)
library(pheatmap)
library(vioplot)
library(ggfortify)   # Plot PCA
library(gridExtra)   # To arrange multiple ggplots into single plots
library(grid)

## Annotation
library(AnnotationHub) # Annotation resources
library(CompoundDb)    # Access small compound annotation data.
library(MetaboAnnotation) # Functionality for metabolite annotation.
```

# Data import

```{r import, message=FALSE}
# Data import
#' read the sample descriptions from an xlsx sheet
pd <- read_xlsx("metadata_lcms_rpos.xlsx", sheet = "samples") |>
  as.data.frame()

#' Import the data
MZML_PATH <- file.path("data/lpos_conv")

lcms1 <- readMsExperiment(file.path(MZML_PATH, "/", pd$mzML_file),
                          sampleData = pd)
```

```{r par-process}
#' Set up parallel processing using 3 cores
if (.Platform$OS.type == "unix") {
  register(MulticoreParam(4)) ### chunksize has to be be the same number. 
 } else {
     register(SnowParam(4))
}
```

## Data organisation

## Indexing per sample group

```{r indexing, message=FALSE}
# Index per sample group
idx_qc <- sampleData(lcms1)$device == "QC"
#idx_qc <- which(sampleData(lcms1)$device == "QC")
idx_blank <- sampleData(lcms1)$device == "BLANK"
idx_blank_qc <- sampleData(lcms1)$device %in% c("QC","BLANK")
idx_double_blank <- sampleData(lcms1)$blank_type %in% c("BLANK_001", "BLANK_002", "BLANK_END")
```

```{r lcms1}
lcms1
```

Below we provide a brief overview of the data structure and content. The
`sampleData()` function extracts sample information from the object.

```{r phenodata}
#| tbl-cap: "Table 1. Samples from the data set."
sampleData(lcms1) |> kable()
```

There are `r length(sampleData(lcms1))` samples in this data set. We only use
QCs here, this will allows to get an idea of the parameters needed for the
overall analysis without wasting too much time during computation.

## Define colors

We will define colors for each of the sample groups based on their sample group
using the *RColorBrewer* package:

```{r define-colors}
#| echo: true
#' Define colors for full dataset
leg_full <- brewer.pal(12, name = "Set3")[c(1, 12, 7, 5, 10, 4, 6)]
names(leg_full) <- unique(sampleData(lcms1)$device)
col_sample <- leg_full[sampleData(lcms1)$device]
#leg_full > leg_full

#' Define colors for only QCs in injection order
leg_qc <- brewer.pal(10, name = "Set3")
names(leg_qc)  <- unique(sampleData(lcms1[idx_qc])$sample_name)
col_qc_inj_order <-leg_qc[sampleData(lcms1[idx_qc])$sample_name]

#' Define colors for full dataset without blanks
leg_noblank <- leg_full[c(-1)]
col_noblank <- leg_noblank[sampleData(lcms1[!idx_blank])$device]

#' Define colors for samples only 
leg_devices <- leg_full[-c(1,2)]
col_devices <- leg_devices[sampleData(lcms1[!idx_blank_qc])$device]

#' Define colors for full dataset with different blank type
leg_blank_type <- brewer.pal(12, name ="Set3")[c(2, 9, 11, 3, 1, 12, 7, 5, 10, 4, 6, 8)]
names(leg_blank_type) <- unique(sampleData(lcms1)$blank_type)
col_sample_blank_type <- leg_blank_type[sampleData(lcms1)$blank_type]
```

The MS data of this experiment is stored as a *`Spectra` object* (from the
`r Biocpkg("Spectra")` Bioconductor package) within the `MsExperiment` object
and can be accessed using `spectra()` function. Each element in this object is a
spectrum.

```{r spectra-objec, eval = FALSE}
#' Access Spectra Object
spectra(lcms1)
```

We can also summarize the number of spectra and their respective MS level. We
only have MS1 here.

```{r number_spectra, eval = FALSE}
#' Count the number of spectra with a specific MS level per file.
spectra(lcms1) |>
    msLevel() |>
    split(fromFile(lcms1)) |>
    lapply(table) |>
    do.call(what = cbind)
```

The present data set thus contains only MS1 data, which is ideal for
quantification of the signal. Also all the samples seem to have a similar number
of scans, which is good.

As another example, we below determine the retention time range for the entire
data set.

```{r rt-range}
#' Retention time range for entire dataset
spectra(lcms1) |>
    rtime() |>
    range()
```

Data obtained from LC-MS experiments are typically analyzed along the retention
time axis, while MS data is organized by spectrum, orthogonal to the retention
time axis.

# Data visualization and general quality assessment

Effective visualization is paramount for inspecting and assessing the quality of
MS data. For a general overview of our LC-MS data, we can:

-   Combine all mass peaks from all (MS1) spectra of a sample into a single
    spectrum in which each mass peak then represents the maximum signal of all
    mass peaks with a similar *m/z*. This spectrum might then be called **Base
    Peak Spectrum (BPS)**, providing information on the most abundant ions of a
    sample.
-   Aggregate mass peak intensities for each spectrum, resulting in the **Base
    Peak Chromatogram (BPC)**. The BPC shows the highest measured intensity for
    each distinct retention time (hence spectrum) and is thus orthogonal to the
    BPS.
-   Sum the mass peak intensities for each spectrum to create a **Total Ion
    Chromatogram (TIC).**
-   **Compare the BPS** of all samples in an experiment to evaluate similarity
    of their ion content.
-   **Compare the BPC** of all samples in an experiment to identify samples with
    similar or dissimilar chromatographic signal.

In addition to such general data evaluation and visualization, it is also
crucial to investigate **specific signal** of e.g. internal standards or
compounds/ions known to be present in the samples. By providing a reliable
reference, internal standards help achieve consistent and accurate analytical
results.

## Chromatographic Data Visualization: BPC and TIC

```{r bpc1}
#| fig-cap: "Figure 1. BPC before rt filtering of all samples colored by device."
#| code-fold: true
#| code-summary: "Show the code"
#' Extract and plot BPC for full data
bpc <- chromatogram(lcms1, aggregationFun = "max")

plot(bpc, col = paste0(col_sample, 80), main = "BPC", lwd = 1.5)
grid()
legend("topright", col = leg_full, legend = names(leg_full), 
       lty = 1, lwd = 2, horiz = FALSE,
       bty = "n", cex = 0.9)
```

Evaluating the BPC, we detected a intense peak between 0 to 60 seconds of
features that didn't retain in the columns and a typical base line peak coming
from the solvent. Also after 700 there shouldn 't be any interesting features.
Rentention time filtering will need to be performed.

We can also compare similarities of the TICs in the heatmap. Thus we bin() the
chromatographic signal per sample along the retention time axis into a bins of 2
seconds resulting in the same number of bins/data points. We can then calculate
the pairwise similarities between these data vectors using the cor() function
and visualize the result using pheatmap().

```{r heatmap}
#| fig-cap: "Figure 2. Heatmap of TIC similarities before filtering."
#| code-fold: true
#| code-summary: "Show the code"
#' Total ion chromatogram
tic <- chromatogram(lcms1, aggregationFun = "sum") |>
  bin(binSize = 2)

#' Calculate similarity (Pearson correlation) between BPCs
ticmap <- do.call(cbind, lapply(tic, intensity)) |>
  cor()

#' Add sample names as rownames and colnames
rownames(ticmap) <- colnames(ticmap) <- sampleData(lcms1)$sample_name
ann <- data.frame(device = sampleData(lcms1)[, "device"])
rownames(ann) <- rownames(ticmap)

#' Plot heatmap
pheatmap(ticmap, annotation_col = ann,
         annotation_colors = list(device = leg_full),
         show_rownames = FALSE, show_colnames = FALSE)
```

We can see separation of samples per group. Only Blood, Whatman and Mitra
samples are grouped together and some of the samples mixed between groups,
probably due to the same blood extraction. However, Capitainer is separated from
the others. This disappears if we filter the rt and plot the heatmap again.

After filtering, we realised the Heatmap had a different layout. At the
beginning it seems the Capitainer devices were different to the rest. However,
now it seems all devices are grouped together. In that way, I want to plot the
heatmap again to check this variation is due to filtering or due to another
modification I am not aware of.

Evaluating which part of the chromatogram, along the retention time contributes
most to the observed differences, e.g. of Whatman against the other samples.

```{r evalution-chromatogram-differences}
#| fig-cap: "Figure 3. Heatmap of TIC similarities per 2-second retention time bins along the x-axis"
#| code-fold: true
#| code-summary: "Show the code"
#| fig-height: 5
#| fig-width: 10
ticm <- do.call(cbind, lapply(tic, intensity))
rownames(ticm) <- rtime(tic[, 1L])
colnames(ticm) <- sampleData(lcms1)$sample_name
#' transpose matrix such that rows are samples and columns retention times
ticm <- t(ticm)

#' log2 transforming the intensities and centering them by retention time bin
#' such that each column contains the relative tic for that retention time
ticm_c <- log2(ticm) |>
    scale(center = TRUE, scale = FALSE)
pheatmap(ticm_c, annotation_row = ann, cluster_cols = FALSE, 
         annotation_colors = list(device = leg_full))
```

The heatmap above represents the relative TIC signal per 2-second retention time
bins along the x-axis (relative TIC compared to the average TIC across all
samples in that bin). Samples are shown in columns, and clustered based on their
similarity. Microsampling devices differ mostly in the beginning of the LC. The
largest difference can be observed between Capitainer and all other samples.
Below we repeat the plot, focusing on the first 30 seconds of the LC.

```{r evalution-chromatogram-differences-30s}
#| fig-cap: "Figure 4. Heatmap of TIC similarities per 2-second retention time bins along 2 and 30 seconds"
#| code-fold: true
#| code-summary: "Show the code"
tmp <- ticm_c[, as.numeric(colnames(ticm_c)) < 30]
pheatmap(tmp, annotation_row = ann, cluster_cols = FALSE, 
         annotation_colors = list(device = leg_full))

```

The Capitainer samples show a higher signal between 16 and 18 seconds compared
to all other samples.

Looking at the chromatogram and gradient, we could filter rt of injection phase
and reconditioning. Thus, I filtered the data removing everything before 60s and
after 700 seconds.

```{r bpc-rtfilter}
#' Filter the data based on retention time
lcms1 <- filterRt(lcms1, c(60, 700))
```

```{r bpc-filtered}
#| fig-cap: "Figure 5. BPC after rt filtering of all samples colored by device."
#| code-fold: true
#| code-summary: "Show the code"
# BPC filtered
bpc_filtered <- chromatogram(lcms1, aggregationFun = "max")
#' Plot after filtering
plot(bpc_filtered, col = paste0(col_sample, 80),
     main = "BPC after filtering retention time", lwd = 1.5)
grid()
legend("topright", col = leg_full,
       legend = names(leg_full), lty = 1, lwd = 2, horiz = FALSE, bty = "n",
       cex = 0.9)

```

We can also here compare similarities of the BPCs in a heatmap.

Before plotting the heatmap after rt filtering, I create a new column with the
same device grouping, but with different blank groups: BLANK_1, BLANK_2,
BLANK_W, BLANK_M, BLANK_C, BLANK_END.

```{r heatmap-filtered}
#| fig-cap: "Figure 6. Heatmap of TIC similarities after rt filtering."
#| code-fold: true
#| code-summary: "Show the code"

#' Total ion chromatogram
tic <- chromatogram(lcms1, aggregationFun = "sum") |>
  bin(binSize = 2)

#' Calculate similarity (Pearson correlation) between BPCs
ticmap <- do.call(cbind, lapply(tic, intensity)) |>
  cor()

#' Add sample names as rownames and colnames
rownames(ticmap) <- colnames(ticmap) <- sampleData(lcms1)$sample_name
ann <- data.frame(sample_type = sampleData(lcms1)[, "blank_type"])
rownames(ann) <- rownames(ticmap)

#' Plot heatmap
pheatmap(ticmap, annotation_col = ann,
         annotation_colors = list(sample_type = leg_blank_type),
         show_rownames = FALSE, show_colnames = FALSE)
```

### Known compounds

Throughout the entire process, it is crucial to have reference points within the
dataset, such as well-known ions, useful for the visualization of data
throughout the entire analysis. Most experiments nowadays include internal
standards (IS). In this case, no standards were spiked, so 12 known compounds
were selected from the National Phenome Center library. This library was built
using the same LCMS method and metabolites were annotated using standards
(<https://github.com/phenomecentre/npc-open-lcms/blob/main/LC-MS%20Metabolite%20Annotations/LPOS_ROI_V_5_1_2.csv>).

-   PC(16:0/18:2)
-   LPC(0:0/18:2)
-   PC(18:0/18:2)
-   Cholesterol
-   LPC(0:0/16:0)
-   LPC(18:0/0:0)
-   LPC(18:1/0:0)
-   LPC(O-18:0/0:0)
-   SM(d16:1/16:0)
-   PC(16:0/20:4)
-   PC(16:0/16:0)
-   SM(d18:2/24:0)

We also plot all of them and save the images locally to check manually.

```{r EIC_extract_for_internal_standard1}
#' get the list
intern_standard <- read_xlsx("standard_compound_addition.xlsx") |> 
  as.data.frame()
intern_standard <- intern_standard[!is.na(intern_standard$POS), ]
rownames(intern_standard) <- intern_standard$abbreviation

#'generate calcualte formula
intern_standard$mass <- calculateMass(intern_standard$formula)
intern_standard$mz <- mapply(intern_standard$mass,
                             intern_standard$POS,
                             FUN = mass2mz)

#' Fit for each standards
intern_standard$mzmin <- intern_standard$mz - 0.02
intern_standard$mzmax <- intern_standard$mz + 0.02
intern_standard$rtmin <- intern_standard$RT - 15
intern_standard$rtmax <- intern_standard$RT + 15
```

```{r EIC_extract_for_internal_standard2, eval=FALSE}
#' Extract the EICs
eics <- chromatogram(lcms1[idx_qc],
                     mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]),
                     rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                     chunkSize = 10)

rownames(fData(eics)) <- intern_standard$abbreviation

dr <- "EIC_IS/full/raw/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eics[i, ], main = intern_standard$name[i],
         col = paste0(col_qc_inj_order, 80))
    grid()
    legend("topright", col = leg_qc,
           legend = names(leg_qc), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}
```

EICs show good rt reproducibility, but intensity variability between samples,
not only between groups. It could be related to injection order dependency.

# Data preprocessing

## Chromatographic peak detection

Thee first step of preprocessing if detection of chromatographic peaks. In XCMS,
we use findChromPeaks() which applies different algorithms for peak detection.
In this case, CentWave will be applied as it is effeectve handling non-Gaussian
shaped chromatographic peaks. The results are evaluated with known compounds
extracted ions chromatograms (EICs).

```{r run-find-chrompeak-on-entire-dataset, eval=!file.exists("peakdetect_lcms1.RData")}
#' Using the same settings, but with default snthresh
param <- CentWaveParam(peakwidth = c(5, 20), ppm = 25, integrate = 2, 
                       snthresh = 7, noise = 600)
lcms1 <- findChromPeaks(lcms1, param = param, chunkSize = 10)
save(lcms1, file = "peakdetect_lcms1.RData")
```

Above we found the parameters that have beeing optimised. We decided to continue
with snthresh = 7 and noise = 600 as there were not that much difference in the
detection of peaks, but the number of features decreased from \~ 5000 to \~3000.

Also, ppm values has being determine before using one known compound:
PC_18_0_18_2 (PPM estimated: 10.94; mzdiff: 0.23). As it was the highest peak
and we still decided to continue with default value ppm 25.

```{r run-find-chrompeak-on-entire-dataset2, eval=!file.exists("peakdetect_lcms1.RData")}
#' Update EIC internal standard object
eics_peakdetect <- chromatogram(lcms1[idx_qc],
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
fData(eics_peakdetect) <- fData(eics)

## plot 
dr <- "EIC_IS/full/peakdetect/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    ## assign the sample color to each chromatographic peak
    col_bg <- leg_qc[chromPeaks(eics_peakdetect[i, ])[, "column"]]
    plot(eics_peakdetect[i, ], main = intern_standard$name[i],
         col = paste0(col_qc_inj_order, 80), peakBg = paste0(col_bg, 40))
    grid()
    legend("topright", col = leg_qc,
           legend = names(leg_qc), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}
```

```{r run-find-chrompeak-on-entire-dataset-2, eval=file.exists("peakdetect_lcms1.RData")}
load("peakdetect_lcms1.RData")
```

All peaks of known compounds were detected so no further optimisation was
performed.

(within the small retention time range in EICs not enough signal is present to
properly estimate the background noise. See metabonaut for further
troubleshooting).

### Refine identified chromatographic peaks

As the CentWave algorith can miss some peaks due to overlapping or split peaks
it is important to use from XCMS refinechromPeaks() and
MergeNeighboringPeaksParam() to merge some split peaks. expandMz is important to
keep it small to avoid merging isotopes. ExpanRt: half size of av rt width.
minProp: to determine

```{r refine1, eval=!file.exists("refine_identified_lcms1.RData")}
#' Apply on whole dataset
param <- MergeNeighboringPeaksParam(expandRt = 7, expandMz = 0.0015,
                                    minProp = 0.75)
lcms1 <- refineChromPeaks(lcms1, param = param, chunkSize = 10)
chromPeakData(lcms1)$merged |>
                      table()
save(lcms1, file = "refine_identified_lcms1.RData")
```

After refinement, the chrompeak detected decreased from 137674 to 132055.

```{r eval=file.exists("refine_identified_lcms1.RData")}
load("refine_identified_lcms1.RData")
```

We check the number of identified chromatographic peaks in all samples of a data
set can help spotting potentially problematic samples. Below we count the number
of chromatographic peaks per sample and show these numbers in a table.

```{r refine3}
#| tbl-cap: "Table 2. Samples and number of identified chromatographic peaks."
#' Count the number of peaks per sample and summarize them in a table.
data.frame(sample_name = sampleData(lcms1)$sample_name,
           peak_count = as.integer(table(chromPeaks(lcms1)[, "sample"]))) |>
    kable(format = "pipe") 
```

There is a decrease of identified peaks related with injection order. In average
there will be \~450 peaks detected in blanks and \~2000 peaks detected in study
samples.

## Retention time alignment

We do an initial correspondence round on the QC samples. The alignment will be
based on the features shared in these samples.

```{r rt_alignment_1, eval=!file.exists("align_lcms1.RData")}
#| fig-cap: "Figure 7. Initial correspondence analysis."
# Initial correspondence analysis
param <- PeakDensityParam(sampleGroups = sampleData(lcms1)$device == "QC",
                          minFraction = 1,
                          binSize = 0.01, ppm = 10,
                          bw = 2)
lcms1 <- groupChromPeaks(lcms1, param = param)
```

```{r rt_alignment_2, eval=!file.exists("align_lcms1.RData")}
#' Define parameters of choice
subset <- which(sampleData(lcms1)$device == "QC")
param <- PeakGroupsParam(minFraction = 0.9, extraPeaks = 50, span = 0.5,
                         subsetAdjust = "average",
                         subset = subset)

#' Perform the alignment
lcms1 <- adjustRtime(lcms1, param = param)
```

```{r rt_alignment_plot, eval = FALSE}
#| fig-cap: "Figure 8. Retention time alignment results."
#| fig-height: 5
#| fig-width: 10
#' Visualize alignment results
plotAdjustedRtime(lcms1, col = paste0(col_sample, 80), peakGroupsPch = 1, cex.lab = 1.2)
grid()
legend("topright", col = leg_full,
       legend = names(leg_full), lty = 1, bty = "n", horiz = TRUE, cex = 0.7)
```

The differences in retention times are, in particular if compared with the
retention time width of the identified chromatographic peaks, only very small.
For the present data set we will thus **not** perform a retention time
alignment.

```{r rt_alignment_4, eval=!file.exists("align_lcms1.RData")}
#' Get data before alignment
lcms1_raw <- dropAdjustedRtime(lcms1)

#' Apply the adjusted retention time to our dataset
lcms1 <- applyAdjustedRtime(lcms1)

# Save aligment object 
save(lcms1, file = "align_lcms1.RData")
```

```{r eval=file.exists("align_lcms1.RData")}
load("align_lcms1.RData")
```

```{r rt_alignment_5, eval=FALSE}
# RUN WITH QCs (idx)
#' Update EIC internal standard object
eics_align <- chromatogram(lcms1[idx_qc],
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]))
fData(eics_align) <- fData(eics)

## plot 
dr <- "EIC_IS/full/align/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eics_align[i, ], main = intern_standard$name[i],
         col = paste0(col_qc_inj_order, 80))
    grid()
    legend("topright", col = leg_full,
           legend = names(leg_full), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}
```

Use the next code to test some extraction of chromatograms.

```{r rt_alignment_6, eval=FALSE, include=FALSE}
# Test chromatogram()
c_test <- chromatogram(lcms1[1L], rt = c(400, 450), mz = c(100, 600)) 
plot(c_test, peakType = "none") 
```

## Correspondence

As described above, for the present data set we don't need to perform a
retention time alignment. Thus we load below the data that was saved **before**
performing the alignment and proceed the analysis with that data.

```{r eval=file.exists("refine_identified_lcms1.RData")}
load("refine_identified_lcms1.RData")
```

Generally, the goal of the correspondence analysis is to identify
chromatographic peaks that originate from the same types of ions in all samples
of an experiment and to group them into LC-MS features. Using the function
PlotChromPeakDensity() form xcms package, it is possible to simulate a
corresopndece analyis with the default values for Peak Groups on the EIC of
known compounds.

```{r correspondence_1, eval=!file.exists("corres_lcms1.RData")}
pd_new <- read_xlsx("metadata_lcms_rpos.xlsx", sheet = "samples") |>
  as.data.frame()
pd_new <- DataFrame(pd_new)
sampleData(lcms1) <- pd_new

sampleData(lcms1)$device
sampleData(lcms1)$sex

vec <- c("sex", "device")
p <- do.call(paste, c(as.list(sampleData(lcms1)[,vec]), sep = "_"))
p[p %in% c("NA_BLANK", "NA_QC")] <- NA
f <- factor(p)
f

#' Define the settings for the param
param <- PeakDensityParam(sampleGroups = f, ## this groups per devices 
                          minFraction = 0.75, binSize = 0.01, ppm = 25,
                          bw = 2)

#' Apply to whole data
lcms1 <- groupChromPeaks(lcms1, param = param)
save(lcms1, file = "corres_lcms1.RData")
nrow(featureDefinitions(lcms1))
```

Binsize depends on the resolution of the instrument. We decreate from 0.25 to
0.01. MinFraction: defines features only if chrompeak was identified in at least
75% of the samples of one of the sample groups.

```{r featureDefinitions}
featureDefinitions(lcms1) |> head()
```

```{r eval=file.exists("corres_lcms1.RData")}
load("corres_lcms1.RData")
```

plot eics:

```{r correspondence_2, eval=FALSE}
#' Plot again
eics_dens <- chromatogram(lcms1[idx_qc],
                       rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                       mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]),
                       chunkSize = 10)
fData(eics_dens) <- fData(eics)

dr <- "EIC_IS/full/density/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plotChromPeakDensity(eics_dens[i, ], param = param,
                         main = intern_standard$name[i],   
                         col = paste0(col_sample, "80"),
    peakCol = leg_qc[chromPeaks(eics_dens[i, ])[, "sample"]],
    peakBg = paste0(col_qc_inj_order[chromPeaks(eics_dens[i, ])[, "sample"]], 20),
    peakPch = 16)
    grid()
    legend("topright", col = leg_qc,
           legend = names(leg_qc), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}
```

Another interesting information to look a the the distribution of these features
along the retention time axis.

```{r correspondence_3}
#| tbl-cap: "Table 3. Distribution of features along the retention time axis."
# Bin features per RT slices
vc <- featureDefinitions(lcms1)$rtmed
breaks <- seq(0, max(vc, na.rm = TRUE) + 1, length.out = 15) |>
    round(0)
cuts <- cut(vc, breaks = breaks, include.lowest = TRUE)

table(cuts) |>
    kable(format = "pipe")
```

The results from the correspondence analysis are now stored, along with the
results from the other preprocessing steps, within our `XcmsExperiment` result
object.

This data frame provides the average *m/z* and retention time (in columns
`"mzmed"` and `"rtmed"`) that characterize a LC-MS feature. Column, `"peakidx"`
contains the indices of all chromatographic peaks that were assigned to that
feature. The actual abundances for these features, which represent also the
final preprocessing results, can be extracted with the `featureValues()`
function:

```{r feature_abundance}
#'Extract feature abundances
# featureValues(lcms1, method = "sum") |>
#    head()
```

We can note that some features (e.g. F0002) have missing values in some samples.
This is expected to a certain degree as not in all samples features,
respectively their ions, need to be present. We will address this in the next
section.

-   this quick check kinda make sense because we saw that there was less peaks
    in the later QCs.

```{r feature_abundance_2}
nrow(featureDefinitions(lcms1))
```

## Gap filling

The missing values (NA) is not always because the ion is not present but because
of the failure of the preceding chromatographic peak detection steps. In that
case, for qualitative and differential abundance analysis we will need to do
imputation. However, this gap filling step we can recover the missing values of
the latter category as much as possible to reduce the eventual need for data
imputation.

```{r gapfilling1}
#' Percentage of missing values
sum(is.na(featureValues(lcms1))) /
    length(featureValues(lcms1)) * 100
```

Gap filling: 38%.

38% of ions detected in different samples are missing. this could be due to some
missing data point in the middle of the peak. Instead of imputing of all them,
the gap filling to recues their original intensity value from the raw data.for
that, fillChromPeaks() function (xcms) was used to define an m/z-rt are for each
feature based on detected peaks, where the signal for the respective ion is
expected. Then it integrates the intensities in that area and report it as
feature abundance.

```{r gapfilling2, , eval=!file.exists("gapfilling_lcms1.RData")}
#' Fill in the missing values in the whole dataset
lcms1 <- fillChromPeaks(lcms1, param = ChromPeakAreaParam(), chunkSize = 10)
save(lcms1, file = "gapfilling_lcms1.RData")
```

```{r eval=file.exists("gapfilling_lcms1.RData")}
load("gapfilling_lcms1.RData")
```

```{r gapfilling3}
#' Percentage of missing values after gap-filling
sum(is.na(featureValues(lcms1))) /
    length(featureValues(lcms1)) * 100
```

The gap-filling rescue up to 1.4% Missing Value.

```{r}
lcms1_noblanks <- lcms1[sampleData(lcms1)$device!="BLANK", keepFeatures = TRUE]
```

NOTE: add keepFeatures=TRUE to avoid removing feature definition.

```{r Detected-vs-filled-signal1}
#| fig-cap: "Figure 9. Detect vs filled signal (all dataset)."
#Distribution of data 

# A. DISTRUBTION in ALL SAMPLES. 

#' Get only detected signal in all samples
vals_detect <- featureValues(lcms1, filled = FALSE)
#' Get detected and filled-in signal
vals_filled <- featureValues(lcms1)
#' Replace detected signal with NA
vals_filled[!is.na(vals_detect)] <- NA
#' Identify features with at least one filled peak
has_filled <- is.na(rowSums(vals_detect))
#' Calculate row averages for features with missing values
avg_detect <- rowMeans(vals_detect[has_filled, ], na.rm = TRUE)
avg_filled <- rowMeans(vals_filled[has_filled, ], na.rm = TRUE)
#' Plot the values against each other (in log2 scale)
plot(log2(avg_detect), log2(avg_filled),
     xlim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     ylim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     pch = 21, bg = "#00000020", col = "#00000080",  
     xlab = "log2(avg_detected)"
     )
grid()
abline(0, 1)

```

Detected and gap-filled values are correlated, but for a considerable number of
features the gap-filled values are lower than the respective data from the
detected chromatographic peaks. Note however that the plot above was created on
the full data set, and differences sample composition respectively presence or
abundance of metabolites and features can be expected.

We next identify the gap-filled features with a more than 8-fold lower
gap-filled value compared to the detected value. We visually inspect this data
using a heatmap to potentially identify any relationship between sample groups
and differences in the gap-filled values.

```{r}
#' Identify gap-filled features with a log2 difference in abundance being > 3
#' Note: log2 difference of 2 represents a 2^3 = 8-fold difference in natural scale
#| fig-cap: "Figure 10. Detect vs filled signal (heatmap)."
sel <- which((log2(avg_detect) - log2(avg_filled)) > 3)

#' Get the feature values for these
vals <- featureValues(lcms1)[has_filled, ]
vals <- vals[sel, ]

colnames(vals) <- sampleData(lcms1)$sample_name
## Just use BLANK instead of the individual BLANKS
ann$device[grep("BLANK", ann$device)] <- "BLANK"

pheatmap(log2(vals), annotation_col = ann,
         annotation_colors = list(device = leg_full),
         show_rownames = FALSE, show_colnames = FALSE)

```

Not unexpectedly, the gap-filled abundances in blanks are much lower for all the
selected features. In addition, a clear difference can also be seen for blood
and plasma samples, and to a lower extend also for Mitra samples. Since these
differences are consistent for all samples of the same *device* it can be
assumed that it stems from differences in either sample composition or
systematic device-specific contamination.

We next evaluate the gap-filling using the QC samples. For these we can assume
the same sample composition and gap-filled signal should thus be more similar to
detected signal.

```{r Detected-vs-filled-signal1-qc}
#Distribution of data 
#| fig-cap: "Figure 11. Detect vs filled signal (QCs)."
qc <- lcms1[sampleData(lcms1)$device == "QC", keepFeatures = TRUE]

#' Get only detected signal in all samples
vals_detect <- featureValues(qc, filled = FALSE)
#' Get detected and filled-in signal
vals_filled <- featureValues(qc)
#' Replace detected signal with NA
vals_filled[!is.na(vals_detect)] <- NA
#' Identify features with at least one filled peak
has_filled <- is.na(rowSums(vals_detect))
#' Calculate row averages for features with missing values
avg_detect <- rowMeans(vals_detect[has_filled, ], na.rm = TRUE)
avg_filled <- rowMeans(vals_filled[has_filled, ], na.rm = TRUE)
#' Plot the values against each other (in log2 scale)
plot(log2(avg_detect), log2(avg_filled),
     xlim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     ylim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     pch = 21, bg = "#00000020", col = "#00000080",  
     xlab = "log2(avg_detected)"
     )
grid()
abline(0, 1)

```

The detected (x-axis) and gap-filled (y-axis) values for QC samples are highly
correlated. Especially for higher abundances, the agreement is very high, while
for low intensities, as can be expected, differences are higher and trending to
below the correlation line. Below we, in addition, fit a linear regression line
to the data and summarize its results

```{r lr}
#' fit a linear regression line to the data
l <- lm(log2(avg_filled) ~ log2(avg_detect))
summary(l)
```

The slope of the linear fit is close to 1 (1.03) and the intercept is also small
suggesting that gap filled signal is very close to the detected signal. Also,
the R squared is over 0.9 reflecting the high correlation of the values.

## Filtering Features: Blank flagging

Now we flagged features present mainly in the blank with Blankflag() function
from xcms package. This can be done in the "XcmsExperiment" or in the
"SummarizedExperiment" object.

There will be an extra column created in "featureDefinition" or "rowData" called
"*possible_contaminants*"with a logical value for each feature.

```{r blankflagfilter}
filter<- BlankFlag(blankIndex = idx_double_blank,
                   qcIndex = idx_qc,
)
lcms1 <- filterFeatures(lcms1, filter)

#number of flagged features : 
featureDefinitions(lcms1)$possible_contaminants |> sum(na.rm = TRUE)
```

Blankflag result filter: 2902 total features remain. 338 features were flagged.

### Ion Map

We will represent the m/z vs the retention time coloring the compounds that have
been flagged in the blank and that are present in possible_contaminant column.

```{r ion-map, eval=FALSE}
ion_map <- ggplot(featureDefinitions(lcms1), aes(x = rtmed, y = mzmed, color = possible_contaminants)) +
    geom_point() +
    scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red", "NA" = "grey")) +
    labs(x = "retention Time (rtmed)", y = "m/z (mzmed)", color = "Possible Contaminants") +
    theme_minimal(base_size = 10) +
    theme(panel.border = element_rect(color = "black", fill = NA, size = 0.8))
grid.arrange(ion_map, ncol = 1, heights = unit(4, "in"),widths = unit(7.5, "in"))

#featureDefinitions(lcms1)$possible_contaminants 
```

Representation with base R.

```{r}

fa <- featureArea(lcms1)
xl <- range(fa[, c("rtmin", "rtmax")])

## BPC
par(mfrow = c(2, 1), mar = c(0, 4.5, 0.5, 0.5))
plot(bpc, xlim = xl, main = "", xlab = "", xaxt = "n")
grid()
par(mar = c(4.5, 4.5, 0, 0.5))
## Feature "area"
## color for features
feat_border <- rep("#00000040", nrow(fa))
feat_border[featureDefinitions(lcms1)$possible_contaminants] <- "#ff000040"
plot(NA, NA, xlim = range(fa[, c("rtmin", "rtmax")]),
     ylim = range(fa[, c("mzmin", "mzmax")]), xlab = "retention time",
     ylab = "m/z")
rect(xleft = fa[, "rtmin"], xright = fa[, "rtmax"],
     ybottom = fa[, "mzmin"], ytop = fa[, "mzmax"], border = feat_border)
grid()
```

Some of the potential contaminant signal can be observed clustered as horizontal
lines along the full retention time range. These could represent contamination
with e.g. salts. Also, vertically aligned groups of features are present. These
could represent multiple ions/adducts of the same compound.

### Noise Comparison

Below we compare the noise signals between devices. We first calculate the
overall signal in the dataset and then calculate the signal that is in the
chromatographic peaks detection. We then subtract the two to get the noise
signal.

```{r Noise_comparison, echo=FALSE}
# overall signal in the dataset 
#' - for each file calculate the sum of intensities 
background  <- spectra(lcms1[!idx_blank_qc]) |>
    split(fromFile(lcms1[!idx_blank_qc])) |>
    lapply(tic) |>
    lapply(sum) |>
    unlist()

# Overall signal that is in the chromatographic peaks detection 
detected <- apply(assay(res_sample), 2, function(x) sum(x, na.rm = TRUE))
 
# substract and plot
names(background) <- names(detected) <- res_sample$device
noise <- background - detected
 
f <- factor(names(noise), levels = unique(names(noise)))
group <- split(noise, f)
 
plot(NULL, xlim = c(1, length(group)), ylim = range(unlist(group)), 
     xaxt = "n", xlab = "Devices", ylab = "Noise", 
     main = "Noise comparison between Devices")
for (i in seq_along(group)) {
  points(rep(i, length(group[[i]])), group[[i]], pch = 19, 
col = leg_devices[i])
}
axis(1, at = seq_along(group), labels = names(group))
```

Plasma shows more background noise than devices and seems to be less
reproducible than the others.

\## Preprocessing results

The final results of the LC-MS data preprocessing are stored within the
`XcmsExperiment` object. As well as the exact preprocessing performed and each
setting used.

```{r Process-history}
#' Check first step of the process history
processHistory(lcms1)[[1]]
```

```{r extract_SumExp}
#' Extract results as a SummarizedExperiment
res <- quantify(lcms1, method = "sum", filled = FALSE)
res
```

```{r different_assays_sumExp}
assays(res)$raw_filled <- featureValues(lcms1, method = "sum",
                                        filled = TRUE )
#' Different assay in the SummarizedExperiment object
assayNames(res)
```

```{r save_preprocessed}
#' Save the preprocessing results
#' d <- file.path(tempdir(), "objects/lcms1")
# saveMsObject(lcms1, AlabasterParam(path = d))
#' for now let's do R object because the previous method is not implemented yet.
save(lcms1, file = "preprocessed_lcms1.RData")
save(res, file = "res.RData")
```
